---
title: "gt, gtsummary & rtables -> rtfreporter: building RTF tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gt, gtsummary & rtables -> rtfreporter: building RTF tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
gt_ok <- requireNamespace("gt", quietly = TRUE)
knitr::opts_chunk$set(eval = gt_ok)
```

```{r setup, eval = TRUE}
library(rtfreporter)
```

```{r, include = FALSE, eval = !gt_ok}
cat("This vignette renders only when the `gt` package is installed.\n")
```

The [gt](https://gt.rstudio.com) package by Posit gives users a
friendly grammar for *display tables* -- `cols_label()`,
`tab_header()`, `tab_spanner()`, `cols_align()`,
`tab_source_note()`, `tab_footnote()`, ...  Clinical summary tables are
very often produced by [gtsummary](https://www.danieldsjoberg.com/gtsummary/)
(`tbl_summary()`, `tbl_regression()`, ...) or
[tfrmt](https://gsk-biostatistics.github.io/tfrmt/), both of which
ultimately render to a `gt_tbl`.  When you move from the HTML world to
the clinical TFL world, you want that same metadata to flow into the
RTF document -- not be re-stated in a different vocabulary.

## The recommended workflow: `as_rtftables()`

[as_rtftables()] is the single entry point.  It takes a table object
(`gt_tbl`, a **gtsummary** table, a `data.frame`/tibble, or a list of
these), reads all of the table's metadata, paginates the body, and
returns a list of ready-to-render [rtftable()] objects -- one per page:

```
  gt / gtsummary / tfrmt          as_rtftables()             rtf_tables()
  ------------------------  ->  read metadata + paginate  ->  add to doc  ->  generate_rtfreport()
        (a gt_tbl)               (list of rtftable pages)      (titles /
                                                                footnotes
                                                                flow through)
```

```{r}
library(gt)

g <- gt(head(mtcars, 3)) |>
  cols_label(mpg = "MPG", cyl = "Cyl", disp = "Disp", hp = "HP",
             drat = "Drat", wt = "WT", qsec = "QSec",
             vs = "VS", am = "AM", gear = "Gear", carb = "Carb") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Demographics and baseline characteristics"
  ) |>
  tab_source_note("Source: built-in mtcars dataset.")

pages <- as_rtftables(g)          # read = TRUE by default
length(pages)                      # one page here
class(pages[[1]])
```

Hand the pages straight to [rtf_tables()].  The page-level title and
source notes travel with each page (as `rtf_titles` / `rtf_footnotes`
attributes) and are picked up automatically -- no `read_gt` flag
needed:

```{r}
doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(pages)

doc$titles[[1]]        # title + subtitle, read from the gt object
doc$footnotes[[1]]     # source note
```

For a single page you can use the convenience wrapper [as_rtftable()],
which returns one `rtftable` instead of a list:

```{r}
tbl <- as_rtftable(g)   # == as_rtftables(g, split = "none")[[1]]
class(tbl)
```

## gtsummary in one line

A gtsummary table is converted to a `gt_tbl` automatically (via
`gtsummary::as_gt()`), so the workflow is identical:

```{r, eval = gt_ok && requireNamespace("gtsummary", quietly = TRUE)}
library(gtsummary)

s <- tbl_summary(
  trial[c("age", "grade", "response", "trt")],
  by = trt
)

pages <- as_rtftables(s)

doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(pages)

out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

gtsummary applies row indentation and bold group-header rows through
HTML/CSS styling that RTF cannot reproduce; those visual details are
dropped (see *Limitations* below).  Everything structural -- column
labels, spanning headers, the stub / group rows, footnote texts and
marks -- comes through.

## rtables & tern

`as_rtftables()` also accepts any **rtables** `VTableTree` -- including the
tables produced by **tern** analysis functions, which are rtables objects.
They are read through `formatters::matrix_form()`, so nested column splits
become spanning headers, the row-label stub keeps its indentation, and the
titles / footers / referential footnotes flow through.  In-cell footnote
marks `{N}` are rewritten to `^{N}`.

```{r, eval = requireNamespace("rtables", quietly = TRUE) && requireNamespace("formatters", quietly = TRUE)}
library(rtables)

lyt <- basic_table(title = "Table 2", main_footer = "Source: ADSL") |>
  split_cols_by("ARM") |>
  split_rows_by("RACE") |>
  analyze("AGE", afun = function(x) {
    in_rows("Mean (SD)" = rcell(c(mean(x), sd(x)), format = "xx.x (xx.x)"))
  })

set.seed(1)
adsl <- data.frame(
  ARM  = rep(c("Placebo", "Active"), each = 30),
  RACE = rep(c("WHITE", "BLACK", "ASIAN"), 20),
  AGE  = rnorm(60, 45, 8)
)
tern_like <- build_table(lyt, adsl)

doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(as_rtftables(tern_like))

out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

The rtables `read` tokens are `"col_header"`, `"alignment"`, `"spanning"`,
`"titles"`, `"footnotes"`, `"indent"`, `"footnote_marks"`; `read = TRUE`
(default) reads them all.

## Splitting a long table across pages

`as_rtftables()` inherits the splitting controls of the (now
deprecated) `paginate()`.  For a table that is too tall for one page,
pass `split` and `max_rows`:

```{r}
long <- data.frame(
  Parameter = sprintf("Row %02d", 1:8),
  Value     = round(rnorm(8), 2),
  stringsAsFactors = FALSE
)
pages <- as_rtftables(long, split = "group_force", max_rows = 4,
                      group_col = "Parameter")
length(pages)                       # 2 pages
vapply(pages, function(p) nrow(p$data), integer(1))
```

Per-cell styles read from a gt object are sliced to match each page
automatically.

## What gets read

`read = TRUE` (the default) is shorthand for a set of independent
tokens, grouped into phases.  Pass `read = c("col_header", "titles")`
to opt in selectively.

* **Phase A (shape-preserving, "safe")**
    * `"col_header"`   -- `cols_label()` -> the rtftable's column header
    * `"alignment"`    -- `cols_align()` -> per-column `col_spec[[j]]$align`
    * `"titles"`       -- `tab_header(title, subtitle)` -> the page
                          `titles[[i]]` block
    * `"source_notes"` -- `tab_source_note()` -> the page
                          `footnotes[[i]]` block

* **Phase B (structural, modifies the column space)**
    * `"spanning"` -- `tab_spanner()` -> multi-row col_header with
                      spanning cells
    * `"widths"`   -- `cols_width()` -> `column_widths_twips`
                      (`"NNpx"`) or `col_rel_width` (`"NN%"`)
    * `"hidden"`   -- `cols_hide()` -> drop the column before
                      rendering

* **Phase C (structural, modifies the row space + extra notes)**
    * `"footnotes"` -- `tab_footnote()` (any anchor) -> appended to
                       the page `footnotes[[i]]` block
    * `"stub"`      -- `groupname_col = ...`, `rowname_col = ...`,
                       `tab_stubhead()` -> group-transition rows and
                       a stubhead-labelled stub column

* **Phase D (per-cell formatting)**
    * `"styles"`         -- `tab_style(cell_text(...))` -> per-cell
                            bold / italic / underline / indent, carried
                            on the rtftable's `cell_styles` field
    * `"footnote_marks"` -- the in-cell footnote marks gt renders as
                            `<sup>N</sup>` -> rtfreporter `^{N}`
                            superscript markup (matching the texts read
                            by `"footnotes"`)
    * `"strip_html"`     -- remove stray HTML left in cell values
                            (`<br>` -> line break; other tags dropped)

Anything listed in `read = c(...)` is read; anything left out falls
back to the rtfreporter default (e.g. column names from `names(data)`).

## End-to-end example: a TFL-style adverse-events table

This is the kind of table you would normally build by hand:

```{r}
ae <- data.frame(
  SOC = c("Nervous system disorders",
          "Nervous system disorders",
          "Nervous system disorders",
          "Gastrointestinal disorders",
          "Gastrointestinal disorders"),
  PT  = c("Headache", "Dizziness", "Insomnia",
          "Nausea",   "Diarrhoea"),
  Placebo_n = c("10 (33.3)", "2 ( 6.7)", "3 (10.0)",
                "6 (20.0)",  "4 (13.3)"),
  Active_n  = c("13 (43.3)", "2 ( 6.7)", "4 (13.3)",
                "7 (23.3)",  "5 (16.7)"),
  Total_n   = c("23 (38.3)", "4 ( 6.7)", "7 (11.7)",
                "13 (21.7)", "9 (15.0)"),
  stringsAsFactors = FALSE
)

ae_gt <-
  gt(ae, groupname_col = "SOC", rowname_col = "PT") |>
  cols_label(PT = "Preferred Term",
             Placebo_n = "Placebo",
             Active_n  = "Active",
             Total_n   = "Total") |>
  tab_stubhead(label = "SOC / PT") |>
  cols_align("right", columns = c(Placebo_n, Active_n, Total_n)) |>
  tab_spanner(label = "Treatment Arm",
              columns = c(Placebo_n, Active_n, Total_n)) |>
  tab_header(title    = "Table 14.2.1",
             subtitle = "Adverse Events Summary (Safety Population)") |>
  tab_source_note("Source: ADaM ADAE.") |>
  tab_footnote("Percentages are based on column N.")
```

Convert it with `as_rtftables()` and add the pages to a document:

```{r}
doc <- rtf_document() |>
  rtf_section(
    page = 1,
    secinfo = list(
      header = rtf_header(rows = list(c(l = "Protocol RTF-101",
                                         r = "Page {AUTO_PAGE}"))),
      footer = NULL
    )
  ) |>
  rtf_tables(as_rtftables(ae_gt))
```

`doc$contents[[1L]]` now has:

```{r}
tbl <- doc$contents[[1L]]
nrow(tbl$data)          # 5 PT rows + 2 SOC group rows
tbl$data$PT[1:3]        # SOC row + first two PTs (PT 1 is the group label)
tbl$col_header          # spanner row + stubhead-labelled bottom row
```

Page-level blocks:

```{r}
doc$titles[[1L]]        # title + subtitle
doc$footnotes[[1L]]     # footnote text + source note (in that order)
```

Render to RTF:

```{r}
out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

## Precedence: explicit arguments beat gt-extracted values

The translation honours the documented precedence chain.  If you
specify `col_header = c("A", "B", ...)` explicitly, gt's
`cols_label()` is ignored for that table:

```{r}
g <- gt(head(mtcars, 1)[, "mpg", drop = FALSE]) |>
  cols_label(mpg = "MPG")
tbl <- as_rtftable(g, read = TRUE, col_header = "Override")
tbl$col_header[[1L]]
```

The same applies to:

| Slot                  | Explicit arg                 |
|-----------------------|------------------------------|
| Column header         | `col_header`                 |
| Per-column alignment  | `col_spec[[j]]$align`        |
| Spanner stack         | (passed via `col_header`)    |
| Absolute widths       | `column_widths_twips`        |
| Relative widths       | `col_rel_width`              |
| Page titles           | `rtf_titles()`               |
| Page footnotes        | `rtf_footnotes()`            |

Per-column `col_spec` entries are **deep-merged** -- a user-supplied
`align = "left"` for column 1 only knocks out gt's column-1
alignment, the other columns keep gt's values.

## Limitations

The following gt / gtsummary features depend on HTML/CSS rendering that
RTF cannot reproduce, so they are intentionally **not** carried through:

* **Row indentation.**  gtsummary nests variable levels by indenting
  them with CSS `padding-left`; the rows render flat in RTF.
* **Bold group-header rows.**  Applied by gtsummary through
  `tab_style()` in a way that the style reader does not currently map
  to a whole-row attribute.  (Cell-level `tab_style(cell_text(...))`
  *is* read -- see the `"styles"` token.)
* **Cell background colour / fill.**  Still gt-only.
* **Markdown inside titles / labels.**  `md("**bold**")` is flattened
  to the raw string `"**bold**"`; rtfreporter has no markdown
  subsystem yet.
* **`column_width` strings other than `"NNpx"` or `"NN%"`.**  Mixed-unit
  widths bail out to `NULL` and the renderer falls back to its default
  proportional split.

What *is* now supported (and was not in earlier releases): per-cell
bold / italic / underline / indent via `tab_style(cell_text(...))`, and
in-cell footnote marks rendered as `^{N}` superscripts.

## Legacy: `rtf_tables(read_gt = ...)`

Before `as_rtftables()` existed, a raw `gt_tbl` could be handed
directly to `rtf_tables()` with `read_gt = TRUE` to read its metadata.
That path still works for a single, non-paginated table, but
`as_rtftables()` is preferred: it is the only path that both paginates
*and* reads metadata, and it keeps `rtf_tables()` free of any gt
knowledge.

## See also

* `?as_rtftables` and `?as_rtftable` for the per-argument docs.
* `?rtf_tables` for appending pages to a document.
* The `gt` package documentation: <https://gt.rstudio.com>.
* The `gtsummary` package: <https://www.danieldsjoberg.com/gtsummary/>.

<!-- where-next -->

## Where next

- [Importing tables](importing-tables.html) --- the same for rtables, flextable and huxtable
- [The pharmaverse catalog](tlg-catalog.html) --- the same tables built five different ways

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
