---
title: "Get started with rtfreporter"
description: >
  The recommended workflow end to end: an ARD from cards, a table plan,
  a document with running headers, and the RTF file.  Then the second path:
  bringing a table you already have.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with rtfreporter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# The chunks below RUN.  Every file goes to a temporary folder, so building
# this vignette writes nothing outside the session, and a broken example fails
# the build instead of rotting unnoticed.  The ARD chunks need cards.
has_cards <- requireNamespace("cards", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE, eval = has_cards)
knitr::opts_knit$set(root.dir = tempdir())
options(width = 100)
```

rtfreporter writes clinical **tables, listings and figures** (TFLs) as RTF
files.  There are two ways to get a table into a report:

1. **From an analysis results dataset (ARD)** -- the recommended path.  The
   statistics come from [cards](https://pharmaverse.github.io/cards/) /
   [cardx](https://insightsengineering.github.io/cardx/); a **plan** says how
   they are laid out.  This is what the rest of this page shows.
2. **Bring your own table** -- a table already built with gt, gtsummary,
   rtables / tern, tfrmt, flextable, huxtable or a plain data frame goes in
   through `as_rtftables()`.  See [the end of this page](#bring-your-own-table).

Either way the document around the table is the same: `rtf_document()`, a
running header and footer, `rtf_tables()`, and `generate_rtfreport()` to
write the file.

```
ADSL ──cards──▶ ARD ──normalize_ard()──▶ flat frame ──table_plan() + plan_*()──▶ plan
                                                                                   │
            generate_rtfreport() ◀── rtf_document() |> rtf_section() |> rtf_tables(plan)
```

```{r setup}
library(rtfreporter)
library(cards)
```

## 1. The statistics: an ARD

The example data is the CDISC pilot `ADSL` that cards ships.  The ARD
summarises age, tabulates age group and sex, and does both by arm.
`ard_stack(.by = )` also counts the subjects in each arm -- the plan reads
those counts for the `(N=xx)` in the column header, so nobody types them.

```{r ard}
adsl <- cards::ADSL
adsl$ARM <- factor(adsl$ARM,
                   levels = c("Placebo", "Xanomeline Low Dose",
                              "Xanomeline High Dose"))

ard <- ard_stack(
  adsl, .by = ARM,
  ard_summary(variables = AGE),
  ard_tabulate(variables = c(AGEGR1, SEX)))
```

## 2. Flatten it: `normalize_ard()`

`normalize_ard()` turns the ARD's group columns into ordinary key columns,
named after their variables, and adds what a table needs: the kind of each
row (`.kind`) and its label.  Its column names are the ones the plan uses.

```{r normalize}
nd <- normalize_ard(ard)
nd[1:4, c("ARM", "variable", "variable_level", "stat_name", "stat", ".kind")]
```

## 3. The plan: `table_plan()` and the `plan_*()` verbs

`table_plan()` names the **roles** -- the key that goes across (`cols`) and
the keys that go down (`rows`).  Each `plan_*()` verb then adds one
declaration; nothing runs until the plan is used.

```{r plan}
p <- nd |>
  table_plan(cols = "ARM", rows = c(group = "variable")) |>
  plan_cells(
    continuous  = c("n"         = "{N:.0f}",
                    "Mean (SD)" = "{mean:.1f} ({sd:.2f})",
                    "Median"    = "{median:.1f}",
                    "Min, Max"  = "{min:.0f}, {max:.0f}"),
    categorical = "{n:.0f} ({p:.1f%})",
    notes = FALSE) |>
  plan_labels(c(AGE    = "Age (years)",
                AGEGR1 = "Age group, n (%)",
                SEX    = "Sex, n (%)")) |>
  plan_levels(AGEGR1 = c("<65", "65-80", ">80"), SEX = c("F", "M")) |>
  plan_stub(name = "row_label") |>
  plan_blanks(where = "between_groups", first = TRUE) |>
  plan_columns(widths = c(4, 2, 2, 2)) |>
  plan_style(border = "tfl", align_count_pct = TRUE) |>
  plan_col_header(values = list(n = TRUE), rtf_col_header(
    c("",               "{col}"),
    c("Characteristic", "(N={n})")))
```

* `plan_cells()` turns statistics into text with templates: `{mean:.1f}` is
  the mean to one decimal, `{p:.1f%}` a proportion shown as a percentage.  A
  continuous variable gets one printed row per name.
* `plan_labels()` / `plan_levels()` name the variables and order their levels.
* `plan_stub()` folds the variable and its rows into one indented column.
* `plan_col_header()` writes the column header; `{col}` is the arm and `{n}`
  its population, read from the ARD.

Before writing any RTF, look at the table the plan makes:

```{r look}
plan_apply(p, stage = "table")
```

## 4. The document: `rtf_document()` to `generate_rtfreport()`

The running header and footer are small tables of rows, each with a left,
centre and right cell.  `{PAGE}` / `{TOTAL_PAGES}`, `{PROGRAM}` and
`{DATETIME}` are filled in when the file is written; `{STUDY}` is a token of
our own, given once to `rtf_document(tokens = )`.

```{r doc}
doc <- rtf_document(
    program = "programs/t_dm.R",
    tokens  = list(STUDY = "CDISCPILOT01")) |>
  rtf_section(secinfo = list(
    header = rtf_header(list(
      c(l = "Protocol: {STUDY}", r = "Page {PAGE} of {TOTAL_PAGES}"),
      c(c = "Table 14.1.1  Demographic Characteristics"),
      c(c = "Safety Analysis Set"))),
    footer = rtf_footer(list(
      c(l = "SD = standard deviation."),
      c(l = "Program: {PROGRAM}", r = "{DATETIME}"))))) |>
  rtf_tables(p)

generate_rtfreport(doc, "t_dm.rtf", overwrite = TRUE)
```

`rtf_tables()` takes the plan as it takes any table and runs it.  Long tables
are paged by the plan (`plan_paginate_rows()`), and the header repeats on
every page.

## 5. Where next on this path

* [Tables from an ARD](tables-from-ard.html) -- the same steps in
  more depth, quartiles, and an adverse-events table sorted by frequency
  within a SOC / PT hierarchy.
* [The plan verbs](plan-verbs.html) -- every `plan_*()` verb,
  grouped by job, and the rule that a later layer wins.
* `widen_ard()` -- the ARD half of a plan in one call, giving a data frame
  now instead of a plan to run later.
* `plan_template()` -- reads an ARD and writes a starting plan to edit.
* [Headers, footers and tokens](headers-footers.html) --
  `{PROGRAM_FULL}`, `program_fallback`, `drop_empty_rows` and tokens of your
  own.

## Bring your own table {#bring-your-own-table}

When the table is already built -- by gt, gtsummary, rtables / tern, tfrmt,
flextable, huxtable, or as a data frame -- `as_rtftables()` reads it,
including its labels, spanning headers and footnotes where the source has
them, and gives `rtftable` pages for the same `rtf_tables()`:

```{r byo, eval = TRUE}
df <- data.frame(
  Parameter = c("Subjects", "Age, mean (SD)"),
  Placebo   = c("86", "75.2 (8.59)"),
  Active    = c("84", "75.7 (8.29)"))

doc2 <- rtf_document() |>
  rtf_tables(as_rtftables(df, border = "tfl"),
             titles = list(c("Table 1", "A table brought as a data frame")))
generate_rtfreport(doc2, "t_byo.rtf", overwrite = TRUE)
```

See [Importing tables with as_rtftables()](importing-tables.html),
[gt, gtsummary & rtables](gt-integration.html), and the *same report, every
framework* articles for [demographics](showcase-dm.html) and
[adverse events](showcase-ae.html).

## Listings and figures

* Listings: [Listings with a plan](plan-listings.html) and
  [Listings end to end](listings.html).
* Figures: [From a plot object to a page](figures.html).
* Joining finished files into one deliverable with a table of contents:
  [Rendering, post-processing and assembly](output.html).
