---
title: "Clinical RTF Reporting with rtfreporter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical RTF Reporting with rtfreporter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# The chunks below RUN.  Every render goes to tempfile(), so building this
# vignette writes nothing outside the session, and a broken example fails the
# build instead of rotting unnoticed.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(rtfreporter)
library(magrittr)
```

## Overview

`rtfreporter` produces clinical-trial–style RTF reports (tables, listings,
figures) without depending on r2rtf or any other RTF library.

**Core workflow:**

```r
rtf_document() %>%
  rtf_section(page = 1, secinfo = list(header = ..., footer = ...)) %>%
  rtf_tables(list(tbl1, tbl2)) %>%
  generate_rtfreport("output.rtf", overwrite = TRUE)
```

---

## 1  Minimal one-page report

```{r minimal}
library(rtfreporter)
library(magrittr)

df <- data.frame(
  Subject = c("001", "002", "003"),
  Arm     = c("Active", "Placebo", "Active"),
  Age     = c(34L, 45L, 28L),
  Sex     = c("M", "F", "M"),
  stringsAsFactors = FALSE
)

doc <- rtf_document() %>%
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(
      c(l = "Protocol: RTF-101",         r = "ACME Pharma"),
      c(l = "Table 14.1.1 Demographics", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}")
    )),
    footer = rtf_footer(rows = list(
      c(l = "Source: DM domain.", r = "CONFIDENTIAL")
    ))
  )) %>%
  rtf_tables(list(
    rtftable(df,
      col_rel_width    = c(2, 2, 1, 1),
      row_height_twips = 280L)
  ))

generate_rtfreport(doc, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

---

## 2  Lab shift table — one section per analyte

Each analyte occupies its own RTF section so the analyte name appears in the
section header, changing from page to page.

```{r shift-table}
make_shift_df <- function(a_low, a_norm, a_high, b_low, b_norm, b_high) {
  data.frame(
    Baseline = c("Low", "Normal", "High"),
    A_Low    = a_low,  A_Normal = a_norm,  A_High = a_high,
    B_Low    = b_low,  B_Normal = b_norm,  B_High = b_high,
    stringsAsFactors = FALSE, check.names = FALSE
  )
}

lab_data <- list(
  list(label = "ALT (Alanine Aminotransferase)",
       df    = make_shift_df(c(4,0,0), c(1,13,1), c(0,2,3),
                             c(3,0,0), c(1,14,0), c(0,1,4))),
  list(label = "AST (Aspartate Aminotransferase)",
       df    = make_shift_df(c(5,0,0), c(0,11,2), c(0,1,5),
                             c(4,0,0), c(1,12,1), c(0,2,4))),
  list(label = "HGB (Hemoglobin)",
       df    = make_shift_df(c(2,1,0), c(1,16,1), c(0,1,2),
                             c(3,0,0), c(0,15,2), c(0,2,2)))
)

spanning_hdr <- list(
  list(from = 2L, to = 4L, label = "Treatment A  (N=24)", underline = TRUE),
  list(from = 5L, to = 7L, label = "Treatment B  (N=24)", underline = TRUE)
)
col_hdr    <- c("Baseline", "Low", "Normal", "High", "Low", "Normal", "High")
col_widths <- c(2160L, 900L, 900L, 900L, 900L, 900L, 900L)

col_spec_shift <- lapply(seq_len(7L), function(j)
  list(col = j, align = if (j == 1L) "left" else "center"))

common_footer <- rtf_footer(
  rows = list(
    c(l = "ALT=Alanine Aminotransferase; AST=Aspartate Aminotransferase; HGB=Hemoglobin"),
    c(l = "Low/Normal/High: site reference range categories.", r = "CONFIDENTIAL")
  )
)

# Build document: one section per analyte
doc <- rtf_document()

for (i in seq_along(lab_data)) {
  item <- lab_data[[i]]

  tbl <- rtftable(
    data                = item$df,
    col_header          = col_hdr,
    spanning_header     = spanning_hdr,
    col_spec            = col_spec_shift,
    column_widths_twips = col_widths,
    border              = "tfl",
    row_height_twips    = 280L
  )

  sec_header <- rtf_header(rows = list(
    c(l = "Protocol: LAB-001",                             r = "ACME Pharma"),
    c(l = "Study Title: Phase III Safety Lab Assessment",   r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = "Table 14.3.1  Laboratory Shift Table (Safety Population)"),
    c(l = item$label)
  ))

  doc <- doc %>%
    rtf_section(page = i, secinfo = list(header = sec_header, footer = common_footer)) %>%
    rtf_tables(list(tbl))
}

generate_rtfreport(doc, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

---

## 3  Embedding figures (ggplot2)

```{r figure}
library(ggplot2)

p <- ggplot(
  data.frame(x = 1:10, y = (1:10)^2 + rnorm(10, sd = 3)),
  aes(x = x, y = y)) +
  geom_point(size = 3, colour = "#2166AC") +
  geom_smooth(method = "lm", se = FALSE, colour = "#D6604D") +
  labs(title = "Lab value vs. Visit (simulated)", x = "Visit", y = "Lab value") +
  theme_bw(base_size = 11)

png_path <- tempfile(fileext = ".png")
ggsave(png_path, plot = p, width = 7, height = 4.5, dpi = 150)

doc <- rtf_document() %>%
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(
      c(l = "Protocol: FIG-001",           r = "ACME Pharma"),
      c(l = "Figure 14.1.1  Lab Value vs. Visit (Safety Population)")
    )),
    footer = rtf_footer(rows = list(c(l = "Source: LB domain.")))
  )) %>%
  rtf_tables(list(
    rtfplot(png_path, width_twips = 9000L)
  ))

generate_rtfreport(doc, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

---

## 4  Figure and summary on consecutive pages

Each page holds exactly one content item (one table **or** one figure).
To present a figure together with a summary table, place them on adjacent
pages — both can share the same section header.

```{r fig-plus-table}
tbl_summary <- rtftable(
  data.frame(
    Statistic = c("n", "Mean", "SD"),
    Value     = c("30", "42.7", "8.4"),
    stringsAsFactors = FALSE
  ),
  table_width_pct  = 40,
  table_align      = "left",
  row_height_twips = 280L
)

doc <- rtf_document() %>%
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(
      c(l = "Protocol: FIG-001", r = "ACME Pharma"),
      c(l = "Figure and Summary Table")
    )),
    footer = rtf_footer(rows = list(c(l = "Source: LB domain.", r = "CONFIDENTIAL")))
  )) %>%
  rtf_tables(list(
    rtfplot(png_path, width_twips = 7200L),  # page 1: figure
    tbl_summary                              # page 2: summary table
  ))

generate_rtfreport(doc, tempfile(fileext = ".rtf"), overwrite = TRUE)
```

---

## 5  Table width and alignment

| Parameter | Description |
|---|---|
| `table_width_pct = 70` | 70 % of writable page width |
| `table_width_twips = 9000L` | Absolute: 9000 twips ≈ 6.25 in |
| `table_align = "center"` | Center the table on the page |
| `table_align = "right"` | Right-align the table |

```{r table-sizing}
df <- data.frame(A = 1:3, B = letters[1:3], C = c(1.1, 2.2, 3.3))

tbl_left   <- rtftable(df, table_width_pct = 70)
tbl_center <- rtftable(df, table_width_twips = 9000L, table_align = "center")
tbl_right  <- rtftable(df, table_width_twips = 7200L, table_align = "right",
                        col_rel_width = c(3, 1, 1))
tbl_full   <- rtftable(df, table_width_pct = 100)
```

---

## 6  Column width control

| Method | Parameter | Notes |
|---|---|---|
| Absolute | `column_widths_twips = c(2880L, 1440L, 1440L)` | One value per column |
| Relative | `col_rel_width = c(2, 1, 1)` | Proportional split |
| Auto     | `auto_col_widths(df)` | Font-aware estimation |

```{r col-widths}
tbl_abs <- rtftable(df, column_widths_twips = c(4320L, 2160L, 2160L))
tbl_rel <- rtftable(df, col_rel_width = c(2, 1, 1), table_width_twips = 10080L)

w       <- auto_col_widths(df, table_width_twips = 14400L)
tbl_auto <- rtftable(df, column_widths_twips = w)
```

---

## 7  Header / footer layout

```{r hf-layout}
# Multi-row header
sec_header <- rtf_header(rows = list(
  c(l = "Protocol: RTF-101",     r = "ACME Pharma"),
  c(l = "Phase III Safety Study", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(c = "Table 14.3.1  Shift Table (Safety Population)"),
  c(l = "ALT (Alanine Aminotransferase)")
))

# Footer with two rows
sec_footer <- rtf_footer(rows = list(
  c(l = "ALT=Alanine Aminotransferase", r = "CONFIDENTIAL"),
  c(l = "Low/Normal/High: site reference range categories.")
))
```

---

## 8  Page tokens

| Token | Behaviour |
|---|---|
| `{AUTO_PAGE}` | Dynamic page number (`\chpgn`) |
| `{AUTO_TOTAL_PAGES}` | Dynamic NUMPAGES field — **recommended** |
| `{TOTAL_PAGES}` | Static total computed at render time |
| `{SECTION_PAGES}` | Dynamic SECTIONPAGES field |

Recommended: `"Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"`.

---

## 9  Spanning headers

```{r spanning}
df_shift <- data.frame(
  Baseline = c("Low", "Normal", "High"),
  A_Low = c(4L, 0L, 0L), A_Norm = c(1L, 13L, 1L), A_High = c(0L, 2L, 3L),
  B_Low = c(3L, 0L, 0L), B_Norm = c(1L, 14L, 0L), B_High = c(0L, 1L, 4L)
)

tbl_span <- rtftable(
  data = df_shift,
  col_header = c("Baseline", "Low", "Normal", "High", "Low", "Normal", "High"),
  spanning_header = list(
    list(from = 2L, to = 4L, label = "Treatment A  (N=24)", underline = TRUE),
    list(from = 5L, to = 7L, label = "Treatment B  (N=24)", underline = TRUE)
  ),
  column_widths_twips = c(2160L, 900L, 900L, 900L, 900L, 900L, 900L),
  col_spec = lapply(seq_len(7L), function(j)
    list(col = j, align = if (j == 1L) "left" else "center")),
  border           = "tfl",
  row_height_twips = 280L
)
```

---

## 10  Multi data.frame table

```{r multi-df}
df_a <- data.frame(Label = c("Low", "Normal", "High"), n = c(4L, 13L, 3L))
df_b <- data.frame(Label = c("Low", "Normal", "High"), n = c(3L, 14L, 4L))

tbl_multi <- rtftable(
  data       = list(df_a, df_b),
  col_header = list(
    c("Baseline", "Arm A"),
    c("Baseline", "Arm B")
  ),
  column_widths_twips = c(3600L, 2160L),
  border              = "tfl",
  row_height_twips    = 280L
)
```

---

## 11  Border customization

```{r borders}
tbl_tfl <- rtftable(df, border = "tfl")

# A border applies to whatever you attach it to.  Here: the whole table.
tbl_framed <- rtftable(df,
  border = rtf_border(top = TRUE, bottom = TRUE,
                      left = TRUE, right = TRUE,
                      inside_h = TRUE)
)

# Or one kind of row at a time.
tbl_custom <- rtftable(df, border = "none") |>
  style_zone(
    header   = rtf_border(top = TRUE, bottom = TRUE),
    last_row = rtf_border(bottom = "double", width = 10L)
  )

hdr_bordered <- rtf_header(
  rows   = list(c(c = "Study Title")),
  border = rtf_border(bottom = TRUE)
)
```

---

## 12  Assembling multiple RTF files

```{r assemble}
make_rtf <- function(df, hdr_text, path) {
  doc <- rtf_document() %>%
    rtf_section(page = 1, secinfo = list(
      header = rtf_header(rows = list(
        c(l = hdr_text, r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}")
      )),
      footer = rtf_footer(rows = list(c(l = "CONFIDENTIAL")))
    )) %>%
    rtf_tables(list(df))
  generate_rtfreport(doc, path, overwrite = TRUE)
}

f1 <- tempfile(fileext = ".rtf"); make_rtf(mtcars[1:3, ], "Table 14.1 DM", f1)
f2 <- tempfile(fileext = ".rtf"); make_rtf(mtcars[4:6, ], "Table 14.2 AE", f2)

assembled <- tempfile(fileext = ".rtf")
assemble_rtf(c(f1, f2), assembled, overwrite = TRUE)
```

---

## 13  Auto column widths

```{r auto-widths}
df_ae <- data.frame(
  USUBJID  = c("001-001", "001-002"),
  AETERM   = c("Headache", "Nausea"),
  AESTDTC  = c("2024-01-15", "2024-02-03"),
  AESER    = c("N", "N"),
  stringsAsFactors = FALSE
)

w <- auto_col_widths(df_ae,
  col_header        = "Subject ID | AE Term | Start Date | Serious",
  table_width_twips = 14400L)

tbl_ae <- rtftable(df_ae,
  col_header          = c("Subject ID", "AE Term", "Start Date", "Serious"),
  column_widths_twips = w,
  border              = "tfl",
  row_height_twips    = 280L)
```

---

## 14  Blank separator rows

`blank_rows` accepts three combinable spec modes — integer positions,
variable-change detection, and regex rule.  Combine them in a `list()`
and the positions are unioned.

```{r blank-rows}
df_demog <- data.frame(
  Group    = c("Sex",    "Sex",     "Sex",
                "Race",   "Race",    "Race",
                "Total"),
  Category = c("Male",   "Female",  "Subtotal",
                "Asian",  "Other",   "Subtotal",
                ""),
  N        = c(45L, 35L, 80L, 40L, 40L, 80L, 80L),
  stringsAsFactors = FALSE
)

# Mode 1: integer positions (-1 = after the last row)
tbl1 <- rtftable(df_demog, blank_rows = c(0, 3, 6, -1))

# Mode 2: blank when a variable's value changes
tbl2 <- rtftable(df_demog,
  blank_rows = blank_rows_by_change("Group",
    include_before_first = FALSE, include_after_last = FALSE))

# Mode 3: regex match (e.g. before every "Total" row)
tbl3 <- rtftable(df_demog,
  blank_rows = blank_rows_by_rule("Group", "^Total", where = "before"))

# Combination — positions are unioned
tbl4 <- rtftable(df_demog,
  blank_rows = list(
    c(-1),
    blank_rows_by_change("Group",
      include_before_first = FALSE, include_after_last = FALSE),
    blank_rows_by_rule("Category", "^Subtotal", where = "after")
  ))

# Or attach an attribute to the data frame and let read_attributes pick it up
attr(df_demog, "rtf_blank_rows") <- c(0, 3, 6, -1)
tbl5 <- rtftable(df_demog)   # read_attributes = TRUE by default
```

---

## 15  Titles & footnotes per page

`rtf_tables()` and `rtf_figures()` accept parallel `titles` and
`footnotes` lists — one entry per page, each a character vector whose
elements become individual paragraph lines.  `""` produces a blank
line; `NULL` falls back to the package defaults (title = one blank
centred line; no footnote).

```{r titles-footnotes}
doc <- rtf_document() %>%
  rtf_tables(
    list(df_demog, df_demog),
    titles = list(
      c("Table 14.1.1", "", "Demographics (Safety Population)"),
      c("Table 14.1.2", "", "Demographics by Region")
    ),
    footnotes = list(
      c("Source: ADSL.", "", "N = 160 subjects."),
      NULL
    ),
    border           = "tfl",
    row_height_twips = 280L
  )

# Or set them after content has been added
doc <- rtf_document() %>%
  rtf_tables(list(df_demog, df_demog)) %>%
  rtf_titles(list("Page One", "Page Two")) %>%
  rtf_footnotes(list(NULL, "Source: ADSL."))
```

Titles render as centred bold paragraphs; footnotes render as
left-aligned paragraphs with a top divider line on the first line.
Both inherit the document font size (one font per document, 9 pt by
default).

---

## 16  Shared table style

`rtf_table_style()` builds a **table-style record** that bundles default
borders, alignment, padding and row-height.  Pass the same record to
many `rtftable()` calls and each table snapshots the defaults it needs
at construction time.  To derive a variant, call `rtf_table_style_with()`.

```{r style-theme}
tfl <- rtf_table_style_tfl()   # built-in TFL preset

# Hand the same record to many tables
tables <- lapply(list(df_demog, df_demog, df_demog),
                  function(df) rtftable(df, style = tfl, row_height_twips = 280L))

# Non-mutating derivation
heavy_tfl <- rtf_table_style_with(tfl,
  border_last_row = rtf_border(bottom = "double", width = 20L)
)
```

The S3 architecture is documented in [`LEARNING.md`](https://github.com/ichirio/rtfreporter/blob/main/LEARNING.md) in the GitHub repository.

---

## Quick-reference

### rtftable parameters

| Parameter | Type | Purpose |
|---|---|---|
| `data` | data.frame or list | Content rows |
| `col_header` | chr / list | Column header text; list elements may be label rows (char) or spanning rows |
| `col_header_align` | NULL / scalar / chr(ncol) | Column-header alignment override (default: inherit `align`) |
| `spanning_header` | list | Spanning row placed above `col_header` (legacy; can mix into `col_header` instead) |
| `col_spec` | list | Per-column align / bold / italic / indent / **border** (`rtf_border`) / header_* |
| `border` | `"tfl"` / object | Outer-frame borders (now: top on first header row, bottom on last) |
| `blank_rows` | int / spec / list | Integer positions, `blank_rows_by_change()`, `blank_rows_by_rule()`, or a list thereof |
| `read_attributes` | logical | Read `attr(data, "rtf_blank_rows")` as fallback (default TRUE) |
| `style` | rtf_table_style | Shared S3 theme record; provides defaults overridden by explicit args |
| `col_rel_width` | numeric vector | Proportional column widths |
| `column_widths_twips` | integer vector | Absolute column widths |
| `table_width_twips` | integer | Table total width (twips) |
| `table_width_pct` | numeric 0-100 | Table width as % of page |
| `table_align` | `"left"` / `"center"` / `"right"` | Table position on page |
| `row_height_twips` | integer / NULL | Data row height; NULL = font-aware default |

### rtf_tables / rtf_figures pipe-API extras

| Parameter | Type | Purpose |
|---|---|---|
| `titles` | list of chr | One entry per page; each char vector = one paragraph per line (`""` = blank) |
| `footnotes` | list of chr | Same shape as `titles`; first line carries the top divider |
| All `rtftable()` formatting args | — | Applied to bare `data.frame` items in the list |


### rtfplot parameters

| Parameter | Default | Purpose |
|---|---|---|
| `path` | - | PNG or JPEG file path |
| `width_twips` | `NULL` (full width) | Display width |
| `height_twips` | `NULL` (aspect ratio) | Display height |
| `align` | `"center"` | `"left"` / `"center"` / `"right"` |

### Header / footer keys

| Keys used | Columns | Alignments |
|---|---|---|
| `l=` only | 1 | left |
| `r=` only | 1 | right |
| `c=` only | 1 | center |
| `l=` + `r=` | 2 | left, right |
| `l=` + `c=` + `r=` | 3 | left, center, right |
| unnamed single value | 1 | center (default) |

<!-- where-next -->

## Where next

- [The pipe API in depth](rtfreporter-pipes.html) --- every verb, in the order you call them
- [Page setup](page-setup.html) --- paper size, orientation, margins
- [Output options](output.html) --- where the file goes and what it contains

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
