---
title: "Adding tables and figures"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

A rtfreporter document is a **flat sequence of pages**, and the rule is simple:
**one page holds exactly one content item** -- a table or a figure -- plus an
optional title and footnote block. You add content with `rtf_tables()` (and
`rtf_figures()` for images); each element you pass becomes one page.

This article is the deep dive on content. For document-wide settings (page size,
fonts) see [Page & document setup](page-setup.html); for converting `gt` /
`gtsummary` / `rtables` objects see
[Importing tables](importing-tables.html); for splitting a long table across
pages see [Paginating with as_rtftables()](pagination.html).

```{r data}
df1 <- data.frame(Parameter = c("Age", "  Mean", "  SD"),
                  Value     = c("", "75.1", "8.2"), stringsAsFactors = FALSE)
df2 <- data.frame(Parameter = c("Sex", "  F", "  M"),
                  Value     = c("", "53%", "47%"), stringsAsFactors = FALSE)
```


## A single table

The quickest way is to hand a data.frame straight to `rtf_tables()` with the
formatting you want -- a single item needs no `list()`:

```{r single-bare}
doc <- rtf_document() |>
  rtf_tables(df1, border = "tfl", row_height_twips = 280L, col_rel_width = c(2, 1))
length(doc$contents)        # one page
```

For full control, build the table object yourself with `rtftable()` (or
`as_rtftables()` for gt/rtables sources) and pass that:

```{r single-built}
tbl <- rtftable(df1, border = "tfl", col_rel_width = c(2, 1),
                col_spec = list(list(col = 2, align = "right")))
doc <- rtf_document() |> rtf_tables(tbl)
```

The formatting arguments of `rtf_tables()` (`border`, `col_rel_width`,
`row_height_twips`, `col_spec`, ...) are documented on `?rtftable`; they are the
same options either way.

### Column alignment and `row_title`

By default the **row-heading column is left-aligned and every data column is
centred** — the usual clinical look. The heading column is the first one unless
you say otherwise with `row_title` (an integer vector of column indices or a
vector of column names):

```{r row-title}
# Columns 1-2 are row headings (left); columns 3+ are data (centre).
tbl <- rtftable(df1, row_title = c(1, 2))
```

`row_title` only sets the *default* alignment; an explicit `col_spec` `align`,
an `rtf_table_style`, or alignment read from a gt/rtables source still wins, and
column headers follow their column's data alignment. `rtf_tables()` and
`as_rtftables()` take `row_title` too, so the same default applies to a whole
batch of tables at once.


## The same settings for several tables

Pass a **list** of bare data.frames and the formatting arguments are applied to
**every** one -- handy for a batch of tables that should look identical. Each
element still becomes its own page:

```{r bulk}
doc <- rtf_document() |>
  rtf_tables(
    list(df1, df2),
    border           = "tfl",
    row_height_twips = 280L,
    col_rel_width    = c(2, 1)
  )
length(doc$contents)        # two pages, same look
```


## Different settings per table

When each table needs its own look, build each one separately (with its own
settings) and pass the list. Pre-built `rtftable()` objects keep their own
settings:

```{r individual}
t1 <- rtftable(df1, border = "tfl", row_height_twips = 280L)
t2 <- rtftable(df2, border = NULL,  row_height_twips = 200L)   # no borders

doc <- rtf_document() |> rtf_tables(list(t1, t2))
doc$contents[[1]]$row_height_twips   # 280 -- t1 keeps its own
doc$contents[[2]]$row_height_twips   # 200 -- t2 keeps its own
```

### Override semantics

A formatting argument you pass **explicitly** to `rtf_tables()` overrides that
field on every pre-built table; arguments you leave at their default do not
touch the tables'' own values. This lets you set a shared override without
rebuilding the tables:

```{r override}
doc <- rtf_document() |>
  rtf_tables(list(t1, t2), row_height_twips = 240L)   # explicit -> overrides both
doc$contents[[1]]$row_height_twips   # 240
doc$contents[[2]]$row_height_twips   # 240
```

(Bare data.frames are always built from the supplied arguments; `rtfplot()`
figure objects are never modified.)


## Titles and footnotes

Each page can carry a title block and a footnote block. One block element becomes
one line. Title lines default to centred + bold; footnote lines default to left,
and the footnote's first line carries a top rule as the separator. Use `NULL` for
a page with none.

By default the **title** renders as plain **text paragraphs** spanning the
writable page width (so it is centred over the page), while the **footnote**
renders as a content-width single-column **table** (which carries the separator
rule). You can switch either form with the document defaults
`title_format` / `footnote_format` (`"text"` or `"table"`):

```{r title-format}
doc <- rtf_document(default_format = list(title_format = "table")) |>  # legacy
  rtf_tables(list(df1), titles = list("Table 14.1.1"))
```

Pass `titles` / `footnotes` as lists parallel to the content (one element per
page):

```{r titles}
doc <- rtf_document() |>
  rtf_tables(
    list(df1, df2),
    border    = "tfl",
    titles    = list(c("Table 14.1.1", "Demographics"),
                     c("Table 14.1.2", "Sex")),
    footnotes = list(c("Source: ADSL"), NULL)
  )
```

### Common (one block for every page)

If you pass a **length-1** list, that single block is applied to every page --
handy for a footnote that repeats on all pages:

```{r titles-common}
doc <- rtf_document() |>
  rtf_tables(list(df1, df2),
             footnotes = list(c("Confidential", "Source: ADSL")))   # on both pages
```

### Per-row styling

A block can be a plain character vector (default styling) **or** a list of rows,
where each row is a string or a styled list with any of `align`, `bold`,
`italic`, `underline`, `color` (a `"#RRGGBB"` hex), and `border` (an
[rtf_border()] -- e.g. to recolour or remove the footnote separator):

```{r titles-styled}
doc <- rtf_document() |>
  rtf_tables(
    list(df1),
    titles    = list(list(
      "Table 14.1.1",
      list(text = "Provisional", italic = TRUE, color = "#C00000")
    )),
    footnotes = list(list(
      list(text = "Source: ADSL", border = rtf_border(top = rtf_border_side()))
    ))
  )
```

You can also set blocks after the content with `rtf_titles()` /
`rtf_footnotes()` (both accept a per-page list or a single common block). When
the content comes from a `gt` table via `as_rtftables()`, its title and source
notes flow through automatically -- see
[Importing tables](importing-tables.html).


## Cell-text markup

Cell text (data cells, column / spanning headers, titles and footnotes) is run
through a small markup step controlled by the `markup` argument of `rtftable()`
(and `rtf_tables()` for bare data.frames; the document-wide default is
`rtf_document(default_format = list(markup = ))`). It takes a character vector of:

- `"script"` -- `^{...}` renders as superscript, `_{...}` as subscript (this is
  how adapter-generated footnote reference marks such as `^{1}` become real
  superscripts);
- `"relational"` -- the literal `">="` is converted to the `>=` symbol
  (U+2265) and `"<="` to `<=` (U+2264).

The default is **`"script"`** -- super/subscript on, but the `>=` / `<=` symbol
conversion **off**, so a literal `>=` stays as typed. Opt into the symbol
conversion with `"all"` (or `"relational"`), or turn everything off with
`"none"`:

```{r markup}
df_rel <- data.frame(Criterion = c("Age >= 65", "CrCl < 30"), n = c("120", "8"))

rtftable(df_rel)                       # default: ">=" stays literal
rtftable(df_rel, markup = "all")       # ">=" -> the >= symbol, ^{}/_{} too
rtftable(df_rel, markup = "none")      # no markup at all (everything literal)
```


## Figures

Images are content too: one figure per page, added with `rtf_figures()`. Pass a
list of image paths (PNG/JPEG) or `rtfplot()` objects; `width_twips` /
`height_twips` / `align` control the display, and `titles` / `footnotes` work
exactly as for tables.

**By default a figure embeds at its native size (100%)** -- the pixel dimensions
divided by the image's own DPI (read from the PNG `pHYs` chunk / JPEG JFIF
density). A 2500 x 1438 px plot saved at 300 DPI therefore appears at
2500/300 x 1438/300 in. If the file carries no DPI, the
`rtfreporter.figure.default_dpi` option (96) is assumed. Save your graphics
device at the resolution you want on the page:

```{r figure}
# Save at 300 DPI so the native size is predictable.
img <- tempfile(fileext = ".png")
png(img, width = 2500, height = 1438, res = 300)
plot(1:10, main = "Mean over time"); dev.off()

doc <- rtf_document() |>
  rtf_figures(list(img), align = "center",           # native size (~8.3 x 4.8 in)
              titles = list(c("Figure 14.2.1", "Mean over time")))
length(doc$contents)        # one figure page
```

Pass an explicit `width_twips` (the height then follows the aspect ratio) to
scale a figure -- for example to shrink one that is wider than the page, since
the native size is **not** auto-capped to the writable width:

```{r figure-scale, eval = FALSE}
rtf_figures(list(img), width_twips = 9000L)          # force 6.25 in wide
```

Tables and figures share the same page model, so a report is just a sequence of
content pages -- some tables, some figures -- each optionally titled and
footnoted. To control the running header/footer across those pages, see the
headers & footers article.

See `?rtf_tables` and `?rtf_figures` for the full argument reference.
