---
title: "Borders and rules"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

Clinical tables live and die by their rules: a top and bottom line around the
column header, a group underline beneath a spanning label, an occasional rule
under the last data row. rtfreporter gives you three levels of control, from a
one-word preset down to a single cell. This article walks through all three,
with runnable code.

The border objects are plain S3 records (no reference semantics), so a border
you build can be reused across tables safely. The vocabulary is small:

* `rtf_border_side(style, width, color)` -- one **edge** (a line).
* `rtf_border(top, bottom, left, right)` -- the four edges of one **cell/row**.
* `rtf_table_border(header, spanning, body, first_row, last_row)` -- the
  per-**zone** borders of a whole table.


## Level 1 -- a preset

The `border` argument of `rtftable()` (and `rtf_tables()`, `as_rtftables()`)
accepts a shorthand string:

```{r preset}
df <- data.frame(Parameter = c("Age (years)", "  Mean", "  SD"),
                 Value     = c("", "75.1", "8.2"), stringsAsFactors = FALSE)

tbl <- rtftable(df, border = "tfl")   # clinical TFL preset
```

* `"tfl"` -- the clinical preset: a rule on **top of the first** header row and
  **below the last** header row, a multi-column spanning cell gets an automatic
  group underline where the grouping changes below it (the next row subdivides
  the span), and the data area has **no** borders. This is the default.
* `"none"` (or `NULL`) -- no borders anywhere.

`"tfl"` is exactly `rtf_border_tfl()`:

```{r tfl-object}
rtf_border_tfl()
```


## Level 2 -- zones with `rtf_table_border()`

When the preset is not enough, build the zones yourself. A table has five
border zones:

| Zone | Applies to |
|------|------------|
| `header` | column-header label rows (top on first, bottom on last) |
| `spanning` | spanning-header rows (falls back to `header` when `NULL`) |
| `body` | every data row |
| `first_row` | the first data row (merged **on top of** `body`) |
| `last_row` | the last data row (merged on top of `body`) |

Each zone is an `rtf_border()`; each side is an `rtf_border_side()`. For
example, a header boxed top and bottom plus a thin rule under the **last data
row** (a common "close the table" look):

```{r zones}
line <- rtf_border_side()                       # single, 15 twips (~0.5pt), black

tbl <- rtftable(
  df,
  border = rtf_table_border(
    header   = rtf_border(top = line, bottom = line),
    last_row = rtf_border(bottom = line)
  )
)
```

`rtf_border_side()` controls the look of a line:

```{r side}
rtf_border_side(style = "double", width = 30L, color = "#003366")
```

`style` is one of `"single"`, `"double"`, `"thick"`, `"dash"`, `"dot"` -- and
`"none"`, which we meet below.

### Body (data-area) borders

Body rules are **per row, not per column**: the `body` / `first_row` /
`last_row` zones apply the same edge to every column of the row. To draw a grid
around every data cell, set all four sides on `body`:

```{r body-grid}
box <- rtf_border_box()        # all four sides, single line

tbl <- rtftable(
  data.frame(A = c("1", "2"), B = c("x", "y"), stringsAsFactors = FALSE),
  border = rtf_table_border(body = box)
)
```


## Level 3 -- single columns and single cells

The header is where fine control matters most. Two mechanisms reach inside it.

### A whole column, with `col_spec`

A per-column `border` in `col_spec` is merged on top of the header zone for the
header cell(s) **in that column** -- e.g. a thicker rule under just column 2:

```{r col-spec-border}
tbl <- rtftable(
  df,
  border   = "tfl",
  col_spec = list(
    list(col = "Value",
         border = rtf_border(bottom = rtf_border_side("thick")))
  )
)
```

### A single header cell, with `col_cell(border = )`

For a multi-row header, the unit of control is one cell. Build the header with
`rtf_col_header()` / `col_cell()` and give the cell its own `border`. Because a
per-cell border overrides **everything** -- the zone border *and* the automatic
group underline -- this is how you add or **remove** an individual rule.

Take a 5-column header with two treatment groups. By default each spanning cell
("Drug A", "Drug B") is underlined:

```{r header-default}
df5 <- data.frame(Item = "Age", A_N = 30L, A_Mean = 45.2,
                  B_N = 30L, B_Mean = 46.1, stringsAsFactors = FALSE)

hdr <- rtf_col_header(
  list(col_cell(1, ""),
       col_cell(c(2, 3), "Drug A"),
       col_cell(c(4, 5), "Drug B")),
  c("Item", "N", "Mean", "N", "Mean")
)
tbl <- rtftable(df5, col_header = hdr, border = "tfl")
```

To **remove** the underline beneath "Drug A" only, give that one cell a `"none"`
bottom. Unlike `NULL` (which just leaves a side unset and so inherits the
automatic rule), a `"none"` side is an explicit "no line" that wins on merge:

```{r header-suppress}
hdr <- rtf_col_header(
  list(col_cell(1, ""),
       col_cell(c(2, 3), "Drug A",
                border = rtf_border(bottom = rtf_border_side("none"))),
       col_cell(c(4, 5), "Drug B")),
  c("Item", "N", "Mean", "N", "Mean")
)
tbl <- rtftable(df5, col_header = hdr, border = "tfl")
```

The same field can **add** a rule to a single cell (e.g. a thick bottom under
"Drug B" only) -- pass an `rtf_border()` with the sides you want instead of
`"none"`.

### How a cell's final border is decided

For a header cell the renderer merges, in order of increasing precedence:

1. the **zone** border (`header` / `spanning`, reduced to top-on-first /
   bottom-on-last by the outer frame);
2. the automatic **group underline** (a multi-column spanning cell, drawn only
   where the column grouping changes below it -- i.e. the next header row
   subdivides the span);
3. `col_spec[[col]]$border` for the cell's leftmost column;
4. `col_cell(border = )` on the cell itself.

Later steps override earlier ones side-by-side, so a `"none"` side at step 3 or
4 erases a line drawn by step 1 or 2, and a real line at step 4 replaces one
from step 3. That single rule -- *last writer wins, per side* -- is all you need
to reason about any combination.


## Level 4 -- restyling an existing table: the styling verbs

Everything above configures a table you are *building*. When the table already
exists -- typically because `as_rtftables(x, read_meta = TRUE)` extracted the
labels, alignment and spanning from a gt / gtsummary / rtables object and only
one rule needs to change -- reach for the **post-hoc styling verbs** instead of
rebuilding the header:

```{r verbs}
df <- data.frame(Item = c("Age", "Sex"), A = c("1", "2"),
                 B = c("3", "4"), C = c("5", "6"))
hdr <- rtf_col_header(
  list(col_cell(1, ""), col_cell(c(2, 4), "Treatment Group")),
  list(col_cell(1, ""), col_cell(c(2, 4), "(N = 254)")),
  c("Item", "Placebo", "Drug A", "Drug B")
)
tbl <- rtftable(df, col_header = hdr, border = "tfl")

tbl <- tbl |>
  # solid rule above -- and no rule below -- the "(N = 254)" span only
  style_header(row = 2, cols = 2:4,
               border = rtf_border(top    = rtf_border_side("single"),
                                   bottom = rtf_border_side("none"))) |>
  # rule under one BODY row (a summary/total row, selected by predicate)
  style_body(rows = ~ Item == "Sex",
             border = rtf_border(bottom = rtf_border_side("single"))) |>
  # double rule under the whole table
  style_zone(last_row = rtf_border(bottom = rtf_border_side("double")))
```

Body borders follow the same two-step chain as everything else: the row's
zone border (`body`, crossed with `first_row` / `last_row`) first, then the
per-cell override from `style_body(border = )` -- last writer wins, per side.
Restrict `cols` to rule only part of a row, or pass an
`rtf_border_side("none")` side to erase a zone rule on selected cells.

`style_header(row =, cols =)` addresses header cells by header-row index and
data-column position; `style_cols()`, `style_body()` and `style_zone()` cover
the column, body-cell and zone layers the same way, and `add_header_row()`
inserts a header row after the fact. Every verb also accepts the **page list**
returned by `as_rtftables()` (the edit is applied to every page), so a
converted multi-page table can be restyled in one pipe:

```r
pages <- as_rtftables(gt_obj, read_meta = TRUE) |>
  style_header(row = 2, cols = 2:4,
               border = rtf_border(top    = rtf_border_side("single"),
                                   bottom = rtf_border_side("none")))
```

The merge semantics are the ones this article already taught: last writer
wins, per side. See `?style_header` for the full family, including how a
character label row is promoted to cells when it needs a per-cell border.

<!-- where-next -->

## Where next

- [Output options](output.html) --- what the rendered file contains

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
