---
title: "Borders and rules"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

Clinical tables live and die by their rules: a top and bottom line around the
column header, a group underline beneath a spanning label, an occasional rule
under the last data row. rtfreporter gives you three levels of control, from a
one-word preset down to a single cell. This article walks through all three,
with runnable code.

The border objects are plain S3 records (no reference semantics), so a border
you build can be reused across tables safely. The vocabulary is small:

```r
rtf_border(all, top, bottom, left, right, inside_h, inside_v,
           style, width, color, from)
```

`top`/`bottom`/`left`/`right` are the selection's outer edges and `all` is the
four of them at once; `inside_h` and `inside_v` are the rules *between* its rows
and *between* its cells. Each takes `TRUE` for a rule, `FALSE` or `"none"` for
none, or a style name such as `"double"`. `style`/`width`/`colour` describe the
line those sides draw, and `from` layers a border onto an existing one.

That is the whole vocabulary -- **one function**. There is no separate
table-border type and no per-edge constructor: what a border applies to is
decided by where you attach it, the same way Word's border dialog acts on
whatever is selected.

| Attach it to | The selection is |
|---|---|
| `rtftable(border = )` | the whole table |
| `style_zone(header = , body = , ...)` | that kind of row |
| `col_cell(border = )`, `cell_styles` | one cell |
| `rtf_header(border = )` / `rtf_footer(border = )` | that block |

An edge always means the outer edge of the selection, and an absent `inside_h`
/ `inside_v` always means "no rule there" -- the reading never depends on which
other arguments are present.


## Level 1 -- a preset

The `border` argument of `rtftable()` (and `rtf_tables()`, `as_rtftables()`)
accepts a shorthand string:

```{r preset}
df <- data.frame(Parameter = c("Age (years)", "  Mean", "  SD"),
                 Value     = c("", "75.1", "8.2"), stringsAsFactors = FALSE)

tbl <- rtftable(df, border = "tfl")   # clinical TFL preset
```

* `"tfl"` -- the clinical preset: a rule on **top of the first** header row and
  **below the last** header row, a multi-column spanning cell gets an automatic
  group underline where the grouping changes below it (the next row subdivides
  the span), and the data area has **no** borders. This is the default.
* `"none"` (or `NULL`) -- no borders anywhere.

The same rules are available as a reusable value from `rtf_table_style_tfl()`,
which is what `"tfl"` resolves to:

```{r tfl-object}
rtf_table_style_tfl()
```


## Level 2 -- the whole table

Attach an `rtf_border()` to `rtftable(border = )` and it describes the table:

```{r whole-table}
line <- TRUE        # single, 15 twips (~0.5pt), black

# The listing look: a frame plus a rule under every row, no vertical rules.
tbl <- rtftable(df,
  border = rtf_border(top = line, bottom = line, left = line, right = line,
                      inside_h = line))
```

Add `inside_v = line` for a full grid; leave both out for a bare frame.


## Level 3 -- one kind of row

`style_zone()` aims a border at one of the five row kinds:

| Zone | Applies to |
|------|------------|
| `header` | column-header label rows (top on first, bottom on last) |
| `spanning` | spanning-header rows (falls back to `header` when `NULL`) |
| `body` | every data row |
| `first_row` | the first data row (merged **on top of** `body`) |
| `last_row` | the last data row (merged on top of `body`) |

Each zone takes an `rtf_border()`, read against that zone as the selection. A
header boxed top and bottom plus a thin rule under the **last data row** (a
common "close the table" look):

```{r zones}
tbl <- rtftable(df, border = "none") |>
  style_zone(
    header   = rtf_border(top = line, bottom = line),
    last_row = rtf_border(bottom = line)
  )
```

`style`, `width` and `color` describe the line that the sides named in the same
call draw:

```{r side}
rtf_border(top = TRUE, style = "double", width = 30L, color = "#003366")
```

Naming the style on the side itself is shorthand for the same thing, so
`rtf_border(top = "double")` and `rtf_border(top = TRUE, style = "double")` agree.
`style` is one of `"single"`, `"double"`, `"thick"`, `"dash"`, `"dot"` -- and
`"none"`, which we meet below. When one side must differ, layer a second call:

```{r side-layer}
rtf_border(from = rtf_border(top = TRUE, width = 30L), bottom = TRUE)
```

### Body (data-area) borders

`body` selects every data row, so its `top`/`bottom` are the top and bottom of
the **block** and `inside_h` is the rule between its rows. Likewise `left` and
`right` are the outer columns and `inside_v` the rules between the cells. A
grid around every data cell therefore needs all six:

```{r body-grid}
tbl <- rtftable(
  data.frame(A = c("1", "2"), B = c("x", "y"), stringsAsFactors = FALSE),
  border = "none"
) |>
  style_zone(body = rtf_border(top = line, bottom = line,
                               left = line, right = line,
                               inside_h = line, inside_v = line))
```

Before 0.5.0 an edge on a zone was uniform -- `left`/`right` were drawn on every
cell and `body`'s `top`/`bottom` on every row -- so `rtf_border(all = TRUE)` alone
produced this grid. See *Before and after 0.5.0* at the end of this article for
each case written both ways.


## Level 3 -- single columns and single cells

The header is where fine control matters most. Two mechanisms reach inside it.

### A whole column, with `col_spec`

A per-column `border` in `col_spec` is merged on top of the header zone for the
header cell(s) **in that column** -- e.g. a thicker rule under just column 2:

```{r col-spec-border}
tbl <- rtftable(
  df,
  border   = "tfl",
  col_spec = list(
    list(col = "Value",
         border = rtf_border(bottom = "thick"))
  )
)
```

### A single header cell, with `col_cell(border = )`

For a multi-row header, the unit of control is one cell. Build the header with
`rtf_col_header()` / `col_cell()` and give the cell its own `border`. Because a
per-cell border overrides **everything** -- the zone border *and* the automatic
group underline -- this is how you add or **remove** an individual rule.

Take a 5-column header with two treatment groups. By default each spanning cell
("Drug A", "Drug B") is underlined:

```{r header-default}
df5 <- data.frame(Item = "Age", A_N = 30L, A_Mean = 45.2,
                  B_N = 30L, B_Mean = 46.1, stringsAsFactors = FALSE)

hdr <- rtf_col_header(
  list(col_cell(1, ""),
       col_cell(c(2, 3), "Drug A"),
       col_cell(c(4, 5), "Drug B")),
  c("Item", "N", "Mean", "N", "Mean")
)
tbl <- rtftable(df5, col_header = hdr, border = "tfl")
```

To **remove** the underline beneath "Drug A" only, give that one cell a `"none"`
bottom. Unlike `NULL` (which just leaves a side unset and so inherits the
automatic rule), a `"none"` side is an explicit "no line" that wins on merge:

```{r header-suppress}
hdr <- rtf_col_header(
  list(col_cell(1, ""),
       col_cell(c(2, 3), "Drug A",
                border = rtf_border(bottom = "none")),
       col_cell(c(4, 5), "Drug B")),
  c("Item", "N", "Mean", "N", "Mean")
)
tbl <- rtftable(df5, col_header = hdr, border = "tfl")
```

The same field can **add** a rule to a single cell (e.g. a thick bottom under
"Drug B" only) -- pass an `rtf_border()` with the sides you want instead of
`"none"`.

### How a cell's final border is decided

For a header cell the renderer merges, in order of increasing precedence:

1. the **zone** border (`header` / `spanning`, reduced to top-on-first /
   bottom-on-last by the outer frame);
2. the automatic **group underline** (a multi-column spanning cell, drawn only
   where the column grouping changes below it -- i.e. the next header row
   subdivides the span);
3. `col_spec[[col]]$border` for the cell's leftmost column;
4. `col_cell(border = )` on the cell itself.

Later steps override earlier ones side-by-side, so a `"none"` side at step 3 or
4 erases a line drawn by step 1 or 2, and a real line at step 4 replaces one
from step 3. That single rule -- *last writer wins, per side* -- is all you need
to reason about any combination.


## Level 4 -- restyling an existing table: the styling verbs

Everything above configures a table you are *building*. When the table already
exists -- typically because `as_rtftables(x, read_meta = TRUE)` extracted the
labels, alignment and spanning from a gt / gtsummary / rtables object and only
one rule needs to change -- reach for the **post-hoc styling verbs** instead of
rebuilding the header:

```{r verbs}
df <- data.frame(Item = c("Age", "Sex"), A = c("1", "2"),
                 B = c("3", "4"), C = c("5", "6"))
hdr <- rtf_col_header(
  list(col_cell(1, ""), col_cell(c(2, 4), "Treatment Group")),
  list(col_cell(1, ""), col_cell(c(2, 4), "(N = 254)")),
  c("Item", "Placebo", "Drug A", "Drug B")
)
tbl <- rtftable(df, col_header = hdr, border = "tfl")

tbl <- tbl |>
  # solid rule above -- and no rule below -- the "(N = 254)" span only
  style_header(row = 2, cols = 2:4,
               border = rtf_border(top    = TRUE,
                                   bottom = "none")) |>
  # rule under one BODY row (a summary/total row, selected by predicate)
  style_body(rows = ~ Item == "Sex",
             border = rtf_border(bottom = TRUE)) |>
  # double rule under the whole table
  style_zone(last_row = rtf_border(bottom = "double"))
```

Body borders follow the same two-step chain as everything else: the row's
zone border (`body`, crossed with `first_row` / `last_row`) first, then the
per-cell override from `style_body(border = )` -- last writer wins, per side.
Restrict `cols` to rule only part of a row, or pass an
`"none"` side to erase a zone rule on selected cells.

`style_header(row =, cols =)` addresses header cells by header-row index and
data-column position; `style_cols()`, `style_body()` and `style_zone()` cover
the column, body-cell and zone layers the same way, and `add_header_row()`
inserts a header row after the fact. Every verb also accepts the **page list**
returned by `as_rtftables()` (the edit is applied to every page), so a
converted multi-page table can be restyled in one pipe:

```r
pages <- as_rtftables(gt_obj, read_meta = TRUE) |>
  style_header(row = 2, cols = 2:4,
               border = rtf_border(top    = TRUE,
                                   bottom = "none"))
```

The merge semantics are the ones this article already taught: last writer
wins, per side. See `?style_header` for the full family, including how a
character label row is promoted to cells when it needs a per-cell border.

<!-- where-next -->

## Where next

- [Output options](output.html) --- what the rendered file contains

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.


## Before and after 0.5.0

Two things changed at 0.5.0, and they compound, so here is every common case
written both ways. `s` is an `TRUE` throughout.

First, `rtf_table_border()` is deprecated: a border is aimed with
`rtftable(border = )` for the whole table, or `style_zone()` for one kind of
row. Second, an edge now always means the **outer** edge of the selection, so
the rules *inside* it are asked for by name -- `inside_h` between the rows,
`inside_v` between the cells.

| What you want | Before (≤ 0.4.x) | Now (0.5.0) |
|---|---|---|
| Rules above and below the column header | `rtf_table_border(header = rtf_border(top = s, bottom = s))` | `style_zone(header = rtf_border(top = s, bottom = s))` |
| A rule under the last data row | `rtf_table_border(last_row = rtf_border(bottom = s))` | `style_zone(last_row = rtf_border(bottom = s))` |
| A rule under **every** data row | `rtf_table_border(body = rtf_border(bottom = s))` | `style_zone(body = rtf_border(inside_h = s))` |
| A vertical rule at every column boundary | `rtf_table_border(body = rtf_border(left = s, right = s))` | `style_zone(body = rtf_border(left = s, right = s, inside_v = s))` |
| A grid around every data cell | `rtf_table_border(body = rtf_border_box())` | `style_zone(body = rtf_border(top = s, bottom = s, left = s, right = s, inside_h = s, inside_v = s))` |
| An outer frame only | *(four `style_header()` / `style_body()` calls on the edge columns)* | `rtftable(border = rtf_border(top = s, bottom = s, left = s, right = s))` |
| Frame plus a rule under every row | *(same four calls, plus `body`)* | `rtftable(border = rtf_border(top = s, bottom = s, left = s, right = s, inside_h = s))` |
| The clinical TFL preset | `border = "tfl"` | `border = "tfl"` -- unchanged |
| A border on one cell | `col_cell(border = rtf_border(bottom = s))` | unchanged -- a cell has no inside |

And the constructors that folded into `rtf_border()` itself:

| Before (≤ 0.5.x) | Now |
|---|---|
| `rtf_border_side()` | `TRUE` |
| `rtf_border_side("none")` | `"none"` (or `FALSE`) |
| `rtf_border_side("double", 20L)` | side `"double"` plus `width = 20L` |
| `rtf_border_none()` | `rtf_border()` |
| `rtf_border_top()` | `rtf_border(top = TRUE)` |
| `rtf_border_bottom()` | `rtf_border(bottom = TRUE)` |
| `rtf_border_box()` | `rtf_border(all = TRUE)` |
| `rtf_border_with(b, bottom = x)` | `rtf_border(from = b, bottom = x)` |
| `rtf_border_tfl()` | `border = "tfl"`, or `rtf_table_style_tfl()` |

All of these still work and warn once per session. They are scheduled for
removal before the CRAN submission.

The first two rows change spelling but not meaning. Rows three to five change
meaning as well: those are the ones where an edge used to be uniform.

```{r migration, eval = FALSE}
s <- TRUE

# was: rtf_table_border(body = rtf_border(bottom = s))
tbl <- rtftable(df, border = "none") |>
  style_zone(body = rtf_border(inside_h = s))

# was: rtf_table_border(body = rtf_border_box())
tbl <- rtftable(df, border = "none") |>
  style_zone(body = rtf_border(top = s, bottom = s, left = s, right = s,
                               inside_h = s, inside_v = s))
```

Because the two readings are spelled identically, rtfreporter cannot rewrite
your code for you -- but it can tell you where to look. It warns **once per
session** when it meets a border that would have rendered differently before,
and naming `inside_h` / `inside_v` says which reading you mean and silences it.
Use `"none"` when the answer is "no rule there".
