---
title: "Adding a table-object adapter"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p align="right">
  <a class="btn btn-outline-secondary btn-sm" href="extending-adapters-ja.html">In Japanese</a>
</p>

rtfreporter can read tables produced by other packages — gt, gtsummary,
rtables/tern, flextable, and huxtable — and turn them into RTF while preserving
their metadata (column labels, alignment, spanning headers, titles, footnotes,
per-cell styles).  This article shows how to teach it about **a new
table-object type**, using the existing `gt_adapter.R`, `rtables_adapter.R`,
`flextable_adapter.R` and `huxtable_adapter.R` as templates.

Read [Architecture & internals](architecture.html) first for the big
picture.

## The contract

An adapter has one job: convert your object into the common **kwargs
list** that [`as_rtftables()`](../reference/as_rtftables.html) consumes.
Every field except `data` is optional.

```r
list(
  data            = <data.frame>,   # REQUIRED: the rendered body (character cells are fine)
  col_header      = <chr vector | stacked spanning rows>,
  col_spec        = list(list(col = 1L, align = "left"), ...),
  cell_styles     = <list, one element per data row, or NULL>,
  titles_block    = <chr vector | NULL>,   # becomes the page title
  footnotes_block = <chr vector | NULL>    # becomes the page footnote
)
```

`as_rtftables()` does everything else (pagination, per-page `cell_styles`
slicing, replicating shared metadata, attaching titles/footnotes).  You do
**not** build the `rtftable` yourself.

## Step 1 — a detector

Add a cheap, dependency-light predicate.  Keep it side-effect free.

```r
# R/mypkg_adapter.R
.is_mypkg_tbl <- function(x) inherits(x, "mypkg_table")
# (for an S4 class use: isS4(x) && methods::is(x, "MyClass"))
```

## Step 2 — token resolution

Tokens let users opt in/out of metadata via `read =`.  Mirror the gt /
rtables resolvers.

```r
.MYPKG_TOKENS_ALL <- c("col_header", "alignment", "titles", "footnotes")

.resolve_mypkg_tokens <- function(read) {
  if (is.null(read) || isFALSE(read)) return(character(0))
  if (isTRUE(read))                    return(.MYPKG_TOKENS_ALL)
  if (!is.character(read))
    stop("`read` must be FALSE/TRUE or a character vector of tokens.",
         call. = FALSE)
  bad <- setdiff(read, .MYPKG_TOKENS_ALL)
  if (length(bad))
    stop("Unknown token(s): ", paste(bad, collapse = ", "), call. = FALSE)
  read
}
```

## Step 3 — the extractor

Return the kwargs list.  Always require the heavy package lazily.

```r
.mypkg_to_rtftable_kwargs <- function(x, tokens = .MYPKG_TOKENS_ALL) {
  if (!requireNamespace("mypkg", quietly = TRUE))
    stop("Reading a mypkg table requires the `mypkg` package.", call. = FALSE)

  out <- list()
  out$data <- mypkg::as_body_data_frame(x)          # however your package exposes the body

  if ("col_header" %in% tokens)
    out$col_header <- mypkg::column_labels(x)

  if ("alignment" %in% tokens)
    out$col_spec <- lapply(seq_len(ncol(out$data)), function(j)
      list(col = j, align = mypkg::col_align(x, j)))

  if ("titles" %in% tokens)
    out$titles_block <- mypkg::title_lines(x)

  if ("footnotes" %in% tokens)
    out$footnotes_block <- mypkg::footnote_lines(x)

  out
}
```

Tips drawn from the existing adapters:

* **Spanning headers** are a multi-row `col_header`: a list whose upper
  elements are `list(col_cell(pos = c(from, to), label = ...), ...)` and
  whose last element is the leaf-label character vector.  See
  `.extract_spanners()` (gt) and `.rtables_header_segments()` (rtables).
* **Per-cell styles** go in `cell_styles`: a list with one element per data
  row, each either `NULL` or
  `list(bold=, italic=, underline=, indent_twips=)` where each is a vector
  of length `ncol` with `NA` meaning "inherit from `col_spec`".
* **Footnote marks already in the text.** If your source renders marks
  inline (gt uses `<sup>N</sup>`, rtables uses `{N}`), convert them to
  rtfreporter's `^{N}` superscript markup **without `gsub` backreferences**
  (some locale-broken R builds drop `\\1`) — use `regmatches()` as in
  `.convert_footnote_marks()` / `.convert_rtables_marks()`.

## Step 4 — wire it into `as_rtftables()`

Three small edits in `R/as_rtftables.R`:

```r
# (a) the list-input guard must NOT treat your object as a list of items
if (is.list(x) && !is.data.frame(x) && !isS4(x) &&
    !.is_gt_tbl(x) && !.is_gtsummary_tbl(x) &&
    !.is_rtables_tbl(x) && !.is_mypkg_tbl(x)) { ... }

# (b) the resolve-body-+-metadata dispatch
} else if (.is_mypkg_tbl(x)) {
  tokens          <- .resolve_mypkg_tokens(read)
  kw              <- .mypkg_to_rtftable_kwargs(x, tokens = tokens)
  body            <- kw$data
  cell_styles     <- kw$cell_styles
  titles_block    <- kw$titles_block
  footnotes_block <- kw$footnotes_block
}
```

Optionally relax the guard in `as_rtftable()` (the singular wrapper) so it
accepts your type too.

## Step 5 — declare the dependency

Add `mypkg` to `Suggests:` in `DESCRIPTION`.  Never add it to `Imports:` —
the zero-hard-dependency rule is deliberate.  The lazy
`requireNamespace()` in Step 3 produces a friendly error when it is
missing.

## Step 6 — tests and docs

* Add `tests/testthat/test-mypkg-adapter.R`, wrapping every test in
  `skip_if_not_installed("mypkg")`.
* Document the new `read` tokens (extend `?as_rtftables`).
* Add a `NEWS.md` entry; for a major/minor change update `CHANGELOG.md`.
* Run `devtools::document()`, `devtools::test()`, `devtools::check()`.

## Supplying a method from your own package

The steps above are for a package that does **not** know rtfreporter: gt,
rtables and the rest never import it, so rtfreporter carries their adapters.
If you are writing a package that **depends on** rtfreporter, do not add a
branch here.  `as_rtftables()` is an S3 generic, so register a method for your
class instead, and rtfreporter never has to name it:

```{r}
# In your package's R/ (roxygen), with rtfreporter in Imports:
#' @importFrom rtfreporter as_rtftables
#' @export
as_rtftables.my_report <- function(x, ...) {
  df <- build_the_table(x)          # your own work
  rtfreporter::as_rtftables(df, ...) # a list of rtftable pages
}
```

This writes `importFrom(rtfreporter, as_rtftables)` and
`S3method(as_rtftables, my_report)` to your NAMESPACE.  The method must return
what the default returns: a list of `rtftable` pages.  `rtf_tables(doc, x)`
then accepts your object directly, converting it up front the way it converts
a `gt_tbl`.  The dependency points the right way: your package depends on
rtfreporter, and rtfreporter does not depend on yours.

## A worked reference

The rtables adapter (`R/rtables_adapter.R`) is a compact, complete example
that exercises every part of the contract — multi-level spanning headers,
per-row indentation as `cell_styles`, inline `{N}` -> `^{N}` mark
conversion, and title/footer extraction — all from a single
`formatters::matrix_form()` call.  It is a good model to copy.
