---
title: "External API specification"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p align="right">
  <a class="btn btn-outline-secondary btn-sm" href="external-api-ja.html">日本語</a>
</p>

This is the contributor-facing reference for rtfreporter's **public API**: the
exported functions, the **S3-class objects they return**, and the pipe flow that
ties them together. For the big picture see
[Architecture & internals](architecture.html); for the internal data structures
see [Internal class design (S3)](internal-design.html).

> rtfreporter is **pure S3** with no hard runtime dependencies. Every public
> object is a plain `list` carrying an S3 `class` attribute, so results are
> `dput()`-able, easy to inspect with `str()`, and serialise without surprises.

## Terminology

"S3 functions" can be ambiguous, so to be precise:

* **Constructor functions** — exported functions that build and return an
  **object of an S3 class** (e.g. `rtftable()` returns an object of class
  `"rtftable"`). These are most of the public API below.
* **S3 generics / methods** — `print()` and `plot()` dispatch on the object's
  class (e.g. `print.rtftable()`).

When this document says "the object returned by `f()`", it means the S3-class
object a constructor produces -- not a generic.

## The pipe flow

A report is built by piping an `rtf_document` through content and section calls,
then rendering:

```{r flow}
library(rtfreporter)

rtf_document() |>                                  # -> rtf_document
  rtf_tables(as_rtftables(df)) |>                  # add a content page
  rtf_section(page = 1, secinfo = list(            # assign a header/footer
    header = rtf_header(...), footer = rtf_footer(...))) |>
  generate_rtfreport("out.rtf", overwrite = TRUE)  # render to a file
```

`rtf_document()` and the `rtf_*()` pipe verbs return a **new** `rtf_document`
(copy-on-modify); nothing is mutated in place.

## Public classes and their constructors

| Class | Constructor(s) | Role |
|-------|----------------|------|
| `rtf_document` | `rtf_document()` | The document being assembled (pages + sections + document-wide settings). |
| `rtftable` | `rtftable()`, `as_rtftable()`, `as_rtftables()` | One table (or one page of a paginated table). |
| `rtfplot` | `rtfplot()` | One embedded figure. |
| `rtf_border` | `rtf_border()`, `rtf_border_side()`, `rtf_border_top()` / `_bottom()` / `_box()` / `_none()`, `rtf_border_with()` | A border for a selection: its outer edges plus the rules between the rows (`inside_h`) and between the cells (`inside_v`) inside it. What it applies to is decided by where you attach it. |
| `rtf_table_border` | `rtf_table_border()` (**deprecated**), `rtf_border_tfl()` | The five-zone map. Superseded: attach an `rtf_border()` to `rtftable(border = )` for the whole table, or to `style_zone()` for one kind of row. |
| `rtf_table_style` | `rtf_table_style()`, `rtf_table_style_with()`, `rtf_table_style_tfl()` | A reusable style (borders + padding + row-height defaults), captured by snapshot at construction. |
| `rtf_col_header` | `rtf_col_header()`, `col_cell()`, `add_col_header_row()` | Multi-row / spanning column headers. |

Not every public function is a constructor. `rtf_config()` is a **configurator**,
not a constructor: it takes an `rtf_document`, updates only the non-`NULL`
fields, and returns a copy of the **same** `rtf_document` class (it sets no new
class) -- just like the other pipe verbs (`rtf_tables()`, `rtf_section()`, …).
Likewise `rtf_header()` / `rtf_footer()` return plain (unclassed) lists
describing a section's header / footer, and `update_header_row()` /
`update_footer_row()` return an updated copy.

> **`rtf_document` vs `rtfreport`.** `rtf_document` is the *public* object you
> build and hold via the pipe. It is converted to a separate *internal* class,
> `rtfreport`, only at render time (by `generate_rtfreport()`); see
> [Internal class design (S3)](internal-design.html). You do not construct or
> handle `rtfreport` directly.

## Content, sections and rendering

* **Content** — `rtf_tables(doc, tables, ...)` and `rtf_figures(doc, figures, ...)`
  append content pages (one element = one page); `rtf_titles()` / `rtf_footnotes()`
  attach per-page title/footnote blocks. Each returns the modified `rtf_document`.
* **Sections** — `rtf_section(doc, page, secinfo)` assigns a header/footer to a
  page range (an overlay on the flat page list; `NULL` inherits the previous
  section).
* **Render** — `generate_rtfreport(doc, file, overwrite)` writes the `.rtf` and
  returns the path invisibly.

## Importing and paginating tables

`as_rtftables(x, ...)` is the unified entry point that converts a `gt` /
gtsummary / rtables-`VTableTree` / `data.frame` / tibble (or a list of these)
into a **list of `rtftable` page objects**, reading the source metadata and
paginating in one call. `as_rtftable(x, ...)` is the single-page form (one
`rtftable`). Pagination strategies are chosen by name (`split = "group_safe"` and the
rest, with their settings passed alongside) and a custom `split=` function is
accepted; blank separator rows are
controlled by `blank_rows` / `set_blank_rows()` (and may be counted toward
`max_rows` with `count_blank_rows = TRUE`). The body can also be **ordered**
before the split (`sort_by` / `sort_desc`), **grouped** on a chosen column
(`group_col` / `group_by`, plus `collapse_repeats` for repeat suppression), and
**columns hidden** from the printed pages while still driving that
grouping/ordering (`drop_cols` — e.g. a sort-key carrier column). All of these
share the input body's column coordinates; see
[Paginating with `as_rtftables()`](pagination.html) for the full treatment.

## Output helpers

* `rtf_replace_text()` — post-processing find/replace on a rendered `.rtf`.
* `assemble_rtf()` (and `assemble_files()` / `assemble_folder()` /
  `assemble_spec()` / `assemble_from_spec()` / `assemble_toc()`,
  `toc_heading()` / `toc_entry()`) — combine several RTFs into one deliverable
  with a table of contents and bookmarks.

## S3 generics

`print()` methods exist for the document and the spec objects; `plot()` methods
give base-graphics wireframes of borders / tables / the document for
sanity-checking a layout before rendering. `print()` on an `rtftable` renders a
visual preview of the table -- the laid-out cells with column (and spanning)
headers, per-column alignment and the border rules -- so you can eyeball the
content in the console (`format()` returns those lines, `summary()` the compact
metadata block). See the
[function reference](../reference/index.html) for the full per-function detail.
