---
title: "Figures: from a plot object to a page of the report"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
options(width = 100)
set.seed(20260909)
```

```{r setup}
library(rtfreporter)
```

A figure is a content page like any other: one figure per page, its own titles
and footnotes, the section's running header and footer around it. What is
particular to a figure is **size** -- a table fills the width it is given,
while a figure has a size of its own and has to be told what to do about it.

This article is about that: getting a plot into the document, deciding how big
it is on the page, and knowing what the file underneath is.

## A plot object goes straight in

`rtf_figures()` takes the plot itself. There is no file to write first:

```{r ggplot-figure}
library(ggplot2)

dat <- data.frame(
  week = rep(0:12, 2),
  arm  = rep(c("Placebo", "Drug"), each = 13),
  mean = c(seq(0, 6,  length.out = 13) + rnorm(13, 0, 0.2),
           seq(0, 11, length.out = 13) + rnorm(13, 0, 0.2))
)

p <- ggplot(dat, aes(week, mean, colour = arm)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Study week", y = "Mean change from baseline", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

doc <- rtf_document() |>
  rtf_figures(list(p),
              titles = list(c("Figure 14.2.1",
                              "Mean Change from Baseline by Week",
                              "Full Analysis Set")),
              footnotes = list("Error bars omitted for clarity."))

length(doc$contents)      # one figure page
```

`rtfplot()` is the same thing one step lower down, for when one figure in a
list of several needs settings of its own:

```{r rtfplot-object}
rtfplot(p)
```

### What counts as a plot

Dispatch is by what the object can do, not by which package made it, so the
list is short and open-ended:

| what you pass | how it is drawn |
|---|---|
| a **ggplot2** plot, a **lattice** trellis object, a **patchwork** | printed |
| a **grid** grob or `gtable` | `grid::grid.draw()` |
| a base plot recorded with `grDevices::recordPlot()` | replayed |
| a **function of no arguments** that draws | called |
| a path to a **PNG or JPEG** file | read as it stands |

The function is the escape hatch, and the natural way to give base graphics --
which draw as a side effect and leave no object to pass:

```{r function-figure}
fig <- rtfplot(function() {
  boxplot(mpg ~ cyl, data = mtcars, xlab = "Cylinders", ylab = "MPG")
})
fig
```

Anything a package can draw, it can draw inside that function.

## Size: inches on the page, pixels in the file

Two sizes are easy to confuse, so the arguments keep them apart.

**`render_width` / `render_height` (inches) say how big the figure is on the
page.** They default to 6.5 x 4.5 in, which fits a portrait letter page inside
one-inch margins.

**`render_dpi` says how sharp it is, and nothing else.** It multiplies the
pixels in the file; it does not change the inches:

```{r dpi-vs-size}
sizes <- lapply(c(150, 300, 600), function(d) {
  f <- rtfplot(function() plot(1:10), render_width = 4, render_height = 3,
               render_dpi = d)
  data.frame(dpi     = d,
             pixels  = paste(f$img_width, "x", f$img_height),
             on_page = sprintf("%.1f x %.1f in", f$img_width / d,
                               f$img_height / d))
})
do.call(rbind, sizes)
```

Three files of very different weight, one size on the page. Pick `render_dpi`
for the medium: 300 for something that will be printed, 150 for a review copy,
600 for a figure with fine hatching.

### Fitting the page

The native size is **not** capped to your page, because a figure's size is a
decision -- a plot squeezed to fit usually reads worse than one drawn at the
right size to begin with. So draw it at the size it should occupy:

```{r page-fit}
# A4 landscape is 11.69 x 8.27 in; the writable width is what the margins
# leave.
writable_in <- 11.69 - 0.75 - 0.75
writable_in

rtfplot(p, render_width = writable_in, render_height = 4.5)
```

That figure now spans the writable width of

```{r page-fit-page}
rtf_page(paper_size = "A4", orientation = "landscape",
         margin_left_in = 0.75, margin_right_in = 0.75)
```

To scale a figure you already have -- a file from someone else, or a plot you
do not want to redraw -- give `width_twips` instead. The height follows the
aspect ratio:

```{r scale-twips}
rtfplot(p, width_twips = as.integer(writable_in * 1440))
```

`width_twips` and `height_twips` are display size in twips (1 in = 1440), the
unit the rest of the package uses. `align` puts the figure `"center"`
(default), `"left"` or `"right"` on the page.

## From a file

The path form is unchanged, and it is what you want when the graphic comes
from somewhere else -- a validated output from another system, a diagram
someone drew:

```{r file-figure}
img <- tempfile(fileext = ".png")
png(img, width = 2400, height = 1500, res = 300)
plot(1:10, type = "l", xlab = "Visit", ylab = "Mean")
invisible(dev.off())

rtfplot(img)
```

A file carries its own resolution, so **its native size is what the file says
it is**: 2400 x 1500 px at 300 dpi is 8 x 5 in. That is read from the PNG's
`pHYs` chunk or the JPEG's JFIF density. A file recording none is assumed to
be at `rtfreporter.figure.default_dpi` (factory `96`), which is worth knowing
when a figure comes out unexpectedly large:

```{r default-dpi}
rtfreporter_options()[["rtfreporter.figure.default_dpi"]]
```

A drawn object has no such doubt -- the package drew it, so it carries the
`render_dpi` it was drawn at whatever the graphics device recorded in the
file.

### Formats

**PNG and JPEG only.** RTF's picture keywords include `\emfblip`,
`\wmetafile`, `\dibitmap` and `\macpict`; `rtfreporter` emits `\pngblip` and
`\jpegblip`, and refuses the rest rather than embedding something a viewer may
not render:

```{r formats, error = TRUE}
svg <- tempfile(fileext = ".svg")
file.create(svg)
rtfplot(svg)
```

For a clinical deliverable that is the right pair: PNG for anything with text
or lines (lossless, and the usual choice for a plot), JPEG for a photograph.
A vector format would print more sharply, but embedded EMF is not rendered the
same way by every word processor, and a figure that differs between reviewers'
machines is worse than one that is 300 dpi.

## A report of figures and tables

They share the page model, so a document is a sequence of content pages:

```{r mixed}
report <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(c(l = "Protocol XYZ-123",
                                      r = "Page {AUTO_PAGE}")))
  )) |>
  rtf_tables(list(head(mtcars[, 1:4], 8)),
             titles = list(c("Table 14.1.1", "Vehicle characteristics"))) |>
  rtf_figures(list(p, function() hist(mtcars$mpg, main = NULL, xlab = "MPG")),
              titles = list(c("Figure 14.2.1", "Mean change by week"),
                            c("Figure 14.2.2", "Distribution of MPG")))

out <- file.path(tempdir(), "figures.rtf")
generate_rtfreport(report, out, overwrite = TRUE)

c(pages = length(report$contents), bytes = unname(file.info(out)$size))
```

Three pages -- one table, two figures -- under one running header. The images
are embedded in the `.rtf` itself, so there is no folder of files to keep
beside it, which is much of the point of the format for a deliverable that
gets emailed.

## Where next

- [Adding content](adding-content.html) --- titles, footnotes and the page model
- [Headers and footers](headers-footers.html) --- the running header around a figure
- [Rendering & assembly](output.html) --- combining figure and table documents into one file

See `?rtfplot` and `?rtf_figures` for the full argument reference.
