---
title: "Importing tables with as_rtftables()"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
.has_gt <- requireNamespace("gt", quietly = TRUE)
.has_flextable <- requireNamespace("flextable", quietly = TRUE)
.has_huxtable <- requireNamespace("huxtable", quietly = TRUE)
```

```{r setup}
library(rtfreporter)
```

`as_rtftables()` is the single entry point for turning a **table object** into
the `rtftable` page objects that `rtf_tables()` consumes. It does two jobs:

1. **Read** the source table -- its rendered body plus the metadata the RTF
   renderer can use (column labels, alignment, spanning headers, titles,
   footnotes, …);
2. **Paginate** -- split the body into one `rtftable` per page.

This article covers the **reading / conversion** side. Pagination has its own
guide: see [Paginating with as_rtftables()](pagination.html).

```{r intro}
df <- data.frame(
  Parameter = c("Age (years)", "  Mean", "  SD"),
  Placebo   = c("", "75.1", "8.2"),
  Active    = c("", "74.4", "7.9"),
  stringsAsFactors = FALSE
)

pages <- as_rtftables(df)     # a list of rtftable page objects
length(pages)
class(pages[[1]])
```


## Supported inputs

`as_rtftables()` accepts any of:

- a **`gt`** table (`gt::gt(...)`),
- a **`gt_group`** -- gt's multi-table container from `gt::gt_group()` /
  `gt::gt_split()`, and what **tfrmt**'s `print_to_gt()` returns when the spec
  has a `page_plan`. Each member table becomes its own page set, so a
  paginated tfrmt render can be handed over as-is,
- a **gtsummary** table (`tbl_summary()`, `tbl_regression()`, …; converted via
  `gtsummary::as_gt()`). A **`tbl_split`** container
  (`tbl_split_by_rows()` / `tbl_split_by_columns()`) is likewise expanded,
  one member per page set,
- an **rtables / tern** `VTableTree`,
- a **`flextable`** table (`flextable::flextable(...)`),
- a **`huxtable`** table (`huxtable::huxtable(...)`),
- a plain **`data.frame` / tibble**, or
- a **`list`** of any of these (flattened to one page each, names propagated).

It always returns a **list** of `rtftable` objects. For a single table where you
do not need a list, use the singular convenience wrapper `as_rtftable()` -- it is
exactly `as_rtftables(x, split = "none")[[1]]`:

```{r singular}
tbl <- as_rtftable(df)        # one rtftable, not a list
class(tbl)
```

Being a single-table wrapper, `as_rtftable()` accepts a `gt_group` or a
gtsummary `tbl_split` only when it holds exactly one table; a multi-member
container errors and points you to `as_rtftables()`.


## What the columns are called (and how to check) {#column-names}

Several `as_rtftables()` arguments select a column **by name or position** --
`group_col` (grouping), `sort_by` (ordering), `drop_cols` (hiding) and
`collapse_repeats` (repeat suppression), all covered in
[Paginating with `as_rtftables()`](pagination.html). They all act on the
**rendered body** -- the intermediate `data.frame` that `as_rtftables()` builds
-- *not* on the visible column headers. **What that body's columns are called
depends on the source**, and this is the single most common stumbling block, so
it is worth knowing the rule up front.

| Source | Body column names | Naming rule |
|--------|-------------------|-------------|
| `data.frame` / tibble | your own column names | exactly what you named them |
| **gt** (`gt(df)`) | gt's column **ids** (default: the original data names) | follows gt |
| **gtsummary** | `label`, `stat_1`, `stat_2`, … | gtsummary's internal ids |
| **tfrmt** | `rowname`, then the `column` values | gt stub id + your `column` levels |
| **rtables / tern** | `V1`, `V2`, `V3`, … | **positional**, content-independent |
| **flextable** | `V1`, `V2`, `V3`, … | positional |
| **huxtable** | `V1`, `V2`, `V3`, … | positional |

Two families fall out of this:

- **gt-based sources (gt, gtsummary, tfrmt)** keep meaningful **id** names -- but
  each family's convention differs (gt = the data names, gtsummary =
  `label` / `stat_N`, tfrmt = `rowname` + your column levels).
- **rtables / flextable / huxtable** discard the names and renumber the columns
  `V1`, `V2`, … -- the name tells you nothing about the content, only the
  position.

### The universal check

Whatever the source, the body column names are exactly
`names(as_rtftables(x)[[1]]$data)` -- the same thing the selection arguments see.
When in doubt, print it:

```{r names-df}
names(as_rtftable(df)$data)            # a data.frame keeps its own names
```

```{r names-gtsummary, eval = requireNamespace("gtsummary", quietly = TRUE)}
library(gtsummary)
tbl <- tbl_summary(trial[c("age", "grade")], by = NULL)
names(as_rtftables(tbl)[[1]]$data)     # gtsummary -> "label", "stat_0"
```

Source packages also offer their own inspectors: **gtsummary**'s
`show_header_names()` lists the internal column names (`label`, `stat_1`, …) that
you pass here, and **gt**'s `gt::extract_body()` shows the rendered body with its
ids.

### Two practical rules

1. **Use the *id*, not the visible header.** A gtsummary `stat_1` column may
   display as "Drug (N = …)", but you must select it as `"stat_1"`. Passing the
   on-screen label fails with "column not found".
2. **Integer indices are the most portable.** Because rtables / flextable /
   huxtable expose only `V1`, `V2`, …, and the gt families differ, a 1-based
   **position** works identically across every source:

   ```r
   as_rtftables(x, group_col = 1, sort_by = 1, drop_cols = 1)  # "the 1st column"
   ```

   Reach for names when they are stable and readable (your own `data.frame`, or a
   known gtsummary layout); reach for indices when writing source-agnostic code.


## Reading metadata -- `read_meta`

For a rich source (gt / gtsummary / rtables / flextable / huxtable),
`as_rtftables()` reads the render-relevant metadata so you do not have to
restate it. What is carried:

| Metadata | gt / gtsummary | rtables / tern | flextable | huxtable |
|----------|----------------|----------------|-----------|----------|
| Column (leaf) labels | yes | yes | yes | yes |
| Per-column alignment | yes | yes | yes | yes |
| Spanning headers | yes | yes | yes | yes |
| Column widths | yes (px/pct) | -- | -- | -- |
| Title + subtitle | yes | yes | yes (caption) | yes (caption) |
| Footnotes / source notes | yes | yes | yes (footer) | -- |
| In-cell footnote marks | yes (superscript) | yes (superscript) | -- | -- |
| Row-group rows + indent | yes (rendered) | yes (rendered) | yes (rendered) | yes (rendered) |

For **flextable** the *displayed* text is read (header labels from
`set_header_labels()` and `colformat_*()` formatting included), not the raw
`$body$dataset`. Image/equation cells and `footnote()` reference marks are not
carried.

For **huxtable** the *displayed* text is read with its `number_format` applied;
header rows (`header_rows()`) become the column header, `colspan()` becomes the
spanning header, and the `caption()` becomes the page title. huxtable has no
footnote concept, so none is extracted.

**Not carried** (RTF cannot reproduce them, so they are intentionally ignored):
per-cell bold / italic / underline, cell background fills, font/size styling, and
Markdown inside labels or titles. A plain `data.frame` / tibble carries two
pieces of metadata: column **`label` attributes** become header labels (see
*Finishing a plain data.frame* below), and **delimited column names** are
reconstructed into a spanning header (see `header_sep`); everything else
(`col_spec`, widths, ...) you set yourself (see *Pass-through* below).

`read_meta` controls the extraction:

- `TRUE` (default) -- read everything in the table above.
- `FALSE` -- use only the rendered body (ignore labels / titles / notes).
- a **character vector of tokens** -- selective opt-in, e.g.
  `c("col_header", "alignment")`. Tokens for gt/gtsummary: `"col_header"`,
  `"alignment"`, `"spanning"`, `"widths"`, `"titles"`, `"footnotes"`. For
  rtables/tern: `"col_header"`, `"alignment"`, `"spanning"`, `"titles"`,
  `"footnotes"`, `"indent"`, `"footnote_marks"`. For flextable:
  `"col_header"`, `"alignment"`, `"spanning"`, `"titles"`, `"footnotes"`. For
  huxtable: `"col_header"`, `"alignment"`, `"spanning"`, `"titles"`. For a
  plain data.frame / tibble: `"labels"` (column `label` attributes as header
  labels); unknown tokens are *ignored* for data.frames, so a `list()` mixing
  table objects and data.frames can share one `read_meta`.

```{r gt-meta, eval = .has_gt}
library(gt)

g <- gt(head(mtcars[, c("mpg", "cyl")], 3)) |>
  cols_label(mpg = "Miles/gallon", cyl = "Cylinders") |>
  cols_align("right", columns = c(mpg, cyl))

tbl <- as_rtftable(g)               # reads the labels + alignment
tbl$col_header                       # -> "Miles/gallon", "Cylinders"
```

```{r gt-meta-off, eval = .has_gt}
# read_meta = FALSE keeps only the rendered body and ignores the labels, so no
# col_header metadata is extracted (the body's own column names are used).
plain <- as_rtftable(g, read_meta = FALSE)
plain$col_header        # NULL -- nothing extracted
names(plain$data)       # the body column names that headers fall back to
```

A **flextable** is read the same way. The *displayed* text is taken (so
`set_header_labels()` relabelling and `colformat_*()` formatting are honoured),
along with the caption (title), footer lines (footnotes), spanning headers and
per-column alignment:

```{r flextable-meta, eval = .has_flextable}
library(flextable)

ft <- flextable(data.frame(Param = c("Mean", "SD"), Value = c(75.13, 8.234))) |>
  set_header_labels(Param = "Characteristic") |>
  colformat_double(j = "Value", digits = 1) |>
  align(j = "Value", align = "right", part = "body") |>
  set_caption("Table 1: Demographics")

tbl <- as_rtftable(ft)
tbl$data[[2]]            # -> "75.1", "8.2"  (colformat applied)
attr(tbl, "rtf_titles") # -> "Table 1: Demographics"  (from the caption)
```

A **huxtable** works the same way -- its `number_format` is applied, header
rows become the column header, and the caption becomes the title:

```{r huxtable-meta, eval = .has_huxtable}
library(huxtable)

hx <- huxtable(
  Characteristic = c("Characteristic", "Age (years)"),
  Value          = c("Value", 75.134),
  add_colnames = FALSE
)
header_rows(hx)[1]    <- TRUE          # row 1 is the column header
number_format(hx)[2, 2] <- "%.1f"      # 75.134 -> 75.1
caption(hx) <- "Table 1: Demographics"

tbl <- as_rtftable(hx)
tbl$data[[2]]            # -> "75.1"  (number_format applied)
attr(tbl, "rtf_titles") # -> "Table 1: Demographics"  (from the caption)
```


## Monospaced alignment -- `cell_format`

Clinical tables often need count/percent columns to line up by character.
`cell_format` re-formats the body **before** pagination, column by column. Pass a
single function (applied to every data column; the row-label column 1 is left
alone) or a list of functions taken positionally. rtfreporter ships a few:

- `fmt_count_paren()` -- right-justify an integer count and the number inside the
  parentheses, e.g. `"10 (11.6%)"` over `" 4 ( 4.7%)"`;
- `fmt_right_align()` -- the minimal "right-justify a column" formatter and a
  template for writing your own;
- `align_count_pct = TRUE` -- the long-standing shorthand for the built-in
  `"n (xx.x)"` realigner.

```{r cellfmt}
counts <- data.frame(
  Group = c("A", "B"),
  n_pct = c("4 (4.7%)", "10 (11.6%)"),
  stringsAsFactors = FALSE
)
tbl <- as_rtftable(counts, cell_format = fmt_count_paren)
tbl$data$n_pct                       # both cells padded to one width
```

### Writing your own `cell_format` function

When none of the built-ins matches your data's notation, write your own. The
contract is small:

* **Signature.** The framework calls your function with **one argument** -- a
  character vector holding the cells of a single column. (The built-ins also
  accept a second `nbsp =` argument with a default, but `cell_format` never
  passes it, so a plain `function(x)` is enough.) Your function is called **once
  per column**, and is not told *which* column it is, so base every decision on
  `x` alone.
* **Return value.** A character vector of the **same length** as `x` -- one
  element per row. Returning a different length is an error; never drop or add
  rows.
* **Leave non-matching cells alone.** Cells you do not intend to touch -- empty
  group-label cells, totals, anything not matching your pattern -- must be
  returned unchanged.
* **Pad with the non-breaking space `" "`, not a normal space.** RTF and
  Word collapse runs of ordinary spaces, which would silently undo your
  alignment.
* **Which columns it sees.** A *single* function is applied to columns
  `2..ncol` (column 1 is the row label, left alone by clinical convention). To
  target specific columns, pass a **list** instead: `cell_format[[j]]` formats
  column `j`, and `NULL` entries leave a column untouched.

Here is a complete formatter that appends a unit and right-aligns a numeric
column, leaving blanks and non-numbers as they are:

```{r cellfmt-custom}
# Right-align a numeric column and append a unit (e.g. "mg"), padding on the
# left with non-breaking spaces so the values line up in a monospaced font.
fmt_with_unit <- function(x, unit = "mg", nbsp = " ") {
  x <- as.character(x)
  is_num <- grepl("^[0-9]+(\\.[0-9]+)?$", trimws(x))   # which cells to touch
  labelled <- x
  labelled[is_num] <- paste0(trimws(x[is_num]), " ", unit)
  width <- max(nchar(labelled[is_num]), 0L)            # widest labelled cell
  out <- x
  out[is_num] <- formatC(labelled[is_num], width = width)   # right-justify
  gsub(" ", nbsp, out, fixed = TRUE)                   # protect the padding
}

dose <- data.frame(
  Arm  = c("Low", "High", "Total"),
  Dose = c("5", "12.5", ""),                           # blank cell left as-is
  stringsAsFactors = FALSE
)

# A list targets one column by position; column 1 (Arm) is left untouched.
tbl <- as_rtftable(
  dose,
  cell_format = list(NULL, function(x) fmt_with_unit(x, unit = "mg"))
)
tbl$data$Dose
```

Because `cell_format` passes only the column vector, wrap a parameterised
formatter in a one-argument closure (`function(x) fmt_with_unit(x, unit = "mg")`)
to fix its extra arguments -- exactly as shown above.


## Column widths -- `auto_width`

By default columns share the width evenly. With `auto_width = TRUE` each column
is sized to its widest content (header label or data cell) so long labels do not
wrap; the widths are computed once and applied to every page:

```{r autowidth}
tbl <- as_rtftable(df, auto_width = TRUE)
tbl$column_widths_twips
```

Add `table_width_twips =` to scale those widths to a total (e.g. to fill, or fit
within, the writable page width). An explicit `column_widths_twips` /
`col_rel_width` always wins over `auto_width`.


## Pass-through to `rtftable()`

Any further argument is forwarded to [rtftable()] for **every** page, so you can
set the table''s appearance in the same call -- and your explicit values always
win over the metadata extracted from the source:

```{r passthrough}
tbl <- as_rtftable(
  df,
  border        = "tfl",
  col_header    = c("Parameter", "Placebo (N=86)", "Active (N=84)"),
  col_rel_width = c(2, 1, 1),
  row_height_twips = 280L
)
tbl$col_header
```

See the [page & document setup](page-setup.html) article for document-wide
settings, and `?rtftable` for the full list of per-table options (alignment,
per-column/-cell styles and colours, blank rows, …).


## Titles and footnotes

When the source carries a title / subtitle or source notes, `as_rtftables()`
attaches them to each page as the `rtf_titles` / `rtf_footnotes` attributes, and
`rtf_tables()` renders them automatically -- you do not pass them again:

```{r titles, eval = FALSE}
doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(as_rtftables(my_gt_table))   # gt's title/source notes flow through
```

For a plain data.frame (no metadata), set them on `rtf_tables()` with `titles =`
/ `footnotes =` instead.


## Spanning headers from delimited names -- `header_sep`

gt, gtsummary, rtables/tern, flextable and huxtable all carry real spanning
(multi-row) header metadata, which `as_rtftables()` reads directly. A **plain
data.frame** has none: the only place the nesting can live is the column *names*.
A common producer is `ydisctools::pivot_stats_wider()`, which spreads grouping
columns into headers joined by `____`:

```
group1 group2 label   cohort1____trt1 cohort1____trt2 cohort2____trt1 cohort2____trt2
Hoge   Sex    Male     11 (55.0)        9 (45.0)        13 (65.0)       7 (35.0)
Hoge   Sex    Female    9 (45.0)       11 (55.0)        7 (35.0)       13 (65.0)
```

`as_rtftables()` reconstructs the two-level header from those names
automatically -- `cohort1` / `cohort2` become spanners over their `trt1` / `trt2`
leaf columns, and the id columns (`group1`, `group2`, `label`) keep their name on
the leaf row:

```{r header-sep, eval = FALSE}
# `df` produced by ydisctools::pivot_stats_wider(); names use the "____" sep.
as_rtftables(df)            # 2-row header reconstructed by default
```

`header_sep` controls the delimiter(s). Its default recognizes **both**
`"____"` (ydisctools) and `"___tlang_delim___"` (the delimiter tfrmt puts in its
flattened column names), so a data.frame from either tool just works. Pass your
own separator(s) to override, or `NULL` to keep the plain one-row header:

```{r header-sep-custom, eval = FALSE}
as_rtftables(df, header_sep = "::")     # split on your own delimiter
as_rtftables(df, header_sep = NULL)     # disable: use names(df) verbatim
```

Adjacent columns merge into one spanning cell only when they share both the
label *and* the same ancestor path, so a `trt1` under `cohort1` never merges with
a `trt1` under `cohort2`. An explicit `col_header =` (passed through to
`rtftable()`) always wins over the reconstructed one.

With the `"____"` separator a **doubled** separator (`"________"`) splits to an
empty middle segment, i.e. a blank cell at that header level -- useful to align a
column that skips an intermediate spanner (e.g. an `Overall________n` column
sitting beside `Drug____Dose____n` columns).

This only applies to plain data.frame input; framework tables keep using their
own spanning metadata.


## Finishing a plain data.frame

A framework table (gt / rtables / ...) arrives *finished*: row-group labels are
interleaved, children are indented, headers are labelled. A tidy data.frame is
not -- it usually looks like "hierarchy columns + statistic columns". Two
finishing tools close that gap without leaving base R.

### Hierarchy columns into a stub -- `stub_cols()`

`stub_cols()` merges hierarchy columns (parent first, leaf last) into **one
stub column**: each parent value becomes its own full-width label row, and the
leaf rows under it are indented with non-breaking spaces -- the same layout the
gt / rtables / tfrmt adapters emit, produced here from plain columns.

```{r stub-cols}
ae <- data.frame(
  soc = c("", "Cardiac disorders", "Cardiac disorders",
          "Gastrointestinal disorders"),
  pt  = c("Any adverse event", "Atrial fibrillation", "Bradycardia",
          "Nausea"),
  n   = c("9 (6.3%)", "3 (2.1%)", "1 (0.7%)", "5 (3.5%)"),
  stringsAsFactors = FALSE
)

tbl <- stub_cols(ae, vars = c("soc", "pt"),
                 label = "System Organ Class / Preferred Term")
tbl
```

Note the `"Any adverse event"` summary row: its parent cell is empty (`""`),
so it gets **no label row and no indent** -- it stays flush left at the top.
Label rows hold `NA` in the statistic columns, which renders as an empty cell.

#### Group-summary rows (the AE pattern) {#stub-group-summary}

Adverse-event tables usually carry a **group-level count** as well: the total
for a System Organ Class, printed on the SOC row itself, above its indented
preferred terms. In the input that count sits on a row whose **leaf** is either
`NA` / `""` or a repeat of the parent value. `stub_cols()` folds such a row's
statistics **onto the group's label row** -- no separate indented leaf -- via
the `group_summary` argument (on by default):

```{r stub-group-summary}
ae_sum <- data.frame(
  soc = c("Cardiac disorders", "Cardiac disorders", "Cardiac disorders"),
  pt  = c(NA, "Atrial fibrillation", "Bradycardia"),   # NA leaf = SOC summary
  n   = c("4 (2.8%)", "3 (2.1%)", "1 (0.7%)"),
  stringsAsFactors = FALSE
)
stub_cols(ae_sum, vars = c("soc", "pt"))
```

The `4 (2.8%)` SOC total lands on the `Cardiac disorders` label row. A leaf that
repeats its parent (e.g. `soc = "Cardiac disorders"`, `pt = "Cardiac
disorders"`) is folded the same way. Control which markers count with
`group_summary`: `"empty"` (the `NA` / `""` leaf), `"parent"` (the
repeated-parent leaf), both (the default `c("empty", "parent")`), or `"none"`
to switch the folding off. Demographic tables -- where the leaf is a statistic
name such as `"n"` or `"Mean (SD)"`, never `NA` and never equal to the parent --
are unaffected either way, so their label rows keep their empty cells and the
stats stay on the indented leaf rows.

Because the output is an ordinary data.frame with baked-in indentation,
everything downstream works unchanged -- `group_by = "auto"` detects the
indent, the group-aware splits keep a label row with its children and append
`" (Cont.)"` when a group continues, and `blank_rows = "between_groups"`
separates the groups:

```{r stub-cols-pipeline}
pages <- as_rtftables(tbl, split = "group_force", max_rows = 20,
                      blank_rows = "between_groups")
```

With more than two `vars` each level indents one step further; a run of the
same parent value gets one label row, so **sort the input first** (or use
`sort_by`) if the hierarchy is scattered.

### Column `label` attributes as header labels

Clinical data.frames read via haven / labelled / xportr usually carry a
`label` attribute on each column. The data.frame branch of `as_rtftables()`
reads it (the `"labels"` `read_meta` token, on by default): a labelled column
uses its label as the header label, an unlabelled one keeps its name, and an
explicit `col_header =` still always wins.

```{r df-labels}
df <- data.frame(param = c("Age", "Sex"), val = c("75.1", "53%"),
                 stringsAsFactors = FALSE)
attr(df$param, "label") <- "Parameter"

as_rtftable(df)$col_header               # "Parameter", "val"
as_rtftable(df, read_meta = FALSE)$col_header  # NULL -> names(df) used
```

The labels feed the `header_sep` reconstruction too, so a label containing
`"____"` splits into a spanning header exactly like a delimited column name.
`stub_cols()` preserves the `label` attributes of the columns it keeps, and
uses the labels of the columns it merges for its default stub header.

### Coming from `reporter::create_table()`

If you know the reporter package's table spec, the finishing features map to
existing arguments -- there is no spec object to build:

| reporter | rtfreporter |
|----------|-------------|
| `stub(vars)` + `define(label_row=)` / `define(indent=)` | `stub_cols(vars=)` before converting |
| `use_attributes = "label"` | `read_meta` `"labels"` token (default on) |
| `define(visible = FALSE)` | `drop_cols` |
| `define(dedupe = TRUE)` | `collapse_repeats` |
| `define(blank_after = TRUE)` | `blank_rows = blank_rows_by_change(...)` |
| `define(page_break = TRUE)` | `split = "by_value", group_col =` |
| `define(align =, width =, label =)` | `col_spec` / `auto_width` / `col_header` |
| `spanning_header()` | `header_sep` delimited names, or `col_header` |
| `create_table(first_row_blank =)` | `blank_row_first` |
| `create_table(show_cols =)` | subset the data.frame, or `drop_cols` |
| sorting | `sort_by` / `sort_desc` |


## Pagination

Everything above is independent of how the table is split into pages. When the
body is longer than one page, control the split with `as_rtftables()`s
`split` / `max_rows` / `group_col` arguments (or a custom function), order the
rows first with `sort_by` / `sort_desc`, and hide a grouping / sort-key column
from the output with `drop_cols` -- all of that is covered in
[Paginating with as_rtftables()](pagination.html).

See `?as_rtftables` for the complete argument reference and the per-source
*what is carried* table.
