---
title: "Listings end to end: from source data to the written RTF"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
options(width = 100)
```

```{r setup}
library(rtfreporter)
```

A table and a listing want different things from the same package. A table is
already a table by the time it reaches `rtfreporter`: some framework counted,
summarised and labelled it, and what is left is rendering. A listing is
different -- it is one row of source data per subject, and the work is
**shape**:

* several source variables belong in one printed column, joined by `/`;
* a long cell has to break over several physical rows so it fits a column that
  is only so many characters wide;
* narrow blank columns sit between the printed ones as gutters;
* one blank row separates one subject's block from the next; and
* **a page break must never land inside a subject's block.**

`listing_col()`, `listing_spec()` and `build_listing()` do the first four.
The fifth is `as_rtftables()`, which could already do it. This article runs the
whole thing, source data to `.rtf`.

## The source data

A small ADSL, one row per subject, shaped like a baseline-characteristics
listing. Note the values that are longer than the column they will print in,
and the missing ones:

```{r data}
adsl <- data.frame(
  USUBJID  = c("63016-204-1015", "63016-204-1023", "63016-205-100028",
               "63016-206-1034", "63016-206-1045"),
  DISPTPD  = c("COMPLETED", "COMPLETED", "DISCONTINUED",
               "ONGOING", "COMPLETED"),
  BRCA     = c("BRCA1", NA, "BRCA2", "BRCA1", NA),
  HIST     = c("ADENOCARCINOMA", "SQUAMOUS CELL CARCINOMA OF THE LUNG",
               "SMALL CELL", "HIGH GRADE SEROUS CARCINOMA", "ADENOCARCINOMA"),
  STAGE    = c("IIIB", "IV", "IIIA", "IV", "IIB"),
  HISTGRD  = c("GRADE 3", "GRADE 2", "GRADE 3", "GRADE 1", NA),
  PRRAD    = c("Y", "N", "Y", "N", "Y"),
  CMBRFST  = c("PARTIAL RESPONSE", "STABLE DISEASE", "COMPLETE RESPONSE",
               "PROGRESSIVE DISEASE", "PARTIAL RESPONSE"),
  CMBRLST  = c("STABLE DISEASE", "PROGRESSIVE DISEASE", "STABLE DISEASE",
               NA, "STABLE DISEASE"),
  ECOGPS   = c("0", "1", "1", "0", "2"),
  stringsAsFactors = FALSE
)
str(adsl)
```

## Describing the columns

`listing_col()` describes **one printed column**: which source variables it is
built from, how wide it may be before its text wraps, and what its header says.

```{r listing-col}
listing_col(c("DISPTPD", "BRCA", "HIST"), width = 22,
            label = "Disposition/\nAny (BRCA) Mutations/\nHistology")
```

Four things are worth reading twice.

**`vars` may name several columns.** Their values are joined with the
listing's separator (`/` by default), and **missing and empty values are
skipped** -- a subject with no `BRCA` prints `COMPLETED/SQUAMOUS...`, not
`COMPLETED//SQUAMOUS...`.

That join is `catx()`, which the package exports for the columns you build
yourself -- so a listing needs nothing outside rtfreporter:

```{r catx}
catx("/", "COMPLETED", NA, "ADENOCARCINOMA")
```

**`width` is a number of characters, not a rendered width.** It decides where
the text breaks onto another physical row, and therefore how tall the subject's
block is. What the column *measures* in the table is `rel_width`, which
defaults to `width`.

**The output column takes the first variable's name** -- `DISPTPD` above --
which is why the reshaped data below has a `DISPTPD` column holding the joined
text. It is an internal name, not a header; pass `name = "COL01"` if you
would rather read the columns positionally.

**`label` may contain `"\n"`**, which starts another header row -- the same
convention as everywhere else in rtfreporter.

`listing_spec()` collects the columns, in the order they print, and the
settings that apply to all of them:

```{r listing-spec}
spec <- listing_spec(list(
  listing_col("USUBJID", width = 15, label = "Unique\nSubject ID"),
  listing_col(c("DISPTPD", "BRCA", "HIST"), width = 22,
              label = "Disposition/\nAny (BRCA) Mutations/\nHistology"),
  listing_col("STAGE", label = "Stage at\nInitial\nDiagnosis"),
  listing_col(c("HISTGRD", "PRRAD"), width = 18,
              label = "Histologic Grade/\nPrior Radiation\nTherapy"),
  listing_col(c("CMBRFST", "CMBRLST"), width = 20,
              label = "Best Response to\nthe 1st Line/\nthe Last Line"),
  listing_col("ECOGPS", label = "ECOG\nPerformance\nStatus")
))
spec
```

## What `build_listing()` does

```{r build}
body <- build_listing(adsl, spec)
body
```

Read that against the source data:

* **`USUBJID` wrapped.** `63016-205-100028` is 16 characters and the column
  takes 15, so it breaks after a hyphen -- `build_listing()` breaks at word
  boundaries (a space, a comma, a hyphen) and takes the break *after* the
  character, so the reader can see why the line ended.
* **The joined column breaks at the separator first.** Each source variable
  starts its own line, which is what makes a listing of this shape readable at
  all. Only a piece that is *still* too long breaks again at a word boundary --
  `SQUAMOUS CELL CARCINOMA OF THE LUNG` becomes two lines under a 22-character
  column.
* **Every column of one subject is padded to the tallest**, so the block stays
  aligned across columns, and **a blank row closes each block**.
* **`.sp1` … `.sp5` are the gutters**, the narrow blank columns between the
  printed ones.
* **`.rtf_record` is bookkeeping**, not a printed column: one id per source row.
  It is what keeps a subject whole across a page break, below.

A token that is *still* too wide -- one with nowhere to break -- is hard-split
at the column width. Cutting a subject id in half is ugly, but the alternative
is worse: the column's relative width follows `width`, so an overflowing token
is one Word wraps onto a second line, making the record a row taller than
`build_listing()` counted and the page taller than `max_rows` allowed for.
Every line this produces fits the column it was measured against.

Widths are **display widths**: a full-width (CJK) glyph occupies two monospaced
columns and counts as two, so a Japanese listing wrapped to 20 really does fit
in twenty.

## Rendering it

`build_listing()` is preparation only -- it hands back an ordinary
`data.frame`, and `as_rtftables()` renders. The body carries its spec, so the
two compose:

```{r render-composed}
pages <- as_rtftables(body)
pages[[1]]
```

The **header, the relative widths and the left alignment all came from the
spec.** None of the three had to be written out: no vector of header strings
with an empty entry for every gutter, no `col_rel_width = c(15, 1, 22, 1, ...)`
to keep in step with the column list, no `col_spec` repeating `align = "left"`
ten times.

Usually you do not want the intermediate object at all, and pass the spec
straight to `as_rtftables()`:

```{r render-hook}
pages <- as_rtftables(adsl, listing = spec)
length(pages)
```

The two forms produce the same tables. Reach for `build_listing()` when you
want to look at -- or patch -- the reshaped data before it is rendered; reach
for `listing =` the rest of the time.

## Keeping a subject whole across a page break

Give it a row budget and it paginates, **never splitting a subject's block**:

```{r paginate}
pages <- as_rtftables(adsl, listing = spec, max_rows = 8)
length(pages)
vapply(pages, function(p) nrow(p$data), integer(1))
pages[[2]]
```

Nothing new implements that. `build_listing()` emitted the `.rtf_record`
column, and the hook pointed the machinery that was already there at it:

```{r equivalent, eval = FALSE}
# what as_rtftables(listing = spec, max_rows = 8) sets for you
as_rtftables(body, max_rows = 8,
             group_col = ".rtf_record",   # a group is one subject's block
             group_by  = "value",         # a new value starts a new group
             split     = "group_safe",    # fill the page, never split a group
             drop_cols = ".rtf_record")   # ... and never print it
```

`drop_cols` columns are hidden **after** pagination, which is why a column can
decide the page breaks and still never appear. Every one of those is a
*default*: pass `split`, `group_col` or `drop_cols` yourself and the hook
leaves your choice alone.

## The whole report

Landscape, because a listing is wide; a title block, a footnote block, and the
file:

```{r report}
pages <- as_rtftables(adsl, listing = spec, max_rows = 8)

doc <- rtf_document(
  page = list(orientation = "landscape", paper_size = "A4",
              margin_top_in = 0.5, margin_bottom_in = 0.5,
              margin_left_in = 0.5, margin_right_in = 0.5),
  default_format = list(font_size_half_points = 16L)
) |>
  rtf_section(secinfo = list(
    header = rtf_header(list(
      c("Listing 16.2.4.2.1.2"),
      c("Baseline Characteristics"),
      c("<Safety Analysis Set>")
    )),
    footer = rtf_footer(list(
      c(l = "ECOG: Eastern Cooperative Oncology Group")
    ))
  )) |>
  rtf_tables(pages)

path <- file.path(tempdir(), "listing-16-2-4-2-1-2.rtf")
generate_rtfreport(doc, path, overwrite = TRUE)
file.exists(path)
```

That is the whole pipeline: `adsl` in, one `.rtf` out, with the column
description as the only thing written by hand.

## Listing types

`listing_spec(type = )` names a template that supplies the defaults --
separator, gutters and their width, the blank row after each record, the
default alignment, and the wrapping rule itself. One ships:

| `type` | what it means |
|---|---|
| `"multiline"` | `/` separator, gutter columns, a blank row after each record and one at the top of every page, left aligned, and text wrapped at the separator first and at word boundaries only where a piece is still too long. |

Any argument you pass explicitly overrides the template, the same relationship
`rtftable(border = "tfl")` has with its preset:

```{r overrides}
listing_spec(c("USUBJID", "STAGE"),
             sep       = " | ",   # join with something else
             spacer    = FALSE,   # no gutter columns
             blank_row = FALSE)   # no blank row between records
```

`record = FALSE` drops the hidden record column too -- do that only if you
intend to paginate some other way, because it is what a page break is told to
respect.

Because the wrapping rule belongs to the template rather than to
`build_listing()`, a future type can bring its own without disturbing this one.

## The hand-written version this replaces

Before this existed, the same listing was a page of pipeline: `catx()` calls to
join the columns, a `split_string()` that broke a cell at `/` and then at word
boundaries, a `get_max_element_counts()` to find each record's height, a
`pad_list_elements()` to pad every column to it, `unnest_longer()`, a manually
interleaved `S00`…`S08`, a blank row bound on top, and a
`split_df_by_blank_rows()` to cut pages at record boundaries -- plus the three
hand-maintained vectors (`col_rel_width`, the header strings, the `col_spec`)
that had to stay in step with all of it.

The package's test suite runs that pipeline -- its three functions verbatim --
beside `build_listing()` on the same data and compares the result cell for
cell, so this article's claim is checked rather than asserted. Four
differences are deliberate:

* **A token wider than the column is hard-split.** The hand-written rule
  pushed an empty accumulator first *and* let the token overflow -- so such a
  cell opened with a blank line, and Word then wrapped the token onto a row
  nobody had counted.
* **Widths are display widths.** The hand-written rule counted characters, so
  a Japanese cell asked for 8 columns and occupied 16.
* **A `"\n"` already in the data is honoured** as a line break, before any
  width is considered.
* **An empty cell still occupies its row.** The hand-written `split_string("")`
  returned no lines at all, so a record whose wrapped columns were all empty
  lost its blank row -- and with it the boundary a page split needs.

One more difference is a matter of placement rather than behaviour: the blank
row that used to be bound on top of the body is now
`as_rtftables(blank_row_first = )`, which the `"multiline"` template turns on,
so it appears at the top of **every** page rather than only the first.

## When not to use this

If your listing comes from **rlistings**, it is already laid out: `disp_cols`
has been applied, key-column repeat suppression is baked into the strings, and
the titles and footers travel with it. Pass it straight to `as_rtftables()`,
which reads all of that:

```{r rlistings, eval = FALSE}
lsting <- rlistings::as_listing(adsl, key_cols = "USUBJID")
as_rtftables(lsting, max_rows = 20)     # no `listing =` argument
```

`build_listing()` and `as_rtftables(listing = )` refuse a `listing_df` for that
reason, with a message pointing here. (An rlistings listing is a `data.frame`
subclass, so without that guard it would quietly be treated as raw source
data.)

## See also

* `?listing_col`, `?listing_spec`, `?build_listing` -- every setting, one page
  each.
* [Paginating tables with `as_rtftables()`](pagination.html) -- the split
  strategies, `group_col` / `group_by`, and `drop_cols` in general.
* [Importing tables](importing-tables.html) -- reading a table object,
  rlistings included, when the layout is already done.
