---
title: "A tfrmt table with several stub columns: one for sections, the rest as a stub"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
.has <- requireNamespace("tfrmt", quietly = TRUE) &&
        requireNamespace("gt",    quietly = TRUE)
```

```{r setup}
library(rtfreporter)
```

When a [tfrmt](https://gsk-biostatistics.github.io/tfrmt/) table places its
row-group levels in **columns** --
`row_grp_plan(label_loc = element_row_grp_loc(location = "column"))` -- each
grouping variable becomes its own **stub column** in the resulting `gt`. A very
common clinical layout then wants to treat those stub columns in **two different
ways at once**:

- **one** stub column drives **page / section splitting** (one section per lab
  test, per cohort, ...), its value dropped into the running header via
  `rtf_tables(auto_section = TRUE)`;
- the **remaining** stub columns are **merged into a single indented row
  heading** (the clinical "stub").

This article shows both, on a lab toxicity-grade table whose three stub columns
are **lab test code** (LBTESTCD -> sections), a **fixed heading**, and
**Toxicity Grade** (the last two -> a two-level indented stub).

## The data

A tidy toxicity-grade summary: for each lab test and grade, a percentage per
arm. `heading` is a constant column -- it becomes the fixed top level of the
stub.

```{r data, eval = .has}
set.seed(1)
tests  <- c("ALT", "AST")
grades <- c("Grade 0", "Grade 1", "Grade 2", "Grade 3")
arms   <- c("Placebo", "Active")

dat <- expand.grid(LBTESTCD = tests, Toxgrade = grades, ARM = arms,
                   stringsAsFactors = FALSE)
dat$heading <- "Toxicity Grade"
dat$param   <- "pct"
dat$value   <- round(runif(nrow(dat), 1, 40), 1)
dat <- dat[c("LBTESTCD", "heading", "Toxgrade", "ARM", "param", "value")]
head(dat)
```

## Build the tfrmt table (three stub columns)

`group = vars(LBTESTCD, heading)` plus `label = Toxgrade`, rendered with
`location = "column"`, makes **all three** row variables stub-type columns.

```{r build, eval = .has}
library(tfrmt)

tf <- tfrmt(
  group  = vars(LBTESTCD, heading),
  label  = Toxgrade,
  column = ARM,
  param  = param,
  value  = value,
  body_plan = body_plan(
    frmt_structure(group_val = ".default", label_val = ".default", frmt("xx.x"))
  ),
  row_grp_plan = row_grp_plan(
    label_loc = element_row_grp_loc(location = "column"))
)

g <- print_to_gt(tf, dat)
```

The `gt`'s boxhead confirms the three stub columns (plus the two arm columns and
one hidden helper tfrmt adds):

```{r boxhead, eval = .has}
bx <- g[["_boxhead"]]
data.frame(var = as.character(bx$var), type = as.character(bx$type))
```

`gt::extract_body()` cannot read a table with more than one stub column, so
`as_rtftables()` reads such a table from the `gt` object's own slots. There the
**first** stub column (LBTESTCD) is renamed to `"rowname"`; the other stub
columns keep their names (`heading`, `Toxgrade`). That is the name you reference
below (an **integer index** -- `group_col = 1` -- is equally portable).

## One call: split by test, merge the rest into a stub

```{r pages, eval = .has}
pages <- as_rtftables(
  g, read_meta = TRUE,
  # 1) sections: split by the LBTESTCD stub column (now called "rowname"),
  #    then hide it from the printed table (it names the section instead).
  split      = "by_value",
  group_col  = "rowname",
  drop_cols  = "rowname",
  # 2) stub: fold the remaining two stub columns into one indented heading.
  stub_vars  = c("heading", "Toxgrade"),
  stub_label = "Toxicity Grade",
  border     = "tfl"
)

names(pages)          # one page per lab test -> section names
```

Each page carries the fixed `"Toxicity Grade"` heading flush-left, with the
grades indented beneath it -- the two stub columns merged into one:

```{r show, eval = .has}
show <- function(p) {                       # NBSP indent -> dots for display
  d <- p$data; d[[1]] <- gsub(intToUtf8(160L), ".", d[[1]], fixed = TRUE); d
}
show(pages[["ALT"]])
```

Hand the named list to `rtf_tables(auto_section = TRUE)` -- each name becomes a
section that starts on its own page with the test code in the running header:

```{r render, eval = FALSE}
doc <- rtf_document() |>
  rtf_section(secinfo = list(header = my_header, footer = my_footer)) |>
  rtf_tables(pages, auto_section = TRUE)

generate_rtfreport(doc, "lab-tox.rtf", overwrite = TRUE)
```

## Why the order matters

For `split = "by_value"`, `as_rtftables()` **splits the body first and builds
the stub per page**. That order is essential here because the `heading` level is
**constant** across the whole table: if the stub were built first (on the full
body), that constant level would collapse into a *single* heading row spanning
both tests, which could not then be divided into per-test sections. Splitting
first gives each section its own copy of the heading. (For every other split the
table stays one logical table paginated across pages, so the stub is built once,
up front -- see `vignette("pagination")`.)

Consequently, under `split = "by_value"` the `group_col` names a **pre-stub**
column (here `"rowname"` = LBTESTCD) and is **not** one of the `stub_vars`: the
outer level splits, the inner levels merge.

## See also

- `vignette("section-splitting")` -- getting section names where
  `auto_section` can see them, for every table object.
- `vignette("importing-tables")` -- what `as_rtftables()` reads from a `gt`
  (and why a multi-stub table is read from slots).
- [`stub_cols()`] -- the same stub merge on a plain `data.frame`, and the
  `stub_vars` argument of [`as_rtftables()`] that folds it into the pipeline.
```

<!-- where-next -->

## Where next

- [Pagination](pagination.html) --- how the rows are cut into pages

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
