---
title: "Rendering, post-processing and assembly"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

Once a document is built, three steps turn it into a deliverable: **render** it
to an `.rtf`, optionally **post-process** the file, and -- for a multi-table
package -- **assemble** several RTFs into one with a table of contents.

```{r doc}
df <- data.frame(
  Parameter = c("Creatinine clearance", "  Mean", "  SD"),
  Result    = c("", "88 ug/mL", "12 ug/mL"),
  stringsAsFactors = FALSE
)
doc <- rtf_document() |>
  rtf_tables(as_rtftables(df), titles = list(c("Table 14.3.1 Renal Function")))
```


## Rendering -- `generate_rtfreport()`

`generate_rtfreport()` writes the document to a file you can open in Word or
LibreOffice. Use `overwrite = TRUE` to replace an existing file.

```{r render}
out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```


## Post-processing -- `rtf_replace_text()`

Sometimes the rendered file needs a small tweak that you do not want to push
back into the (validated) source data. `rtf_replace_text()` does a find/replace
directly on the generated `.rtf`.

A typical case is **units**: the analysis dataset carries an ASCII-safe token
like `ug/mL`, but the report should display `µg/mL`. Rather than putting a
non-ASCII character in the data, render with the ASCII token and swap it in the
finished file. The replacement is the RTF unicode escape for `µ`
(U+00B5 = 181), i.e. `\u181?`:

```{r postprocess}
final <- tempfile(fileext = ".rtf")
rtf_replace_text(out, "ug/mL", "\\u181?g/mL", output_file = final)

rtf <- paste(readLines(final, warn = FALSE), collapse = "\n")
grepl("\\u181?g/mL", rtf, fixed = TRUE)   # Word/LibreOffice show "µg/mL"
```

The same approach swaps any ASCII placeholder for a symbol you cannot (or
prefer not to) carry in the source data. `rtf_replace_text()` also supports vectorised
targets, fixed or regex matching (`use_regex = TRUE`), case-insensitive
matching, in-place editing (with an automatic `.bak`) or writing to a separate
`output_file`. See `?rtf_replace_text`.

> Note: rtfreporter already escapes non-ASCII characters you put in the data
> (so `"µg/mL"` in a cell renders correctly too). Post-processing is the tool
> for the "last mile" -- when the source must stay ASCII, or for a one-off fix
> to a finished file.


## Assembling several RTFs -- `assemble_rtf()`

A clinical deliverable is usually many tables, listings and figures. Render each
to its own `.rtf`, then combine them into a single document with a table of
contents using `assemble_rtf()`.

```{r make-files}
make <- function(title, path) {
  d <- data.frame(A = c("1", "2"), B = c("x", "y"), stringsAsFactors = FALSE)
  doc <- rtf_document() |> rtf_tables(as_rtftables(d), titles = list(c(title)))
  generate_rtfreport(doc, path, overwrite = TRUE); path
}
f1 <- make("Table 14.1.1 Demographics",   tempfile(fileext = ".rtf"))
f2 <- make("Table 14.2.1 Adverse Events",  tempfile(fileext = ".rtf"))
```

### Automatic TOC

`toc = "auto"` builds one TOC entry per file, taking each label from the file's
first title:

```{r assemble-auto}
pkg <- tempfile(fileext = ".rtf")
assemble_rtf(c(f1, f2), pkg, toc = "auto", overwrite = TRUE)
file.exists(pkg)
```

### Manual, multi-level TOC

For sections and explicit labels, pass a list of `toc_heading()` (a section
heading) and `toc_entry()` (a linked entry). Each entry creates an RTF
**bookmark**, so the TOC is navigable in the word processor and the entries jump
to the right table:

```{r assemble-manual}
pkg2 <- tempfile(fileext = ".rtf")
assemble_rtf(
  c(f1, f2), pkg2,
  cover = list(title = "Study XYZ-001", subtitle = "Final Tables", version = "v1.0"),
  toc = list(
    toc_heading("EFFICACY ANALYSES"),
    toc_entry("Table 14.1.1 Demographics",  file = f1),
    toc_heading("SAFETY ANALYSES"),
    toc_entry("Table 14.2.1 Adverse Events", file = f2)
  ),
  toc_page_numbering = "roman",
  overwrite = TRUE
)
file.exists(pkg2)
```

`assemble_rtf()` also takes a `cover` page and `toc_leader` / `toc_page_numbering`
options; see `?assemble_rtf`.

### Assembling a folder

To combine every RTF in a directory, `assemble_files()` lists them (by pattern)
and `assemble_folder()` renders a spec-driven package; a spec file
(`assemble_spec()` / `assemble_from_spec()`) lets you script the order, headings
and labels. See `?assemble_files` and `?assemble_folder`.

> Note: the assembled file is RTF. The TOC entries are RTF bookmarks for
> word-processor navigation; converting to PDF (and confirming the PDF
> bookmarks) is outside the scope of this article.

That completes the basic workflow: build a document, render it, post-process if
needed, and assemble a package. See `?generate_rtfreport`, `?rtf_replace_text`,
and `?assemble_rtf` for full references.

<!-- where-next -->

## Where next

- [The pharmaverse catalog](tlg-catalog.html) --- complete worked deliverables

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
