---
title: "Page and document setup"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

Every report starts with `rtf_document()`, which carries the **document-wide**
settings: page geometry, fonts, colours, and the default font size. The
defaults are tuned for clinical TFL output, so you can usually call it with no
arguments -- but each piece is configurable. This article walks through what you
can set at the document level and how it cascades to the tables.

```{r quick}
df <- data.frame(
  Parameter = c("Age (years)", "  Mean", "  SD"),
  Value     = c("", "75.1", "8.2"),
  stringsAsFactors = FALSE
)

# All defaults -- landscape Letter, 9pt Courier
doc <- rtf_document() |>
  rtf_tables(as_rtftables(df))
```


## Setting the document config: `rtf_document()` (and `rtf_config()`)

A document has exactly **one** document-level configuration, and you can set all
of it on `rtf_document()`. The two calls below are equivalent:

```{r equivalent, eval = FALSE}
rtf_document(default_format = list(font_size_half_points = 20L))

rtf_document() |>
  rtf_config(default_format = list(font_size_half_points = 20L))
```

For a brand-new document `rtf_config()` is **not required** -- you can pass the
same settings straight to `rtf_document()`. Where it earns its keep is **after
you have composed the report**: build the content and sections once, then derive
output variants by changing only a document-level setting. Because an
`rtf_document` is immutable, each variant is an independent copy that shares the
same content.

```{r variants}
# Compose the report once (content + section).
base <- rtf_document() |>
  rtf_tables(as_rtftables(df)) |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL))

# Then render the SAME report in two paper sizes by changing only `page`.
letter <- base                                                  # landscape Letter
a4      <- base |> rtf_config(page = list(paper_size = "A4"))   # landscape A4

# generate_rtfreport(letter, "report-letter.rtf", overwrite = TRUE)
# generate_rtfreport(a4,     "report-a4.rtf",     overwrite = TRUE)
```

`page` and `default_format` are **merged per key**, so the A4 variant above
keeps `base`'s orientation and margins and only swaps the paper size. The same
trick produces a large-print copy
(`rtf_config(default_format = list(font_size_half_points = 24L))`) or a
client-specific font (`rtf_config(font_table = list(list(name = "Arial")))`)
from one composed report.

Both functions take the same four parameters: `page`, `font_table`,
`color_table`, and `default_format`, covered below.


## Page geometry -- `page`

Pass `page = list(...)`. Dimensions and margins are given in **inches**; any key
you omit falls back to its default.

| Key | Meaning | Default |
|-----|---------|---------|
| `paper_size` | Named preset: `"letter"`, `"legal"`, `"A4"`, `"A3"`, `"A5"` (case-insensitive) | `"letter"` |
| `orientation` | `"landscape"` or `"portrait"` | `"landscape"` |
| `width_in` | Paper width (inches) -- explicit override | from `paper_size` |
| `height_in` | Paper height (inches) -- explicit override | from `paper_size` |
| `margin_top_in` | Top margin | `0.75` |
| `margin_bottom_in` | Bottom margin | `0.75` |
| `margin_left_in` | Left margin | `0.75` |
| `margin_right_in` | Right margin | `0.75` |
| `header_dist_in` | Header band distance from page edge | `margin_top_in` |
| `footer_dist_in` | Footer band distance from page edge | `margin_bottom_in` |

The default margins follow the FDA eCTD *Portable Document Format (PDF)
Specifications* minimums (left/binding &ge; 0.75", other sides &ge; 0.375",
landscape top &ge; 0.75"): a uniform **0.75" on all four sides** (the bottom
leaves room for an overall page-number footer).

The header/footer band distances (`\headery` / `\footery` in the RTF) default to
the **full top / bottom margin**, so the band sits at the margin boundary. Set
`header_dist_in` / `footer_dist_in` to pin them explicitly.

### Changing the margins

Pass any of the four `margin_*_in` keys to `rtf_document(page = )` (or
`rtf_page()`) to override a margin for one document; an omitted key keeps its
default. **A wide table (many treatment arms, toxicity-grade shift tables) often
does not fit inside the 0.75" left/right default.** Narrowing the left/right
margins to eCTD's **0.5"** buys an extra inch of horizontal room -- they still
clear the 0.375" floor:

```{r narrow-lr, eval = FALSE}
# Landscape Letter, left/right narrowed to 0.5" for a wide TFL.
# Writable width = 11 - 2*0.5 = 10 in = 14400 twips (vs 9.5 in / 13680 at 0.75").
rtf_document(page = list(margin_left_in = 0.5, margin_right_in = 0.5))
```

To apply the same margins to **every** report in a session or site (e.g. in
`Rprofile.site`), set the matching options instead of repeating them per
document:

```{r narrow-lr-option, eval = FALSE}
options(rtfreporter.page.margin_left_in  = 0.5,
        rtfreporter.page.margin_right_in = 0.5)
```

`auto_width` (and `as_rtftables()`) size columns to the *default* writable width,
so they follow the current margins automatically; only **absolute**
`column_widths_twips` need to be recomputed when you change the margins.

**Pick a paper size by name.** For the common case, `paper_size` + `orientation`
is easier than remembering inch dimensions. One line gives **A4 landscape**:

```{r a4-preset, eval = FALSE}
rtf_document(page = list(paper_size = "A4"))                          # A4 landscape
rtf_document(page = list(paper_size = "A4", orientation = "portrait")) # A4 portrait
```

**Explicit dimensions win and are used as given.** If you pass `width_in` /
`height_in`, those exact dimensions are used and the orientation is *inferred*
from them (`width_in >= height_in` means landscape). An `orientation` that
contradicts the dimensions emits a warning and the dimensions are kept (not
swapped); a `paper_size` given alongside explicit dimensions is ignored (with a
warning). So this is A4 **portrait** (taller than wide), and the
`orientation = "landscape"` is a contradiction that warns:

```{r a4-explicit, eval = FALSE}
rtf_document(page = list(width_in = 8.27, height_in = 11.69))   # inferred portrait
rtf_document(page = list(orientation = "landscape",
                         width_in = 8.27, height_in = 11.69))   # warns; stays portrait
```

The default is **landscape US Letter** (11 × 8.5 in), the usual clinical TFL
page. For a portrait A4 page with wider margins:

```{r portrait}
doc <- rtf_document(
  page = list(
    orientation    = "portrait",
    width_in       = 8.27, height_in = 11.69,   # A4
    margin_left_in = 0.75, margin_right_in = 0.75,
    margin_top_in  = 1.0,  margin_bottom_in = 1.0
  )
) |>
  rtf_tables(as_rtftables(df))
```

The writable width (paper width minus the left/right margins) is what tables are
sized against, so changing the margins automatically reflows full-width tables.


## Fonts -- `font_table`

The report uses a **single** font, given by `font_table`. The default is
**Courier**, a monospaced font: in clinical listings, monospacing is what lets
count/percent columns line up by character (see the count/percent formatters in
the pagination article). Switch to a proportional font only if you do not rely
on character alignment:

```{r font}
doc <- rtf_document(font_table = list(list(name = "Arial"))) |>
  rtf_tables(as_rtftables(df))
```

`font_table` is written as a list for forward compatibility, but **only the
first entry is currently used** -- any further entries are ignored, and there is
no way to select a different font for specific cells or text. (Multiple fonts
and per-text font selection are a planned enhancement.) To use one font
throughout, set that first entry as above.


## Document font size -- `default_format`

The document font size lives in `default_format` and is expressed in
**half-points** (the unit of the RTF `\fs` command): `18` = 9 pt, `20` = 10 pt,
`24` = 12 pt, and so on. 9 pt is the package default.

```{r fontsize}
doc <- rtf_document(default_format = list(font_size_half_points = 22L)) |>  # 11pt
  rtf_tables(as_rtftables(df))
```

The font size also **drives the default row height**: taller text needs taller
rows, so rtfreporter picks a sensible default row height for the chosen size.

```{r rowheight-default}
# Internal lookup, shown here to illustrate the cascade:
rtfreporter:::.default_row_height_twips(18L)   # 9pt  -> 230 twips
rtfreporter:::.default_row_height_twips(22L)   # 11pt -> 270 twips
```


## Default cell / row height

All table-shaped elements -- page header/footer rows, column-header rows, data
rows, blank separator rows, and footnotes -- share a **font-size-aware default
row height** (in twips; 1 twip = 1/1440 inch = 1/20 point). These defaults live
in `inst/resources/rtfreporter_defaults.R` (admin-tunable, package-wide):

| Font size | Default row height |
|-----------|--------------------|
| 8 pt (`16`) | 210 twips |
| 9 pt (`18`) | 230 twips *(default)* |
| 10 pt (`20`) | 250 twips |
| 11 pt (`22`) | 270 twips |
| 12 pt (`24`) | 290 twips |

You rarely need to touch these. To change heights for a specific table, override
them per table on `rtftable()` / `as_rtftables()`:

```{r rowheight-override}
doc <- rtf_document() |>
  rtf_tables(as_rtftables(
    df,
    row_height_twips        = 280L,   # data rows
    header_row_height_twips = 300L,   # column-header rows
    blank_row_height_twips  = 140L    # blank separator rows (e.g. half height)
  ))
```

Cell padding (the gap between the cell border and its text) defaults to `0`
(text sits flush against the border, the typical clinical look) and is likewise
configurable per table via `cell_padding_left_twips` / `cell_padding_right_twips`.

### Document-wide row height / padding

To change the row height or cell padding for the **whole report** -- without
editing the resource file -- set it on the document via `default_format` (or
globally as an option). It applies to every table-shaped element (content table,
page header/footer, title/footnote), and a per-module value still overrides it:

```{r rowheight-doc}
doc <- rtf_document(default_format = list(
  row_height_twips         = 280L,
  cell_padding_left_twips  = 72L,
  cell_padding_right_twips = 72L
))

# ...or set the house style site-wide (e.g. in Rprofile.site):
# options(rtfreporter.row_height_twips = 280L,
#         rtfreporter.cell_padding_left_twips = 72L)
```

Resolution is highest-wins: **per-module** (`rtftable()` / `rtf_header()` /
`rtf_footer()` / `rtf_table_style()`) > **document** `default_format` >
`rtfreporter.*` **option** > the font-aware / resource baseline.


## Site-wide defaults

The document-level defaults above (paper size, orientation, margins, font, font
size) are read from `rtfreporter.*` **options**, resolved as: an explicit
argument you pass wins, otherwise the option value, otherwise the built-in
factory baseline. So a site or organisation can set its house style once -- for
example in `Rprofile.site` -- and every report inherits it without changing any
script:

```{r site-defaults, eval = FALSE}
options(
  rtfreporter.font              = "Arial",
  rtfreporter.page.paper_size   = "A4",
  rtfreporter.page.margin_top_in = 1.0
)
```

Inspect the values a report will use, and restore the factory baseline, with:

```{r defaults-helpers, eval = FALSE}
rtfreporter_options()        # snapshot of the resolved defaults (audit trail)
rtfreporter_reset_defaults() # discard any overrides, back to the factory values
```

**Reproducibility note.** Because options can change the output of identical
code, a validated or reproducible run should pin the configuration *in the
script* (pass the settings explicitly, or call `rtfreporter_reset_defaults()`
first) rather than relying on the environment, and can record
`rtfreporter_options()` alongside the output as an audit trail.


## Colours

Colours are given as `"#RRGGBB"` hex strings, and rtfreporter collects every
colour you reference and builds the RTF colour table automatically -- you never
maintain a palette. Two things can be coloured, which together cover the usual
clinical TFL needs.

**Borders** -- `rtf_border(color = ...)` / `rtf_border_side(color = ...)`,
anywhere a border is accepted (column-header rules, the body / first-row /
last-row zones, spanning headers, and the page header/footer band border). For
example, a navy rule under a header:

```{r colors-border}
navy_rule <- rtf_border(bottom = rtf_border_side(color = "#1F4E79"))

doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(c(l = "Protocol XYZ-001")), border = navy_rule),
    footer = NULL
  )) |>
  rtf_tables(as_rtftables(df))
```

**Body (data-cell) text** -- per column via `col_spec`'s `color`, or per cell
via `cell_styles`'s `color` (which overrides the column colour). Useful, for
example, to flag an out-of-range value in red:

```{r colors-text}
doc <- rtf_document() |>
  rtf_tables(as_rtftables(
    df,
    col_spec = list(list(col = 2, color = "#C00000"))   # column 2 text in red
  ))
```

**What is not coloured (by design, for now):** cell background / fill, and the
text of column headers, the page header/footer, titles, and footnotes -- these
render in the default black. The `color_table` argument on `rtf_document()` only
pre-declares colours in the table; it does not style anything by itself.


## A complete example

Putting it together -- a portrait A4 page, Arial at 11 pt, custom margins, and a
navy header rule:

```{r full, eval = FALSE}
navy_rule <- rtf_border(bottom = rtf_border_side(color = "#1F4E79"))

doc <- rtf_document(
  page = list(
    orientation    = "portrait",
    width_in       = 8.27, height_in = 11.69,
    margin_left_in = 0.75, margin_right_in = 0.75,
    margin_top_in  = 1.0,  margin_bottom_in = 1.0
  ),
  font_table     = list(list(name = "Arial")),
  default_format = list(font_size_half_points = 22L)
) |>
  rtf_section(page = 1, secinfo = list(
    header = rtf_header(rows = list(c(l = "Protocol XYZ-001", r = "Confidential")),
                        border = navy_rule),
    footer = rtf_footer(c(c = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"))
  )) |>
  rtf_tables(as_rtftables(df, row_height_twips = 280L))

generate_rtfreport(doc, "report.rtf", overwrite = TRUE)
```

See `?rtf_document` and `?rtf_config` for the full argument reference.

<!-- where-next -->

## Where next

- [Headers and footers](headers-footers.html) --- what sits in the margins you just set

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
