---
title: "Paginating tables with as_rtftables()"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

`as_rtftables()` turns a table into a **list of `rtftable` page objects** --
one per RTF page -- ready to hand to `rtf_tables()`. Pagination has three
pillars, and this article is organised around them:

1. **How the rows are split into pages** -- the `split =` strategies (and custom
   / pre-split / hybrid variants).
2. **Which column those page-shaping operations act on, and what to do with it**
   -- one shared idea behind `group_col`, `group_by`, `sort_by`,
   `collapse_repeats` and `drop_cols` (see
   [Selecting a column](#selecting-a-column)).
3. **Blank separator rows** -- inserting and counting the gaps between groups.

We start with the split strategies:

1. the **built-in split strategies**,
2. a **custom split function**,
3. **pre-splitting** the data yourself, and
4. an **advanced** hybrid that composes a built-in strategy.

Throughout we use a small demographics-style table. By the clinical-TFL
convention, a row whose first column starts with a non-space character opens a
*group*, and the indented rows below it are that group's members -- which is how
`as_rtftables()` detects groups by default (`group_col = NULL`, `group_by =
"auto"`). Tables that mark groups differently -- a repeated key column, or a
group label with blank member rows -- are handled by `group_col` / `group_by`,
covered under [Shared knobs](#shared-knobs) below.

```{r data}
df <- data.frame(
  Parameter = c("Age (years)", "  Mean", "  SD", "  Min", "  Max",
                "Sex",         "  F", "  M",
                "Race",        "  White", "  Black", "  Asian", "  Other"),
  Value     = c("",     "75.1", "8.2", "50", "89",
                "",     "179 (58%)", "127 (42%)",
                "",     "273 (89%)", "29 (9%)", "2 (1%)", "2 (1%)"),
  stringsAsFactors = FALSE
)
```

A tiny helper to *show* how a result was paginated -- it prints the row-label
column of each returned page:

```{r helper}
show_pages <- function(pages) {
  for (i in seq_along(pages)) {
    nm <- names(pages)[i]
    tag <- if (!is.null(nm) && nzchar(nm)) sprintf(" (%s)", nm) else ""
    cat(sprintf("--- Page %d%s ---\n", i, tag))
    cat(paste0("  ", pages[[i]]$data[[1L]]), sep = "\n")
    cat("\n")
  }
  invisible(pages)
}
```


## 1. Built-in split strategies

Pass `split =` one of the built-in strategy names.

### `"none"` (default) -- one page

The whole table stays on a single page; no row limit is checked.

```{r none}
pages <- as_rtftables(df)            # split = "none"
length(pages)
```

### `"rows"` -- cut at explicit positions

Give `split_rows` the row numbers at which a new page should start. Use this
when you know exactly where the breaks go.

```{r rows}
pages <- as_rtftables(df, split = "rows", split_rows = c(6, 9))
show_pages(pages)
```

### `"group_force"` -- fill `max_rows`, repeat the group label on a cut

Pack up to `max_rows` rows on each page. When a page boundary falls *inside* a
group, a continuation row is inserted at the top of the next page repeating the
group label with a `" (Cont.)"` suffix, so the reader knows the group carries
over.

```{r group_force}
pages <- as_rtftables(df, split = "group_force", max_rows = 6)
show_pages(pages)
```

### `"group_safe"` -- never break a group across pages

Pack *whole* groups onto a page; a group that would overflow is pushed to the
next page instead of being split. (A single group larger than `max_rows` is
still force-split as a fallback.)

```{r group_safe}
pages <- as_rtftables(df, split = "group_safe", max_rows = 6)
show_pages(pages)
```

### `"by_value"` -- one page per group, named by the group

Emit exactly one page per detected group; the returned list is **named** by the
group label (which `rtf_tables()` can use, e.g. with `auto_section = TRUE`).

```{r by_value}
pages <- as_rtftables(df, split = "by_value")
names(pages)
```

### Reusing a strategy

Earlier versions also exported a `page_split_*()` factory per strategy, so a
configured splitter could be built once and passed around. They were retired
before the first release: a factory carried its own `group_col`, which meant
the group column could be declared in two places at once, and the two did not
always agree. Every setting now has exactly one place to live -- alongside
`split` -- so reuse is a list of arguments rather than an object:

```{r reuse}
args  <- list(split = "group_safe", max_rows = 6)
pages <- do.call(as_rtftables, c(list(df), args))
length(pages)
```

### Shared knobs

These arguments shape pagination regardless of the strategy:

| Argument | Effect |
|----------|--------|
| `max_rows` | Maximum data rows per page (required for the group strategies). |
| `group_col` | *Which* column groups are detected on: a name, a 1-based index, or `NULL` for column 1. |
| `group_by` | *How* a group boundary is found on `group_col`: `"auto"`, `"indent"`, `"value"`, or `"filled"` (see below). |
| `min_group_rows` | Widow/orphan control: avoid stranding a lone group header (or too few of its rows) at a page boundary. Set `0` to disable. |
| `cont_label` | The continuation suffix (`" (Cont.)"`) repeated on a mid-group cut. |

### Selecting a column {#selecting-a-column}

Five arguments operate on **one column of the rendered body**, and they all name
that column the same way -- by a column **name** or a 1-based **integer index**,
in the body's coordinate space:

| Argument | What it does to the column |
|----------|----------------------------|
| `group_col` | detects group boundaries on it (with `group_by`) |
| `sort_by` | orders the rows by it, before the split |
| `collapse_repeats` | blanks consecutive repeated values in it |
| `drop_cols` | hides it from the printed pages |

Because they share one coordinate space, the same key works across them -- e.g.
sort on a column, group on it, and hide it:
`as_rtftables(x, sort_by = "k", group_col = "k", drop_cols = "k")`.

Two things make the *name* source-dependent, so they are worth repeating here:

- **It is the body column id, not the visible header.** A gtsummary `stat_1`
  column shows as "Drug (N = …)" but is selected as `"stat_1"`.
- **The names differ by source** -- your own for a `data.frame`, gt ids for
  gt / gtsummary / tfrmt, and `V1`, `V2`, … for rtables / flextable / huxtable.
  An **integer index is the most portable** choice across sources. Check the
  actual names with `names(as_rtftables(x)[[1]]$data)`. The full per-source table
  and check recipes are in
  [What the columns are called](importing-tables.html#column-names).

### Choosing how groups are detected -- `group_by`

`group_col` picks the *column*; `group_by` picks the *logic* used to find where
one group ends and the next begins on that column:

| `group_by` | A row starts a new group when its `group_col` cell ... |
|------------|--------------------------------------------------------|
| `"indent"` | is non-empty and does **not** start with whitespace (space / tab / non-breaking space); indented or empty cells are members. This is the usual clinical row-label layout. |
| `"value"`  | differs from the row above it (each maximal run of equal values is one group). |
| `"filled"` | is non-empty; `NA` / `""` cells belong to the group above (the label appears once, on the group's first row). |
| `"auto"`   | (default) picks one from the column's content: indentation present → `"indent"`; else empty cells interspersed → `"filled"`; else → `"value"`. |

```{r group-by-value}
# Group by a repeated key column (value), regardless of indentation.
df_key <- data.frame(
  Visit = c("Week 1", "Week 1", "Week 2", "Week 2", "Week 4"),
  Result = c("12", "0.3", "14", "0.4", "11")
)
pages <- as_rtftables(df_key, split = "group_safe", max_rows = 2,
                      group_col = "Visit", group_by = "value")
length(pages)                                  # one page per Visit block
```

```{r group-by-filled}
# A group label on its first row, blank members below it (filled).
df_filled <- data.frame(
  Group = c("Group A", "", "", "Group B", ""),
  Value = c("", "1.2", "3.4", "", "5.6")
)
pages <- as_rtftables(df_filled, split = "group_safe", max_rows = 3,
                      group_by = "filled")
length(pages)
```

The grouping column need not be the first column -- name it (or, for portability
across sources, give its integer index; see
[Selecting a column](#selecting-a-column)):

```{r group-by-col}
df2 <- data.frame(
  Parameter = c("ALT", "AST", "ALT", "AST"),
  Visit     = c("Week 1", "Week 1", "Week 2", "Week 2"),
  Value     = c("12", "20", "14", "22")
)
pages <- as_rtftables(df2, split = "group_safe", max_rows = 2,
                      group_col = 2L, group_by = "value")   # group on Visit
length(pages)
```

### Ordering rows before the split -- `sort_by` / `sort_desc`

`group_col` chooses *where* the breaks fall, but it does not reorder anything --
the body is split in the order it arrives. `sort_by` orders the body rows
**before** pagination, so group detection, the `" (Cont.)"` labels and any
`blank_rows` positions all see the sorted order. Name one or more sort keys (in
the input body's coordinates, the same space as `group_col` and `drop_cols`),
with optional per-key direction via `sort_desc`.

```{r sort-basic}
df_s <- data.frame(
  Region = c("West", "East", "West", "East"),
  Site   = c("S2",   "S1",   "S1",   "S2"),
  N      = c("12",   "9",    "20",   "4")
)
# Region ascending, then Site ascending.
as_rtftables(df_s, sort_by = c("Region", "Site"))[[1]]$data[c("Region", "Site")]
```

`sort_desc` is a single `TRUE` (all keys descending) or one logical per key:

```{r sort-desc}
# Region ascending, Site descending.
as_rtftables(df_s, sort_by = c("Region", "Site"),
             sort_desc = c(FALSE, TRUE))[[1]]$data$Site
#> "S2" "S1" "S2" "S1"
```

The sort is **stable** (rows that compare equal keep their input order) and `NA`
keys sort **last**, regardless of direction. It is meant for flat / tabular
bodies -- a `data.frame`, or a pre-flattened table; a gt / rtables body whose
group-label and child rows are already interleaved should be sorted at the
*source* (before the table is built), since reordering its rendered rows would
scramble that hierarchy. To mix name and index keys, pass a `list()`.

A sort key is often a column you do **not** want printed -- a numeric ordering
key, a raw priority. Combine `sort_by` with [`drop_cols`](#hiding-a-grouping-carrier-column----drop_cols)
to order on it and then hide it:

```{r sort-carrier}
df_c <- data.frame(
  soc_ord = c(2, 1, 2, 1),                  # hidden ordering key
  Term    = c("Rash", "Angina", "Pruritus", "Failure"),
  N       = c("9", "12", "4", "7")
)
res <- as_rtftables(df_c, sort_by = "soc_ord", drop_cols = "soc_ord")[[1]]$data
names(res)            # the key is gone
res$Term              # but the rows are ordered by it
#> [1] "Angina" "Failure" "Rash" "Pruritus"
```

### Blanking repeated values -- `collapse_repeats`

A `"value"`-style column carries the group key on *every* row, but tables
usually print the key only once per run and leave the repeats blank -- the
`"filled"` layout. `collapse_repeats` performs exactly that conversion: name one
or more columns and each run of equal values is suppressed to a single visible
cell, the rest replaced with `NA` (which renders as an empty cell -- no row is
removed, only the text is hidden).

```{r collapse-single}
df_rep <- data.frame(
  Group = c("All", "All", "BORC", "BORC", "BORC", "BORC",
            "HOGE", "HAGE", "HOGE", "HAGE", "HAGE")
)
as_rtftables(df_rep, collapse_repeats = "Group")[[1]]$data$Group
#> "All" NA "BORC" NA NA NA "HOGE" "HAGE" "HOGE" "HAGE" NA
```

Where does this sit relative to `group_by`? It runs **after** value-based
grouping and **before** the table is shown as `"filled"`: the suppression is
applied **per page, after the split**, so the pagination still sees the original
repeated values. Group boundaries and the `" (Cont.)"` continuation labels stay
correct, and a group continued onto the next page shows its label again at the
top of that page (each page restarts the suppression).

Several columns are suppressed **hierarchically**: the first listed column is
collapsed on its own value, and each later column on the *combination* of itself
with all earlier listed columns -- so a change in any higher column restarts the
lower column's run.

```{r collapse-hier}
df_hier <- data.frame(
  Soc = c("Cardiac", "Cardiac", "Cardiac", "Vascular", "Vascular"),
  Pt  = c("Angina",  "Angina",  "Failure", "Angina",   "Angina")
)
res <- as_rtftables(df_hier, collapse_repeats = c("Soc", "Pt"))[[1]]$data
res$Pt   # row 3 keeps "Failure"; row 5 blanks the repeated "Angina" under Vascular
#> "Angina" NA "Failure" "Angina" NA
```


### Hiding a grouping / carrier column -- `drop_cols`

`group_col` chooses which column drives the split, but that column is normally
*printed* too. Often you want to break or group on a key that should **not**
appear in the report -- a numeric ordering key, a raw group id, a sort helper.
`drop_cols` names columns (in the input body's coordinates -- the same space as
`group_col` and `collapse_repeats`) that are kept **through** the split and then
removed from every page **before** it is rendered.

```{r drop-cols-byvalue}
df_carrier <- data.frame(
  soc_ord = c(1, 1, 2, 2),                  # hidden ordering / group key
  Term    = c("Angina", "Failure", "Rash", "Pruritus"),
  N       = c("12", "7", "9", "4")
)
# Group + page on the hidden key, but print only Term + N.
pages <- as_rtftables(df_carrier, split = "by_value",
                      group_col = "soc_ord", drop_cols = "soc_ord")
names(pages)                 # pages still named by the group value: "1", "2"
names(pages[[1]]$data)       # the carrier column is gone: "Term", "N"
#> [1] "Term" "N"
```

The column is present while the split runs -- so `group_col`,
`collapse_repeats`, and `blank_rows_by_change()` can all reference it -- and only
then dropped. All position-indexed metadata (column headers, including spanning
headers; `col_spec`; column widths; `col_header_align`; `row_title`; per-cell
`cell_styles`) is reindexed automatically to the columns that remain, so explicit
headers/widths you pass for the full table still line up after the hide.

A few interactions worth knowing:

* `drop_cols`, `group_col`, and `collapse_repeats` all index the **pre-drop**
  input body.
* For `split = "by_value"` the page *names* come from the `group_col` value, so
  that column can be hidden and the pages are still named by it (above).
* Under `"group_safe"` / `"group_force"` the `" (Cont.)"` marker is written into
  the `group_col` cell. If you hide that very column the marker is removed with
  it, so group on the *visible* label column when you want the continuation
  label to show.
* `drop_cols` must leave at least one column to display.

To **mix** column names and integer indices, pass a `list()` (a plain `c()`
would coerce to one type): `drop_cols = list("soc_ord", 3)`.


## 2. A custom split function

When the built-ins don't fit, pass a **function** as `split =`. It is called on
the (cell-formatted) body and must return a **list of data.frames** -- one per
page. Named list elements become page names, exactly like `"by_value"`.

```{r custom-basic}
# Fixed blocks of four rows, ignoring groups entirely.
by_four <- function(df, ...) {
  split(df, (seq_len(nrow(df)) - 1L) %/% 4L)
}
pages <- as_rtftables(df, split = by_four)
length(pages)
```

Your function only implements the split; the rest of the pipeline (blank rows,
metadata, per-page assembly, and header/width/style replication) is applied to
its output unchanged. Write it with a `...` so it tolerates the context
arguments (`max_rows`, `group_col`, `group_by`, `cont_label`, `min_group_rows`)
it does not use.

To re-create the `" (Cont.)"` continuation row the built-in group strategies
add, use `add_cont_label()`:

```{r custom-cont}
by_four_cont <- function(df, cont_label = " (Cont.)", ...) {
  parts <- split(df, (seq_len(nrow(df)) - 1L) %/% 4L)
  for (i in seq_along(parts)) {
    if (i > 1L) parts[[i]] <- add_cont_label(parts[[i]], "…continued",
                                             cont_label = cont_label)
  }
  parts
}
show_pages(as_rtftables(df, split = by_four_cont))
```


## 3. Pre-splitting the data yourself

Sometimes the simplest thing is to split the data.frame *before* calling
`as_rtftables()` and hand it a list. `as_rtftables()` flattens a list input into
one page per element, carrying the list's names through as page names.

```{r presplit}
top    <- df[1:8, ]
bottom <- df[9:13, ]

pages <- as_rtftables(list("Demographics (1 of 2)" = top,
                           "Demographics (2 of 2)" = bottom))
names(pages)
```

This is handy when the split rule lives in your own data-wrangling code (for
example `split(df, df$SOC)` for a list of adverse-event tables), or when each
page comes from a different source entirely.


## 4. Advanced: post-processing a built-in split

A custom `split =` function can run the built-in machinery first and then
adjust its output -- a hybrid of approaches 1 and 2. The simplest way is to
call `as_rtftables()` for the split and post-process the pages:

```{r compose}
pages <- as_rtftables(df, split = "group_safe", max_rows = 6)
pages <- Filter(function(p) nrow(p$data) > 0L, pages)   # keep non-empty pages
length(pages)
```


## Blank separator rows

Clinical tables often need **blank separator rows** -- empty rows that put a
little air between groups. They are not part of the data; they are layout, added
on top of the body.

### Blank rows and the row count

This matters for pagination: **`max_rows` counts data rows only.** Blank rows are
inserted into each page *after* the split, so they do not use up the `max_rows`
budget. For example, packing whole groups with `max_rows = 6` and a blank
between groups:

```{r blank-count}
df_g <- data.frame(
  label = c("Group A", "  a1", "  a2", "Group B", "  b1", "  b2",
            "Group C", "  c1", "  c2"),
  v = as.character(1:9), stringsAsFactors = FALSE
)

pages <- as_rtftables(df_g, split = "group_safe", max_rows = 6,
                      blank_rows = "between_groups")
nrow(pages[[1]]$data)          # 6 data rows -- the max_rows budget
pages[[1]]$blank_rows          # plus a blank after row 3 (A -> B), not counted
```

So page 1 shows six data rows **and** a separator -- seven rows on the page --
but only the six data rows counted toward `max_rows`.

To make the blanks count -- so a page (data + blanks) never exceeds `max_rows`
-- set `count_blank_rows = TRUE`. The blank positions are then resolved on the
full table and counted during the split (a leading blank is suppressed at the
top of each page):

```{r blank-counted}
pages2 <- as_rtftables(df_g, split = "group_safe", max_rows = 6,
                       blank_rows = "between_groups", count_blank_rows = TRUE)
vapply(pages2, function(p) nrow(p$data) + length(p$blank_rows), integer(1))
```

Every page now fits within six rows. (`blank_row_first` / `blank_row_end` below
are page furniture added *after* the split and are never counted, so they may
still push a page past `max_rows`.)

### Adding blank rows while paginating

`as_rtftables()` takes the blank-row controls alongside the split arguments;
they are applied to every page:

- `blank_rows = "between_groups"` -- a blank at each group transition on a page;
- `blank_rows = c(0, 3)` -- explicit positions (`0` = before the first row,
  `k` = after row `k`), relative to each page;
- `blank_row_first = TRUE` / `blank_row_end = TRUE` -- a blank at the top /
  bottom of every page.

### Paginating yourself: `set_blank_rows()`

`set_blank_rows()` is the helper `as_rtftables()` calls on every chunk. If you do
your own splitting, use it to attach the blank positions to a page-sized
data.frame; `rtftable(read_attributes = TRUE)` (the default) then picks them up:

```{r set-blank}
page <- set_blank_rows(df_g, blank_rows = "between_groups")
attr(page, "rtf_blank_rows")
tbl <- rtftable(page)          # read_attributes = TRUE consumes the attribute
```

### Blank rows on a single (un-paginated) table

You do not need pagination to use them: `rtftable(blank_rows = )` accepts integer
positions (here `-1` also means "after the last row"), or the spec constructors
[blank_rows_by_change()] (blank when a column value changes) and
[blank_rows_by_rule()] (blank before/after rows matching a regex), or a `list`
combining them:

```{r blank-single}
rtftable(df_g, blank_rows = blank_rows_by_change("label"))
```

Control the separator height with `blank_row_height_twips` (e.g. half the data
row height); see [Page & document setup](page-setup.html). The full spec forms
are documented in `?rtftable` and `?set_blank_rows`.

### Normalising blank rows -- `blank_row_normalize`

Two render-time clean-ups are applied to blank rows by default, controlled by
the single `blank_row_normalize` argument (on `rtftable()`, and forwarded by
`as_rtftables()` / `rtf_tables()`):

| token | effect |
|-------|--------|
| `"detect"`   | a **data** row whose every cell is `NA` / `""` (empty or ASCII-whitespace only) is rendered as a single full-width blank row -- like an explicit separator -- instead of one empty cell per column (which would draw the internal column rules across the gap). |
| `"collapse"` | a run of two or more consecutive blank rows (separators and/or detected empty rows) is reduced to a single blank row. |

The default is `c("detect", "collapse")` -- the combination wanted for most
clinical TFLs. Because both act on the *rendered* table, they apply **per page**
for a paginated table (i.e. after the split), so a stray empty row left at a page
edge, or two separators meeting at a group boundary, collapse cleanly. Pass
`"none"` (or `NULL`) to switch the normalisation off and render blank data rows
verbatim:

```{r blank-normalize}
df_b <- data.frame(
  A = c("x", "", "", "y"),     # two adjacent all-empty rows
  B = c("1", "", "", "2"),
  stringsAsFactors = FALSE
)
rtftable(df_b)                                   # default: one blank row
rtftable(df_b, blank_row_normalize = "none")     # keep both empty data rows
```

Note `"detect"` only fires when **all** columns are empty; an NBSP-indented
label (the clinical sub-row layout) is not whitespace-empty, so indented rows are
never mistaken for blanks.


## Splitting the other way: `paginate_cols()`

Everything above cuts **rows**. A table that is too *wide* needs the other
axis: `paginate_cols()` splits it by **column** and repeats the row-heading
column(s) on every page.

```{r pagcols, eval = FALSE}
as_rtftables(df, split = "group_safe", max_rows = 20) |>
  paginate_cols(at = c(4, 6))
```

`at` names the columns to cut **before** (so `c(4, 6)` gives blocks `1:3`,
`4:5`, `6:ncol`), symmetric with `split_rows`; `cols = list(2:3, 4:5)` states
the blocks outright. Positions refer to the **final printed columns** -- after
`drop_cols`, `stub_vars` and any `col_header` you supplied -- the same
convention `set_col_header()` follows.

Row splitting happens first and the column split is applied to its result, so
the **row page is the outer level**: a row band sweeps every column block
before the next band starts, so the reader goes across the table first and then
down it. Two row pages by three column blocks come out as

```
page 1  row 1 / cols 1    page 4  row 2 / cols 1
page 2  row 1 / cols 2    page 5  row 2 / cols 2
page 3  row 1 / cols 3    page 6  row 2 / cols 3
```

The columns repeated on each page default to the table's `row_title` (column 1
unless you set it); `carry =` overrides that, and `carry = integer(0)` repeats
nothing. A carry column is removed from the blocks automatically, so it is
never printed twice on one page.

### Column widths after the split -- `width`

`column_widths_twips` is absolute: a subset already carries the right widths
and nothing is rescaled. **Relative** widths need a rule, because
`.compute_cellx()` re-normalises whatever it is given across the page -- a bare
subset would stretch the kept columns to refill the sheet, making a ratio-1
column a different size on every page.

`width = "fill"` (the default) fixes the **twips per ratio unit on page 1** and
reuses it everywhere. Page 1 -- and every block with the same ratio total --
fills the sheet, while a given ratio is the same width throughout. With
`rel = c(3, 1, 1, 1, 1, 1, 1, 1, 1)` on a 13680-twip page:

| blocks | unit | page widths |
|---|---|---|
| 4 + 4 | 1954.3 | 100% / 100% |
| 4 + 3 | 1954.3 | 100% / 85.7% |
| 2+2+2+2 | 2736.0 | 100% on all four |

```{r pagcols_width, eval = FALSE}
tbl <- rtftable(df, col_rel_width = c(3, rep(1, 8)))
paginate_cols(tbl, at = 6)
# page 1: 5863 1954 1954 1954 1955   (13680 -- full)
# page 2: 5863 1954 1954 1954 1955   (13680 -- full)
```

`width = "keep"` measures against the whole table instead, so a kept column has
exactly the width it had before the split and a partial block yields a
proportionally shorter page:

```{r pagcols_keep, eval = FALSE}
paginate_cols(tbl, at = 6, width = "keep")
# page 1: 3731 1244 1244 1244 1242   (8705 -- 64% of the sheet)
```

**Order the blocks widest-first.** A block totalling more ratio than block 1
scales past the sheet under `"fill"` -- three visits then five, say, puts page 2
at 133% of the width. `paginate_cols()` warns and names the blocks; either
reorder the cut or use `"keep"`.

Spanning header cells are clipped to each page's columns. By default a cut may
fall inside a spanning group, and that group's label is repeated over its
remaining columns on both pages; pass `allow_span_break = FALSE` to reject such
a cut instead and be told which cell it would break.

Page names are carried through unchanged, so a table's column pages stay in the
same section under `rtf_tables(auto_section = TRUE)`.

### Worked example: a PK concentration summary

`data-raw/gen_pk_conc.R` builds the case that needs all of this at once -- a
pharmacokinetic concentration table with the **visits across the columns**, so
it is too wide *and* too tall for one page:

```{r pkconc, eval = FALSE}
# ABSOLUTE widths, from what the cells actually need
widths <- auto_col_widths(pk, col_header = names(pk))

as_rtftables(
    pk,
    split = "group_safe", group_by = "indent", max_rows = 21,
    col_header = list(
      list(col_cell(c(2, length(VISITS) + 1), "Plasma Concentration (ng/mL)")),
      c("Nominal Time (h)", VISITS)
    ),
    column_widths_twips = widths,
    border = "tfl"
  ) |>
  set_decimal_split(cols = VISITS) |>   # one column, two cells
  paginate_cols(at = length(VISITS) / 2 + 2)   # visits cut down the middle
```

Three time bands by two visit blocks gives six pages, each with the nominal
time stub, and every visit column the width it has in the full table.

#### Size the table from its content, not from the page

Note the `column_widths_twips` rather than `col_rel_width`. **Relative widths
cannot produce a table that needs `paginate_cols()`**: `.compute_cellx()`
normalises them across the page, so whatever the ratios the full table comes
out exactly as wide as the sheet, and the split would be decorative.
`auto_col_widths()` reports what the content needs instead:

```
stub "Nominal Time (h)"   2021 twips   (widest cell "Min, Max")
each visit column         1804 twips   (widest cell "496.28, 1461.8")
```

so only six visits fit on a landscape sheet's 13680 writable twips. The study
has twelve, which makes the table `2021 + 12 x 1804 = 23669` twips -- **1.73x
too wide to fit** -- and each half `2021 + 6 x 1804 = 12845`, or **94% of the
page**. That is what the example should look like: a table that genuinely
cannot be printed in one piece, cut into blocks that each fill the sheet.

`as_rtftables(auto_width = TRUE)` is deliberately not used here: it clamps the
total to the writable width when the natural size exceeds it, squeezing the
columns back onto one page.

#### What the split does to the numbers

PK reports carry three significant figures, so the decimal count changes with
the magnitude (`1104.5`, `88.012`, `9.0125`) -- `set_decimal_split()` lines the
points up regardless, `<0.500` hangs its `<` to the left of the point, and
`BLQ` or a `Min, Max` pair falls back to a single merged cell.

One consequence is worth knowing: the split decision is made **per page**. A
visit column that is all `BLQ` within one time band carries no decimal
separator there, so that page leaves the column unsplit while the other pages
split it. Each page stays internally consistent, which is what the geometry
needs.

## Choosing an approach

| You want… | Use |
|-----------|-----|
| Everything on one page | `split = "none"` (default) |
| Breaks at known row numbers | `split = "rows"` + `split_rows` |
| As many rows as fit, groups may split (with a continuation label) | `split = "group_force"` + `max_rows` |
| As many *whole groups* as fit, never split a group | `split = "group_safe"` + `max_rows` |
| One page per group, named | `split = "by_value"` |
| A bespoke rule | a custom `split =` function (+ `add_cont_label()`) |
| The split already done upstream | pass a (named) list to `as_rtftables()` |
| A built-in rule plus a tweak | split with a built-in, then post-process the pages |
| The table is too **wide** for the page | `paginate_cols()` (after the row split) |

Whichever you choose, the result is the same shape -- a list of `rtftable`
pages -- so the rest of your pipeline does not change:

```{r pipeline, eval = FALSE}
doc <- rtf_document() |>
  rtf_section(page = 1, secinfo = list(header = NULL, footer = NULL)) |>
  rtf_tables(as_rtftables(df, split = "group_safe", max_rows = 6))

generate_rtfreport(doc, "demographics.rtf", overwrite = TRUE)
```

See `?as_rtftables` for the full `split =` contract and `?page_split` for the
factory functions.

<!-- where-next -->

## Where next

- [Output options](output.html) --- rendering the pages you just cut

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
