---
title: "The plan verbs"
description: >
  What each plan_*() verb declares, grouped by job, with the rule that a later
  layer wins.
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
options(width = 100)
knitr::opts_knit$set(root.dir = Sys.getenv("RTFREPORTER_ARTICLE_OUT", tempdir()))
# The screenshots are committed in inst/rtf-examples/plan/ (made by
# data-raw/gen_plan_articles.R) and copied next to the article to be served.
.here <- normalizePath(".")          # the article's folder, before root.dir moves
.shot <- function(file) {
  src <- file.path(.here, "..", "..", "inst", "rtf-examples", "plan", file)
  if (!file.exists(src)) return(invisible(NULL))
  dir.create(file.path(.here, "figures", "plan"), showWarnings = FALSE,
             recursive = TRUE)
  file.copy(src, file.path(.here, "figures", "plan", file), overwrite = TRUE)
  knitr::include_graphics(file.path("figures", "plan", file), error = FALSE)
}
```

A plan is built from small declarations. [Tables from an ARD](tables-from-ard.html)
walks through one table from start to finish. This article is the reference
to keep open while you write your own: **which verb says what**, grouped by
the job it does, with a short example of each.

Two rules hold for every verb:

1. **Each verb has one job.** Its arguments have the names of the function
   that does that job: `widen_ard()` for the cells, `as_rtftables()` for the
   pages, `set_col_header()` for the header. If you know those functions, you
   already know the arguments.
2. **A later layer wins.** A verb can be called more than once. For the same
   key, the last call decides, so a house style is written once and one table
   changes it by adding a line.

```{r setup}
library(rtfreporter)
library(cards)

arms <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- pharmaverseadam::adsl
adsl <- adsl[adsl$SAFFL == "Y", ]
adsl$TRT01A <- factor(adsl$TRT01A, levels = arms)

ard <- ard_stack(
  adsl, .by = TRT01A,
  ard_summary(variables = c(AGE, TRTDURD),
                 statistic = ~ continuous_summary_fns(
                   c("N", "mean", "sd", "median", "min", "max"))),
  ard_tabulate(variables = c(SEX, RACE)))

base <- ard |>
  normalize_ard() |>
  table_plan(cols = "TRT01A", rows = c(group = "variable"))
```

`base` has the roles and nothing else. Each section below adds to it.

## At a glance

| Job | Verb |
|---|---|
| The roles: what goes across, what goes down | `table_plan(x, cols, rows, label, stat)` |
| How a cell is made | `plan_cells(..., stats, value, na, notes)` |
| The digits | `plan_digits(..., rounding)` |
| The order and text of values | `plan_levels()`, `plan_labels()` |
| The row order | `plan_sort(..., stat, keep)` |
| The row headings | `plan_stub(vars, name, indent, group_summary, before)` |
| Groups down the body | `plan_row_group(mode, collapse)` |
| Blank rows | `plan_blanks(where, first, last, counted)` |
| Columns not printed | `plan_hide(...)` |
| The columns: widths, decimal alignment | `plan_columns(widths, decimal, row_title, auto_width, sep)` |
| The column header | `plan_col_header(header, values)` |
| The whole table's look | `plan_style(border, align_count_pct, font, ..., border_header, ...)` |
| Some cells' look | `plan_cell_style(cols, header, where, bold, italic, align, color, background, border)` |
| A page per value | `plan_paginate_group(col, keep)` |
| How many rows on a page | `plan_paginate_rows(max_rows, split, break_before, min_group_rows, cont_label)` |
| A table too wide for the page | `plan_paginate_cols(at, cut_by, every, keep, col_header, fit, allow_span_break, order)` |
| A listing | `plan_listing(..., type, sep, spacer, spacer_rel_width, layout, wrap)` |
| Titles and footnotes | `plan_titles()`, `plan_footnotes()` |
| Anything else | `plan_after(...)` |
| Run it, look inside | `plan_apply(plan, stage)`, `plan_layers()`, `print()` |

## The cells

### `plan_cells()`: templates

A template is text with `{statistic}` tokens. `{mean:.1f}` has one decimal
and `{p:.1f%}` is a percentage. The shape of the value says how many rows it
makes:

* one string: one row per record, labelled by the record (a category level);
* a **named** vector: one row per name, and the name is the row label;
* a vector with a **guard**, `condition ~ template`: the first element that
  applies wins.

Entries are keyed by an analysis variable, by the kind of variable
(`continuous`, `categorical`) or left unnamed as the default. The most
specific key wins:

```{r cells}
p <- base |>
  plan_cells(
    continuous  = c("n"         = "{N:.0f}",
                    "Mean (SD)" = "{mean:.1f} ({sd:.2f})",
                    "Median"    = "{median:.1f}",
                    "Min, Max"  = "{min:.1f}, {max:.1f}"),
    categorical = c(n == 0 ~ "0", "{n:.0f} ({p:.1f%})"),
    AGE         = c("n" = "{N:.0f}", "Mean (SD)" = "{mean:.1f} ({sd:.2f})"),
    notes = FALSE)
plan_apply(p, "table")[1:7, ]
```

`AGE` has its own entry, so it shows two rows. `TRTDURD` (treatment
duration) uses the `continuous` one.

`stats = "rows"` gives each statistic a row of its own, labelled by the
statistic, without templates. This is the shape of a PK concentration table.
`na` is the text for a cell with no value. `notes = FALSE` turns off the
message listing statistics that no template used.

### `plan_digits()`: the digits, as a house rule

A token that leaves its format open (`{mean}`, `{p:%}`) takes its digits from
`plan_digits()`. The digits can be keyed by statistic and by variable, and a
later call wins, so a house rule and its exceptions are separate lines:

```{r digits}
p <- base |>
  plan_cells(continuous = c("Mean (SD)" = "{mean} ({sd})", "Median" = "{median}"),
             categorical = "{n} ({p:%})", notes = FALSE) |>
  plan_digits(continuous = c(mean = 1, sd = 2, median = 1),
              categorical = c(p = 1)) |>                    # the house rule
  plan_digits(TRTDURD = c(mean = 0, sd = 1))               # one variable
plan_apply(p, "table")[1:6, ]
```

A token that sets its own format (`{mean:.2f}`) keeps it.
`plan_digits(rounding = "sas")` rounds halves away from zero as SAS does. The
default rounds as R does -- or as `options(rtfreporter.rounding = )` says, the
one rule every formatter in the package shares. `round_num()` is that rule on
its own, for numbers you round yourself:

```{r round-num}
round_num(c(0.5, 1.5, 2.5))                    # R: half to even
round_num(c(0.5, 1.5, 2.5), rounding = "sas")  # SAS: half away from zero
```

### `plan_levels()` and `plan_labels()`

`plan_levels()` sets the order of a key's values, or of a variable's levels.
`plan_labels()` sets how a key's values print. Here that is the analysis
variables' names in the `group` column. Both merge one key at a time, so a
later call adds an entry without repeating the others:

```{r labels}
p <- base |>
  plan_cells(continuous = c("Mean (SD)" = "{mean:.1f} ({sd:.2f})"),
             categorical = "{n:.0f} ({p:.1f%})", notes = FALSE) |>
  plan_labels(c(AGE = "Age (years)", TRTDURD = "Treatment duration (days)")) |>
  plan_labels(c(SEX = "Sex", RACE = "Race")) |>
  plan_levels(SEX = c("M", "F"))
plan_apply(p, "table")[, 1:3]
```

When no `plan_levels()` gives an order for the `group` column, the order of
`plan_labels()` becomes its order. So label every variable, in the order you
want them. A variable left out moves to the end.

`plan_levels(.drop_empty = )` names variables whose levels no subject has --
`n` 0 in every column -- and leaves those rows out; a level some column counts
stays.  An ARD made after a code list's full set of levels was applied counts
an unused value as 0, which a table shows by default.

The text of a **category level** (`F`, `M`) is the data's. Recode it where the
ARD is made, for example as a factor's labels, and the table prints what the
ARD holds.

### `plan_sort()`: the row order

Rows keep the order the data and `plan_levels()` give them.
`plan_sort()` names keys to sort by, most important first. A minus sign means
descending, and a statistic name sorts by that statistic summed across the
columns. `.overall` is the "any event" row and `.depth` a level's own row
before the rows under it. The usual adverse-event order is "any event
first, then each SOC with its own row first and its PTs most frequent first":

```r
plan_sort(".overall", "SOC", ".depth", "-n", "PT")
```

Keep a hierarchy's outer key before the statistic. `-n` alone orders every
row by its count, and the PTs lose their SOC.

`keep = FALSE` also hides the columns used only for sorting.

## The rows

### `plan_stub()`: the row headings

A demographics table usually prints the variable and its levels in one
indented column, the **stub**. `plan_stub()` folds the row keys into it:
`name` is the column's name, and `indent` the number of spaces for each
level.

`before = TRUE` folds the stub before the table is paginated, so that
`plan_cell_style(where = )` can see the printed rows. Leave it `FALSE` (the
default) when a page split must still read a row key.

### `plan_row_group()`, `plan_blanks()`, `plan_hide()`

* `plan_row_group(mode = "indent")` treats each heading and the rows under it
  as a group, so pagination keeps them together. `collapse = TRUE` prints a
  repeated value once.
* `plan_blanks(where = "between_groups", first = TRUE, last = TRUE)` adds a
  blank row between groups and at the top and bottom of the body.
* `plan_hide("COL")` keeps a column in the data but does not print it.

## The columns and the header

### `plan_columns()`

`widths` are relative. When they are named, a reordered table keeps them, and
`.values` stands for every value column. `decimal` lines numbers up on the
decimal point:

```r
plan_columns(widths = c(row_label = 4, .values = 2), decimal = ".values")
```

### `plan_col_header()`

The header is an ordinary `rtf_col_header()`, and its cells can use tokens:

* `{col}`: the column's key value (the arm); with several `cols` keys,
  `{col1}`, `{col2}`, ... are the levels;
* `{n}`: the column's population. `values = list(n = TRUE)` reads it from the
  ARD. Here that is the arm counts `ard_stack(.by = )` wrote;
* `{n:sum}`: the total over the columns a cell spans, for a spanner.

A header row shorter than the table has its last cell repeated across the
value columns, so the header does not need to know how many arms there are:

```{r header}
p_hdr <- base |>
  plan_cells(continuous = c("Mean (SD)" = "{mean:.1f} ({sd:.2f})"),
             categorical = "{n:.0f} ({p:.1f%})", notes = FALSE) |>
  plan_stub(name = "row_label") |>
  plan_col_header(values = list(n = TRUE), rtf_col_header(
    list(col_cell(1, ""), col_cell(c(2, 4), "All arms (N={n:sum})")),
    c("",               "{col}"),
    c("Characteristic", "(N={n})")))
p_hdr
```

Printing the plan lists every token and the value it will take, so you can
check a header before anything is run. If a population is not in the ARD, the
header prints `NA` and a warning says which cells. Give the numbers yourself
with `values = list(n = c(Placebo = 86, ...))`.

## The look

### `plan_style()`: the whole table

`plan_style()` takes the whole-table settings `rtftable()` and
`as_rtftables()` have: `border`, `font_size_half_points`, `row_height_twips`,
`align_count_pct`, and the rules of each kind of row (`border_header`,
`border_body`, `border_first_row`, ...).

### `plan_cell_style()`: some cells

`plan_cell_style()` chooses cells with `cols`, `header` or `where`, and says
how they look. A value applies to every chosen cell. A **one-sided formula**
is computed row by row, and `NA` leaves a cell as it is:

```{r cell-style}
p_style <- base |>
  plan_cells(continuous = c("Mean (SD)" = "{mean:.1f} ({sd:.2f})"),
             categorical = "{n:.0f} ({p:.1f%})", notes = FALSE) |>
  plan_labels(c(AGE = "Age (years)", TRTDURD = "Treatment duration (days)",
                SEX = "Sex, n (%)",
                RACE = "Race, n (%)")) |>
  plan_stub(name = "row_label", before = TRUE) |>
  plan_blanks(where = "between_groups", first = TRUE, last = TRUE) |>
  plan_columns(widths = c(row_label = 4, .values = 2)) |>
  plan_style(border = "tfl", align_count_pct = TRUE) |>
  plan_cell_style(header = TRUE, bold = TRUE) |>
  plan_cell_style(cols = "row_label", bold = ~ is.na(Placebo)) |>   # the headings
  plan_col_header(values = list(n = TRUE), rtf_col_header(
    c("",               "{col}"),
    c("Characteristic", "(N={n})")))

doc <- rtf_document() |>
  rtf_section(secinfo = list(header = rtf_header(list(
    c(c = "Table 14.1.2  Demographics and Treatment Duration"),
    c(c = "Safety Analysis Set"), c(c = ""))))) |>
  rtf_tables(p_style)
generate_rtfreport(doc, "t_style.rtf", overwrite = TRUE)
```

```{r shot-style, echo = FALSE}
.shot("t_style.png")
```

The heading rows are the ones with no numbers, so `is.na(Placebo)` picks
them. The formula can see the heading rows only because the stub is folded
first (`before = TRUE`). Without that, they do not exist yet when the
styles are worked out.

## Pages

Three verbs cover the three directions in which a table can split into pages:

* **by value**: `plan_paginate_group(col)` starts a new page for each value of
  a column (a lab parameter, a visit). Left out, `col` is the outermost row
  key. The page is named after the value, which is what
  `rtf_tables(auto_section = TRUE)` starts a section on.
* **by rows**: `plan_paginate_rows(max_rows, split = "group_safe")` sets how
  many rows fit on a page and keeps groups whole.
* **by columns**: `plan_paginate_cols(every = 4, keep = "row_label")` splits
  a table too wide for the page into blocks. Each block repeats the `keep`
  columns.

A vital signs table with a page per parameter, and two columns (value and
change from baseline) under each arm:

```{r vs}
visits <- c("Baseline", "Week 8", "Week 16", "Week 24")
advs <- pharmaverseadam::advs
advs <- advs[advs$PARAMCD %in% c("SYSBP", "DIABP") &
               advs$ATPT %in% "AFTER LYING DOWN FOR 5 MINUTES" &
               (advs$ABLFL %in% "Y" | advs$ANL01FL %in% "Y") &
               advs$AVISIT %in% visits & advs$USUBJID %in% adsl$USUBJID, ]
advs$TRTA   <- factor(advs$TRT01A, levels = arms)
advs$AVISIT <- factor(advs$AVISIT, levels = visits)

stats <- ~ continuous_summary_fns(c("N", "mean", "sd"))
ard_vs <- bind_ard(
  ard_summary(advs, by = c(PARAM, AVISIT, TRTA), variables = AVAL,
                 statistic = stats),
  ard_summary(droplevels(advs[advs$AVISIT != "Baseline", ]),
                 by = c(PARAM, AVISIT, TRTA), variables = CHG,
                 statistic = stats))

p_vs <- ard_vs |>
  normalize_ard() |>
  table_plan(cols = c("TRTA", "variable"),
             rows = c(PARAM = "PARAM", Visit = "AVISIT")) |>
  plan_levels(variable = c("AVAL", "CHG")) |>
  plan_cells(continuous = c("n" = "{N:.0f}", "Mean (SD)" = "{mean:.1f} ({sd:.2f})"),
             na = "", notes = FALSE) |>
  plan_stub(name = "Visit") |>
  plan_paginate_group() |>                         # a page per parameter
  plan_blanks(where = "between_groups", first = TRUE, last = TRUE) |>
  plan_columns(widths = c(5, rep(2, 6))) |>
  plan_style(border = "tfl") |>
  plan_col_header(values = list(n = c(table(adsl$TRT01A))), rtf_col_header(
    c(list(col_cell(1, "")),
      lapply(1:3, function(i) col_cell(c(2 * i, 2 * i + 1), "{col1} (N={n})"))),
    c("Visit", rep(c("Value", "Change from\nbaseline"), 3))))

doc <- rtf_document(page = rtf_page(orientation = "landscape")) |>
  rtf_section(secinfo = list(header = rtf_header(list(
    c(c = "Table 14.3.5.1  Vital Signs by Visit"),
    c(c = "Safety Analysis Set"), c(c = ""))))) |>
  rtf_tables(p_vs)
generate_rtfreport(doc, "t_vs.rtf", overwrite = TRUE)
length(plan_apply(p_vs))
```

```{r shot-vs, echo = FALSE}
.shot("t_vs.png")
```

Here the arm counts come from ADSL (`values = list(n = ...)`), because the
columns are arm × variable and the vital-signs ARD does not state the arms'
populations.

## Titles, footnotes, and the way out

`plan_titles()` and `plan_footnotes()` put the same block on every page. With
`pages =`, you give one block for each page instead. Titles are often shared
by a whole report and set in the running header, as in the examples above.
These verbs are for the text that belongs to one table.

`plan_after(function(pages) ...)` runs your own functions on the finished
pages. It is for what no verb declares. Prefer a verb when there is one: a
declaration names columns and so survives a reordered table, while a function
written against positions does not. A plan written with verbs alone can also
be saved to a table spec by tflspec; a `plan_after()` step cannot.

## Looking inside

* `print(plan)`: the layers, the column names the roles may use, and the
  header tokens with their values.
* `plan_apply(plan, stage = )`: `"input"` is the frame going in, `"table"`
  the table data frame, `"args"` the calls the plan resolves to (without
  running them), `"pages"` the pages.
* `plan_layers(plan)`: the plan read back, layer by layer.

`stage = "args"` answers "which of my lines decided this?". It shows the
argument lists after every layer has been merged:

```{r args}
a <- plan_apply(p_vs, "args")
names(a)
str(a$rtf[c("split", "blank_rows", "blank_row_first", "blank_row_end")])
```
