---
title: "From as_rtftables() to a plan"
description: >
  Which plan verb takes over each as_rtftables() argument, and how to see the
  call a plan resolves to.
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
options(width = 100)
knitr::opts_knit$set(root.dir = Sys.getenv("RTFREPORTER_ARTICLE_OUT", tempdir()))
```

If you already write `as_rtftables()` calls, you already know most of the
plan. The display half of a plan does not add new settings: each verb passes
its arguments on to `as_rtftables()`, `rtftable()`, `stub_cols()`,
`set_col_header()` or `paginate_cols()`, under the same names. This article
lists where each `as_rtftables()` argument went, and shows how to see the call
a plan makes.

```{r setup}
library(rtfreporter)
```

## A plan on a table you already have

A plan does not have to start from an ARD. Given a data frame that is
already the table, `table_plan()` keeps it as it is and the plan uses only its
display half. This is a small laboratory summary in the usual shape: a
parameter, a statistic label, and one column per arm.

```{r data}
lab <- data.frame(
  PARAM   = rep(c("ALT (U/L)", "AST (U/L)", "Creatinine (umol/L)"), each = 4),
  STAT    = rep(c("n", "Mean (SD)", "Median", "Min, Max"), 3),
  Placebo = c("86", "17.6 (8.21)", "15.0", "6, 59",
              "86", "22.4 (6.12)", "21.0", "11, 50",
              "86", "95.3 (18.9)", "93.0", "58, 159"),
  Active  = c("168", "18.4 (9.03)", "16.0", "5, 71",
              "168", "23.1 (7.40)", "22.0", "10, 66",
              "168", "96.1 (20.2)", "94.0", "53, 177"))
```

The same pages, written both ways:

```{r both}
by_hand <- as_rtftables(
  lab,
  group_by = "value", collapse_repeats = TRUE,
  blank_rows = "between_groups", blank_row_end = TRUE,
  max_rows = 10, split = "group_safe",
  border = "tfl", read_meta = FALSE)

as_plan <- table_plan(lab) |>
  plan_row_group(mode = "value", collapse = TRUE) |>
  plan_blanks(where = "between_groups", last = TRUE) |>
  plan_paginate_rows(max_rows = 10, split = "group_safe") |>
  plan_style(border = "tfl") |>
  plan_apply()

length(as_plan)
all.equal(by_hand, as_plan)
```

Neither names the group column. With `group_by = "value"`, a run of equal
values in the first column is a group. A plan built from an ARD names it
for you: it is the outermost row key of `table_plan(rows = )`.

## The call a plan makes

`plan_apply(plan, stage = "args")` gives the arguments the plan resolves to,
without running anything. `$rtf` is the `as_rtftables()` call:

```{r args}
p <- table_plan(lab) |>
  plan_row_group(mode = "value", collapse = TRUE) |>
  plan_blanks(where = "between_groups", last = TRUE) |>
  plan_paginate_rows(max_rows = 10, split = "group_safe") |>
  plan_style(border = "tfl")
str(plan_apply(p, "args")$rtf)
```

This is the quickest way to find which line of a long plan decided a
setting. Every layer has been merged, so what you see is what runs.

## Where each argument went

**Pages**

| `as_rtftables()` | Plan |
|---|---|
| `max_rows`, `split`, `split_rows`, `min_group_rows`, `cont_label` | `plan_paginate_rows(max_rows, split, break_before, min_group_rows, cont_label)` (`split_rows` is `break_before`) |
| `split = "by_value"`, `group_col` | `plan_paginate_group(col)`: a page per value |
| `page_by` | no verb; use `plan_paginate_group()` for a page per value |
| then `paginate_cols(pages, ...)` | `plan_paginate_cols(at, cut_by, every, keep, col_header, fit, allow_span_break, order)` |

**Rows**

| `as_rtftables()` | Plan |
|---|---|
| `group_by`, `collapse_repeats` | `plan_row_group(mode, collapse)` |
| `group_col` | the outermost row key of `table_plan(rows = )`; `plan_paginate_group(col = )` for a page split |
| `sort_by`, `sort_desc` | `plan_sort("KEY", "-OTHER")`: a minus sign for descending |
| `blank_rows`, `blank_row_first`, `blank_row_end`, `count_blank_rows` | `plan_blanks(where, first, last, counted)` |
| `drop_cols` | `plan_hide(...)`; also `keep = FALSE` on `plan_sort()` / `plan_paginate_group()` |
| `stub` (and the deprecated `stub_vars`, `stub_label`, `stub_indent`, `stub_group_summary`) | `plan_stub(vars, name, indent, group_summary)` |
| `listing` | `plan_listing(listing_col(...), ...)` |

**Columns and header**

| `as_rtftables()` / other | Plan |
|---|---|
| `auto_width` | `plan_columns(auto_width = )` |
| `rtftable(col_rel_width = )`, `row_title` | `plan_columns(widths = , row_title = )`; named widths follow the columns |
| `set_decimal_split(pages, cols = )` | `plan_columns(decimal = )` |
| `set_col_header(pages, header, values = )` | `plan_col_header(header, values)` |
| `header_sep` | the separator of several `cols` keys: `plan_columns(sep = )` |

**Look**

| `as_rtftables()` / other | Plan |
|---|---|
| `border`, `align_count_pct`, and `rtftable()`'s `font`, `font_size_half_points`, `row_height_twips`, `cell_valign`, `markup`, ... | `plan_style()`, same names |
| `style = rtf_table_style(header = , body = , ...)` | `plan_style(border_header = , border_body = , ...)` |
| `style_header()`, `style_cols()` | `plan_cell_style(header = TRUE, ...)`, `plan_cell_style(cols = , ...)` |
| `style_zone()` | `plan_style(border_* = )` |
| `rtftable(cell_styles = )`, `style_body()` by condition | `plan_cell_style(where = ~ ..., ...)`, or a formula value |
| `realign_count_pct()` | `plan_style(align_count_pct = TRUE)` |
| `fmt_numeric()` on the data | `plan_digits(<column> = 2)`, `plan_digits(.rows = c(Mean = 1))` |

**Handled for you**

| `as_rtftables()` | In a plan |
|---|---|
| `read_meta` | `FALSE`: a table from an ARD is a plain data frame with no adapter metadata |
| `na`, `cell_format` | the cells are text already: `plan_cells(na = )` for an empty cell |

## Rules that differ from a single call

A few things behave differently because a plan is built from layers:

* **A later layer wins** for every setting except one. `plan_hide()` **adds**:
  two `plan_hide()` calls hide both columns, because a later one un-hiding the
  first would be a surprise.
* **Contradictions are errors.** `plan_paginate_group()` with
  `plan_paginate_rows(max_rows = )` says both "a page per value" and "this
  many rows a page". A single call would quietly ignore `max_rows`. A plan
  stops and says which to drop.
* **Declarations name columns, not positions.** `plan_columns(widths =
  c(row_label = 4, .values = 2))` and `plan_cell_style(cols = "Placebo")`
  still apply after a column is added or reordered. `.values` stands for every
  value column.

## When to stay with `as_rtftables()`

Both routes produce the same pages, so choose by what fits the program:

* **A table from gt, rtables, flextable and the like**: `as_rtftables()`
  reads the object's own header, spans and styles. A plan starts from a data
  frame.
* **A table from an ARD**: a plan. It fills the cells, reads the header
  populations from the ARD, and keeps a program short.
* **A study with many tables in one house style**: a plan. The style is a set
  of layers written once, and each table adds its own. tflspec can also save a
  plan to a workbook and build it again from there.

See [Tables from an ARD](tables-from-ard.html) and
[The plan verbs](plan-verbs.html).
