---
title: "Same report, every framework: Demographics (DM)"
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE, eval = FALSE)

# Show a screenshot of the rendered RTF if it has been captured.  The PNGs sit
# next to the .rtf files in inst/rtf-examples/showcase/; we copy the one we need
# into the article's own figures/ folder (so pkgdown serves it) and include it.
# If the screenshot does not exist yet, a small placeholder is shown instead so
# the article still builds.
.snapshot <- function(file, alt = "rendered RTF") {
  src <- file.path("..", "..", "inst", "rtf-examples", "showcase", file)
  if (file.exists(src)) {
    dir.create(file.path("figures", "showcase"), showWarnings = FALSE,
               recursive = TRUE)
    dest <- file.path("figures", "showcase", file)
    file.copy(src, dest, overwrite = TRUE)
    knitr::include_graphics(dest)
  } else {
    knitr::asis_output(sprintf(
      "<p><em>(snapshot <code>inst/rtf-examples/showcase/%s</code> not captured yet)</em></p>",
      file))
  }
}
```

The R ecosystem for clinical tables is wonderfully diverse -- **rtables/tern**,
**tfrmt**, **gtsummary** (on its own or on a **cards/cardx** ARD), **Tplyr**,
**gt**, **flextable**, **huxtable** -- and teams pick whichever fits their style
and validation story. What they all have in common is the *last* step: a
**regulatory RTF** that opens cleanly in Word.

That last step is exactly what `rtfreporter` does, for **all of them**. This
article builds the **same demographics report from the same data** with several
frameworks and renders each to RTF through `rtfreporter`. The point is not that
one framework is best -- it is that **whatever you already use, rtfreporter is
the RTF back end**. A companion article,
[Adverse events](showcase-ae.html), does the same for a longer, *paginated*
adverse-events table.

> **Every block here is copy-paste runnable, from data creation to the written
> `.rtf`.** Run a section's **data** and **furniture** blocks once (they prepare
> the data and the shared rtfreporter header / footer / column spec), then copy
> any single framework block: it builds that framework's table and then calls
> rtfreporter **directly** -- `as_rtftables()`, `rtf_document()`,
> `generate_rtfreport()` -- to write its `.rtf`. We render with `eval = FALSE` in the
> article itself -- so pkgdown does not have to run the entire framework stack --
> and show the Word output as a screenshot. The committed `.rtf` files are
> produced by **running these very chunks**:
> [`data-raw/showcase_dm.R`](https://github.com/ichirio/rtfreporter/blob/main/data-raw/showcase_dm.R)
> extracts them with `knitr::purl()` and writes them to
> [`inst/rtf-examples/showcase/`](https://github.com/ichirio/rtfreporter/tree/main/inst/rtf-examples/showcase)
> -- so the code you read here *is* the generator.

## The data

We use [`pharmaverseadam`](https://pharmaverse.github.io/pharmaverseadam/)'s
`adsl`, restricted to the three randomised arms, and summarise **age** (as `n`,
`Mean (SD)`, `Median`, `Min, Max`), an **age category**, **sex** and **race** by
treatment -- a production-style Table 14.1 demographics layout.

```{r data}
library(rtfreporter)
library(dplyr)

arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- pharmaverseadam::adsl |>
  filter(TRT01A %in% arm_levels) |>
  mutate(
    TRT01A = factor(TRT01A, levels = arm_levels),
    SEX    = factor(SEX, levels = c("M", "F"), labels = c("Male", "Female")),
    AGEGR  = cut(AGE, breaks = c(-Inf, 65, 81, Inf),
                 labels = c("<65", "65 - 80", ">80"), right = FALSE),
    # All race categories as a factor, so even zero-count levels (Asian,
    # American Indian) appear as "0 (0.0%)" in every framework.
    RACE   = factor(RACE, levels = c(
      "WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN",
      "AMERICAN INDIAN OR ALASKA NATIVE"),
      labels = c("White", "Black or African American", "Asian",
                 "American Indian or Alaska Native")))
arm_n <- table(adsl$TRT01A)
```

## The shared rtfreporter furniture, defined once

The running **header** (sponsor, protocol, page number and the table title), the
**footer** (analysis note, program path and run time), the **Letter landscape**
page, the identical column header with the arm Ns, the 40 : 20 : 20 : 20 column
widths and the left / centred alignment are all rtfreporter settings -- the same
for every framework below. We build them **once here**, as plain rtfreporter
objects and values, and every framework block then references them. Only the
table *body* changes from framework to framework.

```{r furniture}
# Running header: sponsor / protocol / page number, then the title block.
dm_header <- rtf_header(rows = list(
  c(l = "Acme Biopharma, Inc.", r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(l = "Protocol ABC-2026-001", r = "Status: Draft"),
  c(c = "Table 14.1.1"),
  c(c = "Summary of Demographic and Baseline Characteristics"),
  c(c = "Safety Population"),
  c(c = "")))                                       # gap before the content

# Footer: analysis note, program path and run time.
dm_footer <- rtf_footer(rows = list(
  c(l = "Note: Percentages are based on the number of non-missing subjects in each treatment group."),
  c(l = "Program: /prod/abc/tfl/t_14_1_1_dm.R", r = "Generated: 2026-06-17 09:14"),
  c(l = "Source: ADSL", r = "CONFIDENTIAL")))

# One identical column header for every framework (arm Ns underneath), the
# 40 : 20 : 20 : 20 relative widths, and row labels left / data columns centred.
dm_col_header <- c("Characteristic",
                   paste0(arm_levels, "\nN = ", as.integer(arm_n)))
dm_col_spec <- c(
  list(list(col = 1L, align = "left",   header_align = "center")),
  lapply(2:4, function(j) list(col = j, align = "center", header_align = "center")))
dm_widths <- c(40, 20, 20, 20)
```

Every framework block below ends with the **same three rtfreporter steps**,
written out in full so nothing is hidden:

1. `as_rtftables()` turns the framework's table into a list of rtfreporter pages,
   taking the shared `dm_col_header` / `dm_col_spec` / `dm_widths` and a blank row
   between groups (`blank_rows = "between_groups"`);
2. `rtf_document() |> rtf_section() |> rtf_tables()` wraps those pages with the
   shared `dm_header` and `dm_footer` on a Letter-landscape page;
3. `generate_rtfreport()` writes the `.rtf`.

## A table per framework

Each framework below builds the demographics **body** its own way; we then hand
that object straight to the shared rtfreporter pipeline. `as_rtftables()` reads
the rendered body (and, with `read_meta = TRUE`, any structural metadata), while
our explicit `dm_col_header` / `dm_col_spec` / `dm_widths` force the one identical
regulatory look across all of them.

### gtsummary

```{r gtsummary}
library(gtsummary)
tbl <- adsl |>
  select(TRT01A, AGE, AGEGR, SEX, RACE) |>
  tbl_summary(
    by = TRT01A,
    type = list(AGE ~ "continuous2"),                       # the 4-line age block
    label = list(AGE = "Age (years)", AGEGR = "Age category",
                 SEX = "Sex", RACE = "Race"),
    statistic = list(
      AGE ~ c("{N_nonmiss}", "{mean} ({sd})", "{median}", "{min}, {max}"),
      all_categorical() ~ "{n} ({p}%)"),
    digits = list(AGE ~ list(N_nonmiss = 0, mean = 1, sd = 2,
                             median = 1, min = 0, max = 0)))
# (No modify_header: rtfreporter supplies the shared column header below.)

# align_count_pct lines up the "n (xx.x%)" cells; the shared furniture supplies
# the header / spec / widths.  Now render straight to RTF:
pages <- as_rtftables(tbl, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_gtsummary.rtf", overwrite = TRUE)
```

```{r img-gtsummary, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("dm_gtsummary.png", "DM summary via gtsummary")
```

### gtsummary on a cards / cardx ARD

The modern, ARD-first workflow: compute an **Analysis Results Dataset** with
`cards`, summarise it with `gtsummary::tbl_ard_summary()`, then hand the table
to `rtfreporter` exactly as above.

```{r gtsummary-ard}
library(cards); library(gtsummary)
ard <- ard_stack(
  data = adsl, .by = TRT01A,
  ard_continuous(variables = AGE,
                 statistic = ~ list(N = length, mean = mean, sd = sd,
                                    median = median, min = min, max = max)),
  ard_categorical(variables = c(AGEGR, SEX, RACE)), .total_n = TRUE)

tbl <- tbl_ard_summary(
  ard, by = TRT01A,
  type = list(AGE = "continuous2"),
  label = list(AGE = "Age (years)", AGEGR = "Age category",
               SEX = "Sex", RACE = "Race"),
  statistic = list(AGE = c("{N}", "{mean} ({sd})", "{median}", "{min}, {max}"),
                   all_categorical() ~ "{n} ({p}%)"))

pages <- as_rtftables(tbl, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_gtsummary_ard.rtf", overwrite = TRUE)
```

```{r img-gtsummary-ard, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("dm_gtsummary_ard.png", "DM summary via gtsummary on a cards/cardx ARD")
```

### rtables / tern

`tern::analyze_vars()` produces a `VTableTree`; `rtfreporter` reads its row
labels, column counts and spanning structure via `formatters::matrix_form()`.

```{r rtables}
library(rtables); library(tern)
lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  analyze_vars(
    vars = c("AGE", "AGEGR", "SEX", "RACE"),
    var_labels = c(AGE = "Age (years)", AGEGR = "Age category",
                   SEX = "Sex", RACE = "Race"),
    .stats = c("n", "mean_sd", "median", "range", "count_fraction"),
    # render the count/percent cells as "n (xx.x%)" to match the other frameworks.
    .formats = c(count_fraction = function(x) {
      sprintf("%d (%.1f%%)", round(x[1]), 100 * x[2])
    }))
tbl <- build_table(lyt, adsl)

pages <- as_rtftables(tbl, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_rtables.rtf", overwrite = TRUE)
```

```{r img-rtables, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("dm_rtables.png", "DM summary via rtables/tern")
```

### cards + tfrmt

The *same* cards ARD as above, rendered with **tfrmt** instead of gtsummary --
so the one ARD drives two different formatters. The ARD is reshaped to tfrmt's
`group / label / column / param / value` long form, a `tfrmt` spec formats it,
and `print_to_gt()` produces the `gt` that `rtfreporter` reads.

```{r cards-tfrmt}
library(cards); library(tfrmt)
ard <- ard_stack(
  data = adsl, .by = TRT01A,
  ard_continuous(variables = AGE,
                 statistic = ~ list(n = length, Mean = mean, SD = sd,
                                    Median = median, Min = min, Max = max)),
  ard_categorical(variables = c(AGEGR, SEX, RACE)))

# Reshape the ARD to tfrmt's long form: group / label / column / param / value.
# sc1()/nm1() pull the first element of each list-column as character / numeric.
d   <- as.data.frame(ard)
sc1 <- function(x) vapply(x, function(v) if (length(v)) as.character(v[[1]]) else NA_character_, character(1))
nm1 <- function(x) vapply(x, function(v) if (length(v)) suppressWarnings(as.numeric(v[[1]])) else NA_real_, numeric(1))

long <- d |>
  mutate(column = sc1(group1_level), vlev = sc1(variable_level), value = nm1(stat)) |>
  filter((variable == "AGE" & stat_name %in% c("n", "Mean", "SD", "Median", "Min", "Max")) |
         (variable %in% c("AGEGR", "SEX", "RACE") & stat_name %in% c("n", "p"))) |>
  transmute(
    column = factor(column, levels = arm_levels),
    group  = recode(variable, AGE = "Age (years)", AGEGR = "Age category",
                    SEX = "Sex", RACE = "Race"),
    label  = case_when(
      variable == "AGE" & stat_name == "n" ~ "n",
      variable == "AGE" & stat_name %in% c("Mean", "SD") ~ "Mean (SD)",
      variable == "AGE" & stat_name == "Median" ~ "Median",
      variable == "AGE" & stat_name %in% c("Min", "Max") ~ "Min, Max",
      TRUE ~ vlev),
    param  = ifelse(stat_name == "p", "pct", stat_name),
    value  = ifelse(stat_name == "p", value * 100, value))

# `long` is now ready.  The two render chunks below each write out their own
# tfrmt spec in full -- they share this `long`, but nothing else.
```

#### Who formats the `n (xx.x%)` cell?

This is a nice place to see the same table formatted **two ways**, because the
count/percent cell is exactly where formatters differ. (Both render chunks
reuse `long` from the block just above, and each writes out its own `tfrmt`
spec in full -- the only line that changes is the count/percent `frmt`.)

A small but important detail first: write the percent as **`x.x`**, not `xx.x`.
`frmt("xx.x")` right-pads the integer part, so a one-digit percent prints with a
leading space *inside* the parentheses -- `( 3.5%)`. `frmt("x.x")` gives a clean
`(3.5%)`.

**Way 1 -- tfrmt formats and aligns.** The percent uses `x.x`; the count is
padded to `xx`. `rtfreporter` reads the strings and renders them *as-is*
(`align_count_pct` left off):

```{r tfrmt-self}
# tfrmt formats AND aligns: the count is padded to "xx", the percent is "x.x".
spec <- tfrmt(
  group = group, label = label, column = column, param = param, value = value,
  body_plan = body_plan(
    frmt_structure(".default", ".default", frmt("xx.x")),
    frmt_structure("Age (years)", "n", frmt("xx")),
    frmt_structure(".default", "Mean (SD)",
                   frmt_combine("{Mean} ({SD})",
                                Mean = frmt("xx.x"), SD = frmt("xx.xx"))),
    frmt_structure(".default", "Min, Max",
                   frmt_combine("{Min}, {Max}",
                                Min = frmt("xx"), Max = frmt("xx"))),
    frmt_structure(c("Age category", "Sex", "Race"), ".default",
                   frmt_combine("{n} ({pct}%)", n = frmt("xx"), pct = frmt("x.x")))))
g1 <- print_to_gt(spec, long)

# tfrmt already aligned the strings; rtfreporter renders them as-is (no
# align_count_pct).  Direct render:
pages <- as_rtftables(g1, read_meta = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_tfrmt.rtf", overwrite = TRUE)   # tfrmt's alignment
```

```{r img-tfrmt, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("dm_tfrmt.png", "DM summary via cards + tfrmt, formatted by tfrmt")
```

**Way 2 -- rtfreporter formats.** tfrmt does the bare minimum -- just clean
`n (x.x%)` strings, no count padding -- and `rtfreporter` re-aligns them with
`align_count_pct = TRUE`:

```{r tfrmt-rtfreporter}
# Same spec as Way 1, except the count is just "x" (no padding): tfrmt emits
# clean "n (x.x%)" strings and leaves the column alignment to rtfreporter.
spec <- tfrmt(
  group = group, label = label, column = column, param = param, value = value,
  body_plan = body_plan(
    frmt_structure(".default", ".default", frmt("xx.x")),
    frmt_structure("Age (years)", "n", frmt("xx")),
    frmt_structure(".default", "Mean (SD)",
                   frmt_combine("{Mean} ({SD})",
                                Mean = frmt("xx.x"), SD = frmt("xx.xx"))),
    frmt_structure(".default", "Min, Max",
                   frmt_combine("{Min}, {Max}",
                                Min = frmt("xx"), Max = frmt("xx"))),
    frmt_structure(c("Age category", "Sex", "Race"), ".default",
                   frmt_combine("{n} ({pct}%)", n = frmt("x"), pct = frmt("x.x")))))
g2 <- print_to_gt(spec, long)

# rtfreporter owns the count/percent alignment via align_count_pct = TRUE.
pages <- as_rtftables(g2, read_meta = TRUE, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_tfrmt_rtfreporter.rtf", overwrite = TRUE)  # rtfreporter aligns
```

```{r img-tfrmt-rtfreporter, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("dm_tfrmt_rtfreporter.png", "DM summary via cards + tfrmt, formatted by rtfreporter")
```

The two are deliberately *not* byte-identical. tfrmt anchors the `(` right after
the count, so the percentages stay left-aligned within the parentheses.
`rtfreporter`'s `align_count_pct` instead right-aligns the whole parenthetical so
the **decimal points line up**, and drops the redundant decimal at 100%
(`100%`, not `100.0%`). Use whichever convention your SAP calls for -- the point
is that you can hand `rtfreporter` already-formatted strings *or* let it own the
count/percent alignment, from the very same tfrmt table.

## When the framework returns only the numbers

Some frameworks return just the **formatted body** -- the numbers as strings --
with no layout. You assemble the labelled body yourself, then hand it to the
*same* rtfreporter pipeline as everything else.

### Tplyr

`Tplyr::build()` returns a data frame of formatted strings (plus ordering
helpers). We reshape it into a labelled body -- group headers with indented
stats and a "Number of Subjects" row on top -- and hand that plain data frame to
the shared rtfreporter render.

```{r tplyr}
library(Tplyr)
built <- tplyr_table(adsl, TRT01A) |>
  add_layer(group_desc(AGE, by = "Age (years)") |>
              set_format_strings("n"         = f_str("xx", n),
                                 "Mean (SD)" = f_str("xx.x (xx.xx)", mean, sd),
                                 "Median"    = f_str("xx.x", median),
                                 "Min, Max"  = f_str("xx, xx", min, max))) |>
  # "x.x" (not "xx.x") so a one-digit percent has no leading space inside the
  # parens -- a clean " 1 (1.2%)" that align_count_pct can re-align.
  add_layer(group_count(AGEGR, by = "Age category") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  add_layer(group_count(SEX, by = "Sex") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  add_layer(group_count(RACE, by = "Race") |>
              set_format_strings(f_str("xx (x.x%)", n, pct))) |>
  build()

# Reshape `built` into `disp`: one column per arm, a label column with each
# group header followed by its indented stat rows, in a fixed display order.
val_cols <- paste0("var1_", arm_levels)
ord <- c("Age (years)", "n", "Mean (SD)", "Median", "Min, Max",
         "Age category", "<65", "65 - 80", ">80",
         "Sex", "Male", "Female",
         "Race", "White", "Black or African American", "Asian",
         "American Indian or Alaska Native")
groups <- c("Age (years)", "Age category", "Sex", "Race")
body <- built[order(built$ord_layer_index, built$ord_layer_1, built$ord_layer_2), ]
disp <- do.call(rbind, lapply(split(body, factor(body$row_label1, groups)), function(g) {
  hdr <- data.frame(Characteristic = g$row_label1[1], stringsAsFactors = FALSE)
  hdr[val_cols] <- ""
  rows <- data.frame(Characteristic = paste0("  ", g$row_label2),
                     stringsAsFactors = FALSE)
  rows[val_cols] <- lapply(val_cols, function(c) g[[c]])
  rbind(hdr, rows)
}))
disp <- disp[order(match(sub("^  ", "", disp$Characteristic), ord)), ]
# "Number of Subjects" line on top (just N).
top <- data.frame(Characteristic = "Number of Subjects", stringsAsFactors = FALSE)
top[val_cols] <- as.character(as.integer(arm_n))
disp <- rbind(top, disp)
names(disp) <- c("Characteristic", arm_levels)
rownames(disp) <- NULL

# Metadata lives HERE, in rtfreporter -- not in Tplyr.  We pass the body plus the
# shared furniture straight to as_rtftables() and render it.
pages <- as_rtftables(disp, align_count_pct = TRUE,
                      col_header = dm_col_header, col_spec = dm_col_spec,
                      col_rel_width = dm_widths, blank_rows = "between_groups")
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = dm_header, footer = dm_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "dm_tplyr.rtf", overwrite = TRUE)
```

```{r img-tplyr, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("dm_tplyr.png", "DM summary via Tplyr")
```

## Two ways to divide the work

The frameworks above split into two camps -- not in how `rtfreporter` renders
them (that pipeline is identical), but in **how much layout the framework hands
you**:

| | **Ready-made body** | **Body you assemble** |
|---|---|---|
| Examples | rtables/tern, gtsummary, cards + gtsummary, cards + tfrmt | Tplyr (and other body-only tools) |
| What the framework gives you | group headers + indented rows, laid out | formatted numbers only |
| Your job before rtfreporter | none -- pass the object | reshape into a labelled body |
| The rtfreporter render | identical | identical |

Either way the regulatory furniture (`dm_col_header` / `dm_col_spec` /
`dm_widths`, plus the shared header and footer) lives **once in rtfreporter**,
and both routes produce the **same RTF**.

<!-- where-next -->

## Where next

- [Showcase: adverse events](showcase-ae.html) --- the hierarchical counterpart
- [Worked example: demographics](example-demog.html) --- the shorter version of this table

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
