---
title: "Same report, every framework: Adverse events (AE)"
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE, eval = FALSE)

# Show a screenshot of the rendered RTF if it has been captured.  The PNGs sit
# next to the .rtf files in inst/rtf-examples/showcase/; we copy the one we need
# into the article's own figures/ folder (so pkgdown serves it) and include it.
# If the screenshot does not exist yet, a small placeholder is shown instead so
# the article still builds.
.snapshot <- function(file, alt = "rendered RTF") {
  src <- file.path("..", "..", "inst", "rtf-examples", "showcase", file)
  if (file.exists(src)) {
    dir.create(file.path("figures", "showcase"), showWarnings = FALSE,
               recursive = TRUE)
    dest <- file.path("figures", "showcase", file)
    file.copy(src, dest, overwrite = TRUE)
    knitr::include_graphics(dest)
  } else {
    knitr::asis_output(sprintf(
      "<p><em>(snapshot <code>inst/rtf-examples/showcase/%s</code> not captured yet)</em></p>",
      file))
  }
}
```

The R ecosystem for clinical tables is wonderfully diverse -- **rtables/tern**,
**tfrmt**, **gtsummary** (on its own or on a **cards/cardx** ARD), **Tplyr**,
**gt**, **flextable**, **huxtable** -- and teams pick whichever fits their style
and validation story. What they all have in common is the *last* step: a
**regulatory RTF** that opens cleanly in Word.

That last step is exactly what `rtfreporter` does, for **all of them**. This
article builds the **same adverse-events report from the same data** with several
frameworks and renders each to RTF through `rtfreporter`. The point is not that
one framework is best -- it is that **whatever you already use, rtfreporter is
the RTF back end**. Its companion,
[Demographics](showcase-dm.html), does the same for a single-page demographics
table; this article tackles a longer table that `rtfreporter` must **paginate**.

> **Every block here is copy-paste runnable, from data creation to the written
> `.rtf`.** Run a section's **data** and **furniture** blocks once (they prepare
> the data and the shared rtfreporter header / footer / column spec), then copy
> any single framework block: it builds that framework's table and then calls
> rtfreporter **directly** -- `as_rtftables()`, `rtf_document()`,
> `generate_rtfreport()` -- to write its `.rtf`. (The data block also defines the
> shared `AE_SORT` rule and PT orderings, so a framework block runs on its own
> once the data block has.) We render with `eval = FALSE` in the
> article itself -- so pkgdown does not have to run the entire framework stack --
> and show the Word output as a screenshot. The committed `.rtf` files are
> produced by **running these very chunks**:
> [`data-raw/showcase_ae.R`](https://github.com/ichirio/rtfreporter/blob/main/data-raw/showcase_ae.R)
> extracts them with `knitr::purl()` and writes them to
> [`inst/rtf-examples/showcase/`](https://github.com/ichirio/rtfreporter/tree/main/inst/rtf-examples/showcase)
> -- so the code you read here *is* the generator.

The demographics table fits on a page; a real **adverse-events** table does not.
This is where `rtfreporter`'s **pagination** earns its keep: a long
System-Organ-Class / Preferred-Term table is split across pages, and a class
that straddles a page boundary has its label **repeated with `(Cont.)`** at the
top of the next page.

We count the **number of subjects** with at least one treatment-emergent AE,
by treatment group, shown as `n (xx.x%)`:

- **"Subjects with any adverse event"** overall on top, then per **System Organ
  Class (SOC)**, and -- indented -- per **Preferred Term (PT)**.
- Each level is an **independent distinct-subject count**: the SOC count sits on
  the **SOC row itself** and counts a subject once even if they had several PTs
  in that class, so a SOC count is **not** the sum of its PTs.
- Limited to PTs occurring in **>= 3% of subjects in any group** (the overall /
  SOC counts are still over *all* TEAEs), so the table is a readable few pages.
- **SOCs are listed alphabetically**; within each SOC the **PTs are ordered by
  subject count (all arms), descending**, ties broken **alphabetically**.

## The data

We use [`pharmaverseadam`](https://pharmaverse.github.io/pharmaverseadam/)'s
`adsl` (for the treatment groups and their denominators) and `adae`, both
restricted to the three randomised arms. This block also defines the canonical
sort rule and the PT orderings/filters that the framework blocks reuse -- each
commented with *which* blocks use it -- so that, once this block has run, any one
framework block below is self-contained.

```{r ae-data}
library(rtfreporter)
library(dplyr)

arm_levels <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- pharmaverseadam::adsl |>
  filter(TRT01A %in% arm_levels) |>
  mutate(TRT01A = factor(TRT01A, levels = arm_levels))
arm_n <- table(adsl$TRT01A)

# Treatment-emergent AEs only, in the same three arms.
adae <- pharmaverseadam::adae |>
  filter(TRT01A %in% arm_levels, TRTEMFL == "Y") |>
  mutate(TRT01A = factor(TRT01A, levels = arm_levels))
ANY_AE <- "Subjects with any adverse event"

# AE_SORT: native per-hierarchy-level sort rule -- SOCs alphanumeric, PTs by total
# distinct-subject count (all arms) descending, ties A -> Z.  USED BY the
# gtsummary-family blocks: gtsummary, cards + gtsummary, and flextable / huxtable.
AE_SORT <- list(AESOC ~ "alphanumeric", AEDECOD ~ "descending")

# The next three are USED BY the transpose-and-set blocks (Tplyr and cards +
# tfrmt), which assemble the body themselves and so need the orderings as data:
#   .pt_order = canonical PT order (subject count desc, ties A -> Z);
#   .keep_pt  = PTs kept (>= 3% of subjects in any arm);
#   .soc_pt   = the real SOC / PT pairs (a PT belongs to exactly one SOC).
.pt_order <- adae |> distinct(USUBJID, AEDECOD) |>
  count(AEDECOD, name = "tot") |>
  arrange(desc(tot), AEDECOD) |> pull(AEDECOD) |> as.character()
.keep_pt <- adae |> distinct(USUBJID, TRT01A, AEDECOD) |>
  count(TRT01A, AEDECOD, name = "n") |>
  mutate(p = 100 * n / as.integer(arm_n[as.character(TRT01A)])) |>
  group_by(AEDECOD) |> summarise(mx = max(p), .groups = "drop") |>
  filter(mx >= 3) |> pull(AEDECOD) |> as.character()
.soc_pt <- adae |> distinct(AESOC, AEDECOD) |>
  mutate(across(everything(), as.character))
```

## The shared rtfreporter furniture, defined once

The running **header** and **footer**, the **Letter landscape** page, the column
header with the arm Ns, the column spec and widths, and the `fmt_ae()` cell
formatter are the same for every framework below. We build them **once here**;
each framework block then references them. Only the table *body* changes from
framework to framework.

```{r ae-furniture}
# Shared column header / spec / widths / page budget.
ae_col_header <- c("System Organ Class /\nPreferred Term",
                   paste0(arm_levels, "\nN = ", as.integer(arm_n)))
ae_col_spec <- c(
  list(list(col = 1L, align = "left",   header_align = "left")),
  lapply(2:4, function(j) list(col = j, align = "center", header_align = "center")))
ae_widths   <- c(0.52, 0.16, 0.16, 0.16)
AE_MAX_ROWS <- 16L   # rows/page: small enough that a SOC is split across a page
                     # (so "(Cont.)" shows) and the table runs to 3 pages

# Running header and the shared footer (one footer for every framework).
ae_header <- rtf_header(rows = list(
  c(l = "Acme Biopharma, Inc.",  r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
  c(l = "Protocol ABC-2026-001", r = "Status: Draft"),
  c(c = "Table 14.3.1"),
  c(c = "Treatment-Emergent Adverse Events Occurring in >= 3% of Subjects in Any Treatment Group"),
  c(c = "by System Organ Class and Preferred Term -- Safety Population"),
  c(c = "")))
ae_footer <- rtf_footer(rows = list(
  c(l = "Note: A subject with multiple events within a category is counted once for that category. Percentages use the number of subjects in the safety population per group."),
  c(l = "Program: /prod/abc/tfl/t_14_3_1_ae.R", r = "Generated: 2026-06-17 09:14"),
  c(l = "Source: ADSL, ADAE", r = "CONFIDENTIAL")))

# fmt_ae(): cell formatter used by EVERY framework (tern included). Expand a bare
# zero "0" to "0 (0.0%)" so it lines up with the count/percent cells -- a bare "0"
# is otherwise left ragged by the aligner -- then align the column. tern's
# count_occurrences() emits bare zeros; gtsummary / tfrmt / Tplyr already emit
# "0 (0.0%)", so for them the expansion is a no-op.
fmt_ae <- function(x, nbsp = " ") {
  x[trimws(x) == "0"] <- "0 (0.0%)"
  fmt_count_paren_bare(x, nbsp = nbsp)
}
```

Every framework block below ends with the **same three rtfreporter steps**,
written out in full so nothing is hidden:

1. `as_rtftables()` paginates the framework's table into a list of pages with
   `split = "group_force"` (`max_rows = AE_MAX_ROWS`), so a System Organ Class
   that straddles a page boundary repeats with `(Cont.)`;
2. `rtf_document() |> rtf_section() |> rtf_tables()` wraps the pages with the
   shared `ae_header` / `ae_footer` on a Letter-landscape page;
3. `generate_rtfreport()` writes the `.rtf`.

Each `as_rtftables()` call passes `read_meta = FALSE`: we supply the regulatory
column header / spec / widths (and the footer, via `rtf_section()`) ourselves, so
a framework's *own* column header -- and its `{n} ({p}%)` statistic footnote --
are deliberately not read. (Tplyr already hands over a plain data frame, which
carries no metadata to begin with.)

## rtables / tern

`tern` is the pharmaverse standard for this table. `analyze_num_patients()`
gives the overall "any AE" row; a SOC-level content summary
(`summarize_row_groups()` with `child_labels = "hidden"`) puts each SOC's
distinct-subject count **on the SOC name row**; `count_occurrences()` gives the
indented PT rows. We build on the full data, **prune** PT rows below 3% with
`rtables::prune_table()`, and **sort** by subject count -- alphabetical factor
levels plus a stable `sort_at_path()` break ties A -> Z. `rtfreporter` then
paginates with **`split = "group_force"`** (`max_rows` per page), which is what
inserts the `(Cont.)` rows.

```{r ae-tern}
library(rtables); library(tern)
# Alphabetical factor levels so SOCs / PTs tie-break A -> Z.
adae_f <- adae |>
  mutate(AESOC   = factor(AESOC,   levels = sort(unique(AESOC))),
         AEDECOD = factor(AEDECOD, levels = sort(unique(AEDECOD))))

# Distinct subjects with any AE in a SOC, per column -- shown on the SOC row.
# Zero counts are emitted bare ("0") here and by count_occurrences() for the PT
# rows; the shared fmt_ae() later expands every "0" to "0 (0.0%)" so the column
# aligns.
soc_count <- function(df, labelstr, .N_col, ...) {
  n <- length(unique(df$USUBJID))
  cell <- if (n == 0L) rcell(0L, format = "xx")
          else         rcell(c(n, n / .N_col), format = "xx (xx.x%)")
  in_rows(cell, .labels = labelstr)
}

ae <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("TRT01A") |>
  analyze_num_patients(vars = "USUBJID", .stats = "unique",
                       .labels = c(unique = ANY_AE)) |>
  split_rows_by("AESOC", child_labels = "hidden", nested = FALSE,
                split_fun = drop_split_levels) |>
  summarize_row_groups(cfun = soc_count) |>              # SOC count on the SOC row
  count_occurrences(vars = "AEDECOD", .indent_mods = 1L) |>
  build_table(df = adae_f, alt_counts_df = adsl) |>
  prune_table(keep_rows(has_fraction_in_any_col(atleast = 0.03))) |>    # >= 3%
  # SOCs stay alphabetical (factor order); only the PTs within each SOC are
  # ordered by subject count (all arms), ties A -> Z.
  sort_at_path(path = c("AESOC", "*", "AEDECOD"), scorefun = score_occurrences)

# group_force = fill each page to max_rows and repeat a split SOC with "(Cont.)"
pages <- as_rtftables(ae, read_meta = FALSE, split = "group_force", max_rows = AE_MAX_ROWS,
                      blank_rows = "between_groups", cell_format = fmt_ae,
                      col_header = ae_col_header, col_spec = ae_col_spec,
                      col_rel_width = ae_widths, row_height_twips = 200)
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = ae_header, footer = ae_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "ae_tern.rtf", overwrite = TRUE)
```

```{r img-ae-tern, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("ae_tern.png", "AE summary via rtables/tern, paginated with (Cont.)")
```

The same `as_rtftables()` call returns a **list of pages**; `rtf_tables()`
renders them in order, and the running header's `{AUTO_PAGE}` /
`{AUTO_TOTAL_PAGES}` number them.

## gtsummary

`gtsummary::tbl_hierarchical()` builds the **same** table directly:
`overall_row = TRUE` is the any-AE row, each SOC row carries its own subject
count, and PTs are nested beneath. `filter_hierarchical()` keeps the >= 3% PTs,
and `sort_hierarchical()` puts them in the canonical order natively -- its `sort`
takes one rule **per hierarchy level** (`AE_SORT`, from the data block: SOCs
alphabetical, PTs by subject count, ties A -> Z). The overall row is labelled to
match tern via the native `label` argument, and the shared `fmt_ae()` expands a
zero cell to a clean, aligned `0 (0.0%)`.

```{r ae-gtsummary}
library(gtsummary)
tbl <- tbl_hierarchical(
    data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
    denominator = adsl, id = USUBJID, overall_row = TRUE,
    label = list("..ard_hierarchical_overall.." ~ ANY_AE),
    statistic = ~ "{n} ({p}%)", digits = everything() ~ list(p = 1)) |>
  filter_hierarchical(p >= 0.03) |>
  sort_hierarchical(sort = AE_SORT)    # SOCs A->Z, PTs by subject count

pages <- as_rtftables(tbl, read_meta = FALSE, split = "group_force", max_rows = AE_MAX_ROWS,
                      blank_rows = "between_groups", cell_format = fmt_ae,
                      col_header = ae_col_header, col_spec = ae_col_spec,
                      col_rel_width = ae_widths, row_height_twips = 200)
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = ae_header, footer = ae_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "ae_gtsummary.rtf", overwrite = TRUE)
```

```{r img-ae-gtsummary, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("ae_gtsummary.png", "AE summary via gtsummary tbl_hierarchical()")
```

## cards + gtsummary

The ARD-first variant: compute the hierarchical **Analysis Results Dataset** with
`cards::ard_stack_hierarchical()`, then summarise it with `tbl_ard_hierarchical()`.
`tbl_ard_hierarchical()` has no `overall_row` argument -- but the any-AE row is
still **the ARD's job, not the raw data's**: a second `ard_stack_hierarchical()`
with **`over_variables = TRUE`** computes the distinct-subject "any event" count
across all the hierarchy variables, and `add_overall_row()` reads it straight off
that ARD (no re-counting of `adae`). Ordering uses the same native
`sort_hierarchical(sort = AE_SORT)`, so the rendered table matches the two above
cell-for-cell.

```{r ae-cards-gtsummary}
library(cards); library(gtsummary)

# add_overall_row(): used ONLY here. tbl_ard_hierarchical() has no overall_row
# argument, so we prepend the "any event" row ourselves -- generically and
# ARD-driven: the overall n / N / p are read straight off an ARD built with
# `ard_stack_hierarchical(..., over_variables = TRUE)` (the rows whose variable
# is "..ard_hierarchical_overall.."), mapped to the table's stat_* columns via
# its own header and formatted with the table's glue `statistic` template.
add_overall_row <- function(tbl, ard, label = ANY_AE,
                            statistic = "{n} ({p}%)", digits = 1L) {
  hdr <- tbl$table_styling$header
  hdr <- hdr[startsWith(hdr$column, "stat_"), ]
  hdr <- hdr[order(hdr$column), ]                  # stat_1, stat_2, ... order
  ov <- ard |>
    dplyr::filter(variable == "..ard_hierarchical_overall..",
                  stat_name %in% c("n", "N", "p")) |>
    dplyr::transmute(level = as.character(unlist(group1_level)),
                     stat_name, stat = unlist(stat)) |>
    tidyr::pivot_wider(names_from = stat_name, values_from = stat)
  cells <- vapply(hdr$modify_stat_level, function(lvl) {
    r <- ov[match(lvl, ov$level), ]
    glue::glue_data(list(n = r$n, N = r$N,
                         p = formatC(100 * r$p, format = "f", digits = digits)),
                    statistic)
  }, character(1))
  tb  <- tbl$table_body
  row <- tb[1, ]; row[seq_len(ncol(row))] <- NA
  row$variable <- "..ard_hierarchical_overall.."
  row$row_type <- "level"
  row$label    <- label
  for (j in seq_len(nrow(hdr))) row[[hdr$column[j]]] <- cells[j]
  tbl$table_body <- dplyr::bind_rows(row, tb)
  # Un-indent the injected row via the public API so it lines up with the SOCs.
  gtsummary::modify_indent(tbl, columns = "label",
                           rows = variable == "..ard_hierarchical_overall..",
                           indent = 0L)
}

# Hierarchical ARD for the SOC / PT body.
ard <- ard_stack_hierarchical(
  data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
  denominator = adsl, id = USUBJID)
# "Any event" ARD: over_variables = TRUE adds the distinct-subject overall count.
ard_any <- ard_stack_hierarchical(
  data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
  denominator = adsl, id = USUBJID, over_variables = TRUE)

tbl <- tbl_ard_hierarchical(
    cards = ard, variables = c(AESOC, AEDECOD), by = TRT01A,
    statistic = ~ "{n} ({p}%)") |>
  filter_hierarchical(p >= 0.03) |>
  sort_hierarchical(sort = AE_SORT) |> # SOCs A->Z, PTs by subject count (native)
  add_overall_row(ard_any)             # any-AE row read from the ARD, generically

pages <- as_rtftables(tbl, read_meta = FALSE, split = "group_force", max_rows = AE_MAX_ROWS,
                      blank_rows = "between_groups", cell_format = fmt_ae,
                      col_header = ae_col_header, col_spec = ae_col_spec,
                      col_rel_width = ae_widths, row_height_twips = 200)
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = ae_header, footer = ae_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "ae_gtsummary_ard.rtf", overwrite = TRUE)
```

```{r img-ae-cards-gtsummary, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("ae_gtsummary_ard.png", "AE summary via cards + gtsummary tbl_ard_hierarchical()")
```

> The three AE tables above are the **same report** -- same numbers, same SOC /
> PT order -- built three different ways and all rendered by `rtfreporter`.

## Tplyr (transpose-and-set)

`Tplyr` returns **formatted strings**; as with the DM table we do the
*transpose-and-set* here -- Tplyr counts the distinct subjects, and we assemble
the layout (overall row, SOC count on the SOC row, indented PTs in subject-count
order) and hand the column header to `rtfreporter` via the setup block's
`ae_col_header` / `ae_col_spec`.

```{r ae-tplyr}
library(Tplyr)
varcols <- paste0("var1_", arm_levels)
rn <- function(df) df |> dplyr::rename_with(~ arm_levels, dplyr::all_of(varcols))

# distinct-subject counts: one layer for SOC totals, one for PT-by-SOC.
soc <- tplyr_table(adae, TRT01A) |> set_pop_data(adsl) |> set_pop_treat_var(TRT01A) |>
  add_layer(group_count(AESOC) |> set_distinct_by(USUBJID) |>
              set_format_strings(f_str("xx (xx.x%)", distinct_n, distinct_pct))) |>
  build() |> dplyr::transmute(AESOC = as.character(row_label1),
                              dplyr::across(dplyr::all_of(varcols))) |> rn()
pt <- tplyr_table(adae, TRT01A) |> set_pop_data(adsl) |> set_pop_treat_var(TRT01A) |>
  add_layer(group_count(AEDECOD, by = vars(AESOC)) |> set_distinct_by(USUBJID) |>
              set_format_strings(f_str("xx (xx.x%)", distinct_n, distinct_pct))) |>
  build() |> dplyr::transmute(AESOC = as.character(row_label1),
                              AEDECOD = as.character(row_label2),
                              dplyr::across(dplyr::all_of(varcols))) |>
  dplyr::semi_join(.soc_pt, by = c("AESOC", "AEDECOD")) |>
  dplyr::filter(AEDECOD %in% .keep_pt) |> rn()

# Assemble `disp`: an overall any-AE row, then each SOC row (its count) and its
# PTs indented in subject-count order.
any_n <- adae |> dplyr::distinct(TRT01A, USUBJID) |> dplyr::count(TRT01A)
ov <- vapply(arm_levels, function(a) {
  n <- any_n$n[match(a, as.character(any_n$TRT01A))]
  sprintf("%d (%.1f%%)", n, 100 * n / as.integer(arm_n[a]))
}, character(1))

rows <- list(setNames(data.frame(ANY_AE, t(ov), check.names = FALSE,
                                 stringsAsFactors = FALSE),
                      c("Characteristic", arm_levels)))
for (s in sort(unique(pt$AESOC))) {
  sc <- soc |> dplyr::filter(AESOC == s)
  rows[[length(rows) + 1]] <- setNames(
    data.frame(s, sc[, arm_levels], check.names = FALSE, stringsAsFactors = FALSE),
    c("Characteristic", arm_levels))
  pts <- pt |> dplyr::filter(AESOC == s) |> dplyr::arrange(match(AEDECOD, .pt_order))
  if (nrow(pts))
    rows[[length(rows) + 1]] <- setNames(
      data.frame(paste0("    ", pts$AEDECOD), pts[, arm_levels],
                 check.names = FALSE, stringsAsFactors = FALSE),
      c("Characteristic", arm_levels))
}
disp <- dplyr::bind_rows(rows)

# Metadata lives HERE, in rtfreporter -- not in Tplyr.
pages <- as_rtftables(disp, col_header = ae_col_header, split = "group_force",
                      max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
                      cell_format = fmt_ae, col_spec = ae_col_spec,
                      col_rel_width = ae_widths, row_height_twips = 200)
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = ae_header, footer = ae_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "ae_tplyr.rtf", overwrite = TRUE)
```

```{r img-ae-tplyr, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("ae_tplyr.png", "AE summary via Tplyr (transpose-and-set)")
```

## cards + tfrmt

`tfrmt` is driven by a long **group / label / column / param / value** table. The
trick for an AE table is that a row whose **`group` equals its `label`** renders
as a *single* line that also carries its statistic -- so giving each SOC a
`group = label = <SOC>` row puts the SOC count **on the SOC row**, just like the
others; the PT rows (`group = SOC`, `label = PT`) become the indented children,
and the overall row is its own one-row group. We complete the arm x row grid so
a zero is a real `0`, and sort via `sorting_cols`.

```{r ae-tfrmt}
library(tidyr); library(tfrmt)
N          <- as.integer(arm_n[arm_levels]); names(N) <- arm_levels
real_pairs <- .soc_pt |> dplyr::filter(AEDECOD %in% .keep_pt)
kept_socs  <- sort(unique(real_pairs$AESOC))

# PT counts per arm, completed to a full grid so a missing cell is a real 0.
pt_c <- adae |> dplyr::distinct(TRT01A, USUBJID, AESOC, AEDECOD) |>
  dplyr::count(TRT01A, AESOC, AEDECOD, name = "n") |>
  dplyr::mutate(AESOC = as.character(AESOC), AEDECOD = as.character(AEDECOD)) |>
  dplyr::semi_join(real_pairs, by = c("AESOC", "AEDECOD")) |>
  tidyr::complete(TRT01A, tidyr::nesting(AESOC, AEDECOD), fill = list(n = 0)) |>
  dplyr::mutate(p = n / N[as.character(TRT01A)])
# SOC distinct-subject counts (independent of the PTs).
soc_c <- adae |> dplyr::distinct(TRT01A, USUBJID, AESOC) |>
  dplyr::count(TRT01A, AESOC, name = "n") |>
  dplyr::mutate(AESOC = as.character(AESOC)) |>
  dplyr::filter(AESOC %in% kept_socs) |>
  tidyr::complete(TRT01A, AESOC = kept_socs, fill = list(n = 0)) |>
  dplyr::mutate(p = n / N[as.character(TRT01A)])
# Overall any-AE count.
ov_c <- adae |> dplyr::distinct(TRT01A, USUBJID) |> dplyr::count(TRT01A, name = "n") |>
  dplyr::mutate(p = n / N[as.character(TRT01A)])

# long form: a SOC row has group == label (count shown on the SOC line); PT rows
# are group = SOC, label = PT; the overall row is its own group.  ord1 / ord2 are
# the SOC and PT sort keys.
long <- dplyr::bind_rows(
    ov_c  |> dplyr::transmute(group = ANY_AE, label = ANY_AE,
                              column = as.character(TRT01A), n, p),
    soc_c |> dplyr::transmute(group = AESOC,  label = AESOC,
                              column = as.character(TRT01A), n, p),
    pt_c  |> dplyr::transmute(group = AESOC,  label = AEDECOD,
                              column = as.character(TRT01A), n, p)) |>
  tidyr::pivot_longer(c(n, p), names_to = "param", values_to = "value") |>
  dplyr::mutate(
    ord1   = ifelse(group == ANY_AE, 0L, match(group, kept_socs)),
    ord2   = ifelse(label == group, 0L, match(label, .pt_order)),
    column = factor(column, levels = arm_levels),
    group  = factor(group, levels = c(ANY_AE, kept_socs)))

spec <- tfrmt(
  group = group, label = label, column = column, param = param, value = value,
  sorting_cols = c(ord1, ord2),                       # SOCs A->Z, PTs by count
  body_plan = body_plan(frmt_structure(".default", ".default",
    frmt_combine("{n} ({p}%)", n = frmt("x"),
                 p = frmt("x.x", transform = ~ . * 100)))),
  col_plan = col_plan(group, label, Placebo, `Xanomeline Low Dose`,
                      `Xanomeline High Dose`, -ord1, -ord2))

pages <- as_rtftables(print_to_gt(spec, long), read_meta = FALSE, split = "group_force",
                      max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
                      cell_format = fmt_ae, col_header = ae_col_header,
                      col_spec = ae_col_spec, col_rel_width = ae_widths,
                      row_height_twips = 200)
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = ae_header, footer = ae_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "ae_tfrmt.rtf", overwrite = TRUE)
```

```{r img-ae-tfrmt, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("ae_tfrmt.png", "AE summary via cards + tfrmt (group == label puts the count on the SOC row)")
```

## flextable and huxtable (convert and render)

No need to rebuild the table for these two: the gtsummary table converts straight
to a **flextable** or a **huxtable**, and `rtfreporter` reads either. (We bake
the PT indent into the label text first with `bake_indent()`, because the
converters turn gtsummary's indent into cell padding that the flextable /
huxtable adapters do not read as a row-label indent.)

```{r ae-flextable-huxtable}
library(gtsummary); library(flextable); library(huxtable)

# bake_indent(): used ONLY here. as_flex_table() / as_hux_table() turn
# gtsummary's indent into cell padding that the flextable / huxtable adapters do
# not read as a row-label indent, so we bake the PT indent into the label TEXT
# (NBSP) before converting.
bake_indent <- function(tbl, n = 4L) {
  tb <- tbl$table_body
  pt <- tb$variable == "AEDECOD"
  tb$label[pt] <- paste0(strrep(" ", n), tb$label[pt])
  tbl$table_body <- tb
  tbl$table_styling$indent <- tbl$table_styling$indent[0, ]
  tbl
}

# The same gtsummary table as the gtsummary section above (AE_SORT is from the
# data block).
g <- tbl_hierarchical(
    data = adae, variables = c(AESOC, AEDECOD), by = TRT01A,
    denominator = adsl, id = USUBJID, overall_row = TRUE,
    label = list("..ard_hierarchical_overall.." ~ ANY_AE),
    statistic = ~ "{n} ({p}%)", digits = everything() ~ list(p = 1)) |>
  filter_hierarchical(p >= 0.03) |>
  sort_hierarchical(sort = AE_SORT)

# flextable
pages <- as_rtftables(gtsummary::as_flex_table(bake_indent(g)), read_meta = FALSE, split = "group_force",
                      max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
                      cell_format = fmt_ae, col_header = ae_col_header,
                      col_spec = ae_col_spec, col_rel_width = ae_widths,
                      row_height_twips = 200)
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = ae_header, footer = ae_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "ae_flextable.rtf", overwrite = TRUE)

# huxtable
pages <- as_rtftables(gtsummary::as_hux_table(bake_indent(g)), read_meta = FALSE, split = "group_force",
                      max_rows = AE_MAX_ROWS, blank_rows = "between_groups",
                      cell_format = fmt_ae, col_header = ae_col_header,
                      col_spec = ae_col_spec, col_rel_width = ae_widths,
                      row_height_twips = 200)
doc <- rtf_document(page = list(paper_size = "letter", orientation = "landscape")) |>
  rtf_section(page = 1, secinfo = list(header = ae_header, footer = ae_footer)) |>
  rtf_tables(pages)
generate_rtfreport(doc, "ae_huxtable.rtf", overwrite = TRUE)
```

```{r img-ae-flextable, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("ae_flextable.png", "AE summary via gtsummary -> flextable")
```

```{r img-ae-huxtable, eval = TRUE, echo = FALSE, out.width = "100%", purl = FALSE}
.snapshot("ae_huxtable.png", "AE summary via gtsummary -> huxtable")
```

> **Seven ways, one report.** rtables/tern, gtsummary, cards + gtsummary, Tplyr,
> cards + tfrmt, and gtsummary converted to flextable / huxtable -- all the same
> AE table, all rendered to the same paginated RTF by `rtfreporter`.

---

*Coming next, under the same showcase: real Word snapshots in place of the
current placeholders.*

<!-- where-next -->

## Where next

- [The pharmaverse catalog](tlg-catalog.html) --- every showcase table in one place

The four recipes (`?rtfreporter-recipes`) are the same ground covered as runnable help-page examples: demographics, adverse events, PK and laboratory, each data-in to RTF-out.
