---
title: "Tables from an ARD"
description: >
  From a cards / cardx analysis results dataset to RTF pages: normalize_ard(),
  table_plan() and the plan_*() verbs, and rtf_tables().
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
options(width = 100)
# The RTF files are written to a scratch folder; data-raw/gen_plan_articles.R
# points this at the folder it turns into the screenshots shown below.
knitr::opts_knit$set(root.dir = Sys.getenv("RTFREPORTER_ARTICLE_OUT", tempdir()))
# The screenshots are committed in inst/rtf-examples/plan/ (made by
# data-raw/gen_plan_articles.R) and copied next to the article to be served.
.here <- normalizePath(".")          # the article's folder, before root.dir moves
.shot <- function(file) {
  src <- file.path(.here, "..", "..", "inst", "rtf-examples", "plan", file)
  if (!file.exists(src)) return(invisible(NULL))
  dir.create(file.path(.here, "figures", "plan"), showWarnings = FALSE,
             recursive = TRUE)
  file.copy(src, file.path(.here, "figures", "plan", file), overwrite = TRUE)
  knitr::include_graphics(file.path("figures", "plan", file), error = FALSE)
}
```

Most of rtfreporter starts from a table that already exists: a data frame, a
`gt`, an `rtables` object, and `as_rtftables()` turns it into pages. This
article starts one step earlier, from an **analysis results dataset** (ARD):
the long frame of statistics that [cards](https://insightsengineering.github.io/cards/)
and [cardx](https://insightsengineering.github.io/cardx/) produce, one number a
row.

An ARD holds the numbers but not how the table shows them. Somebody still has to
say which key goes across and which goes down, how a mean and an SD become one
cell, what the rows are called and what the header says. In rtfreporter that
is written as a **plan**:

```
ARD ──normalize_ard()──▶ flat frame ──table_plan()──▶ plan ──plan_*() verbs──▶ plan
                                                                       │
                                     rtf_tables(doc, plan) ◀───────────┘
```

* `normalize_ard()` flattens the ARD: groups and hierarchy become ordinary key
  columns.
* `table_plan()` names the **roles**: the column that goes across, the ones
  that go down.
* each `plan_*()` verb adds one declaration: the cell templates, the labels,
  the header, the widths. Nothing runs yet.
* `rtf_tables(doc, plan)` (or `plan_apply(plan)`) runs the plan and gives
  `rtftable` pages, the same objects `as_rtftables()` makes.

```{r setup}
library(rtfreporter)
library(cards)
```

## The data and the ARD

The examples use the CDISC pilot study as the
[pharmaverseadam](https://pharmaverse.github.io/pharmaverseadam/) package
ships it. The treatment is made a factor so the arms come out in study order.

```{r data}
arms <- c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
adsl <- pharmaverseadam::adsl
adsl <- adsl[adsl$SAFFL == "Y", ]
adsl$TRT01A <- factor(adsl$TRT01A, levels = arms)
```

A demographics ARD: age summarised, three characteristics tabulated, all by
arm. `continuous_summary_fns()` adds the quartiles, which cards leaves out by
default.

```{r ard-dm}
ard_dm <- ard_stack(
  adsl, .by = TRT01A,
  ard_continuous(
    variables = AGE,
    statistic = ~ continuous_summary_fns(
      c("N", "mean", "sd", "median", "p25", "p75", "min", "max"))),
  ard_categorical(variables = c(AGEGR1, SEX, RACE)))
```

`ard_stack(.by = )` also counts the arms themselves. The plan reads those
counts for the `(N=86)` in the column header, so nobody types them.

## Step 1: flatten it

A cards ARD keeps its groups as pairs of columns (`group1`, `group1_level`)
whose values are lists. `normalize_ard()` turns each group into a column of
its own, named after the variable, and adds the columns a table needs: the
kind of each row (`.kind`), its label and its depth in a hierarchy.

```{r normalize}
nd <- normalize_ard(ard_dm)
nd[1:4, c("TRT01A", "variable", "variable_level", "context",
          "stat_name", "stat", ".kind")]
```

This frame is what the plan works on. Its columns are the names you will
use in `table_plan()`: look at them first.

## Step 2: the roles

`table_plan()` takes the roles and nothing else, like `ggplot(aes())`:

* `cols`: the key that goes **across**, one column per arm;
* `rows`: the keys that go **down**, here the analysis variable, called
  `group` in the table.

```{r roles}
p <- nd |>
  table_plan(cols = "TRT01A", rows = c(group = "variable"))
```

## Step 3: the cells

`plan_cells()` says how the statistics become text. A cell is a template:
`{mean:.1f}` is the mean to one decimal, `{p:.1f%}` a proportion shown as a
percentage. Entries can be keyed by the kind of variable:

* `continuous`: a **named** vector, one printed row per name. The name is the
  row label.
* `categorical`: one template for every level, `n (%)`. A formula entry comes
  first and wins where it applies: `n == 0 ~ "0"` prints a zero count without
  its `(0.0)`.

```{r cells}
p <- p |>
  plan_cells(
    continuous  = c("n"               = "{N:.0f}",
                    "Mean (SD)"       = "{mean:.1f} ({sd:.2f})",
                    "Median (Q1, Q3)" = "{median:.1f} ({p25:.1f}, {p75:.1f})",
                    "Min, Max"        = "{min:.0f}, {max:.0f}"),
    categorical = c(n == 0 ~ "0", "{n:.0f} ({p:.1f%})"),
    notes = FALSE)
```

`notes = FALSE` turns off the message listing the statistics no template
used. Leave it on while you write a plan: it is how you notice a statistic
you forgot.

## Step 4: how it looks

The rest is display. The variables get readable labels and their levels an
order, then the stub, blank rows, widths and header are declared:

```{r display}
p <- p |>
  plan_labels(c(AGE    = "Age (years)",
                AGEGR1 = "Age group, n (%)",
                SEX    = "Sex, n (%)",
                RACE   = "Race, n (%)")) |>
  plan_levels(AGEGR1 = c("18-64", ">64"), SEX = c("F", "M")) |>
  plan_stub(name = "row_label") |>
  plan_blanks(where = "between_groups", first = TRUE, last = TRUE) |>
  plan_columns(widths = c(4, 2, 2, 2)) |>
  plan_style(border = "tfl", align_count_pct = TRUE) |>
  plan_col_header(values = list(n = TRUE), rtf_col_header(
    c("",               "{col}"),
    c("Characteristic", "(N={n})")))
```

* `plan_stub()` folds the variable and its rows into one indented column, the
  usual look of a demographics table.
* `plan_col_header()` writes the header with the same `rtf_col_header()` as
  anywhere else. Its cells can use tokens: `{col}` is the column's arm and
  `{n}` its population. `values = list(n = TRUE)` says to read that
  population from the ARD.
* `align_count_pct = TRUE` lines up the `n` and the `(%)` of every count in a
  column.

The plan still has not run. Printing it lists what it declares, and the
values the header tokens will take:

```{r print-plan}
p
```

## Step 5: run it

`rtf_tables()` takes the plan as it takes any table, and runs it:

```{r rtf-dm}
doc <- rtf_document() |>
  rtf_section(secinfo = list(
    header = rtf_header(list(
      c(l = "Protocol: CDISCPILOT01", r = "Page {PAGE} of {TOTAL_PAGES}"),
      c(c = "Table 14.1.1  Demographic Characteristics"),
      c(c = "Safety Analysis Set"),
      c(c = ""))),
    footer = rtf_footer(list(
      c(l = "SD = standard deviation; Q1, Q3 = first and third quartiles."))))) |>
  rtf_tables(p)
generate_rtfreport(doc, "t_dm.rtf", overwrite = TRUE)
```

```{r shot-dm, echo = FALSE}
.shot("t_dm.png")
```

To see a stage without writing RTF, use `plan_apply()`. It gives the pages
by default. `stage = "table"` gives the table data frame before it is split
into pages, which is often the quickest way to check a cell:

```{r apply-table}
tb <- plan_apply(p, stage = "table")
tb[1:7, ]
```

## A layer can be overridden

A plan is a list of layers, and **a later layer wins** over what an earlier
one said about the same key. A house style can be written once and one
table adjusted by adding a line, not by editing the original:

```{r last-wins}
p2 <- p |>
  plan_labels(AGE = "Age at baseline (years)") |>
  plan_cells(AGE = c("n" = "{N:.0f}", "Mean (SD)" = "{mean:.1f} ({sd:.1f})"))
plan_apply(p2, stage = "table")[1:4, ]
```

The `AGE` entry is narrower than `continuous`, so it applies to `AGE` alone.
The other variables keep the kind-wide template.

## Adverse events: a hierarchy, sorted by frequency

A table of adverse events by system organ class (SOC) and preferred term (PT)
starts from `ard_stack_hierarchical()`. `over_variables = TRUE` adds the
"any event" row, and the denominator is the safety population in ADSL, so the
percentages are of the arm.

```{r ard-ae}
adae <- pharmaverseadam::adae
adae <- adae[adae$TRTEMFL %in% "Y" & adae$SAFFL %in% "Y", ]
adae$TRT01A <- factor(adae$TRT01A, levels = arms)

ard_ae <- ard_stack_hierarchical(
  adae, variables = c(AEBODSYS, AEDECOD), by = TRT01A,
  denominator = adsl, id = USUBJID, over_variables = TRUE)
```

`normalize_ard()` is told which variables are the hierarchy, and what to
call the overall row:

```{r plan-ae}
p_ae <- ard_ae |>
  normalize_ard(hierarchy = c("AEBODSYS", "AEDECOD"),
                overall = "Any TEAE") |>
  table_plan(cols = "TRT01A", rows = c(SOC = "AEBODSYS", PT = "AEDECOD"),
             label = NA) |>
  plan_sort(".overall", "SOC", ".depth", "-n", "PT") |>
  plan_cells("{n:.0f} ({p:.1f%})", notes = FALSE) |>
  plan_stub(name = "System Organ Class\n  Preferred Term", indent = 2) |>
  plan_blanks(where = "between_groups", first = TRUE) |>
  plan_paginate_rows(max_rows = 30, split = "group_safe") |>
  plan_columns(widths = c(5, 2, 2, 2)) |>
  plan_style(border = "tfl", align_count_pct = TRUE) |>
  plan_col_header(values = list(n = TRUE), rtf_col_header(
    c("",   "{col}"),
    c("",   "(N={n})\nn (%)")))
```

* `plan_sort(".overall", "SOC", ".depth", "-n", "PT")` is the row order, key
  by key: the overall row first, then the SOCs (alphabetically). Within a
  SOC, its own row comes first (`.depth`), then its PTs by the number of
  subjects in all arms, most first (the minus sign means descending), with
  ties by name. `-n` alone would order every row by its count and take the
  PTs away from their SOC. The keys are what keeps the hierarchy together.
* `plan_paginate_rows(split = "group_safe")` breaks pages between SOCs, not
  inside one, when a SOC fits on a page.
* The header's `{n}` here is the arm's denominator. It comes from the
  hierarchical ARD, not from a count of the adverse events.

```{r rtf-ae}
doc <- rtf_document() |>
  rtf_section(secinfo = list(
    header = rtf_header(list(
      c(l = "Protocol: CDISCPILOT01", r = "Page {PAGE} of {TOTAL_PAGES}"),
      c(c = "Table 14.3.1.1  Treatment-Emergent Adverse Events by SOC and PT"),
      c(c = "Safety Analysis Set"),
      c(c = ""))),
    footer = rtf_footer(list(
      c(l = "A subject is counted once per SOC and once per PT."))))) |>
  rtf_tables(p_ae)
generate_rtfreport(doc, "t_ae.rtf", overwrite = TRUE)
length(plan_apply(p_ae))    # pages
```

```{r shot-ae, echo = FALSE}
.shot("t_ae.png")
```

## Without a plan: `widen_ard()`

A plan defers everything until it runs, which is what lets a later layer win
and a plan be saved and reused. When you want the table data frame now,
`widen_ard()` does the ARD half in one call. It takes the same templates, and
gives you a frame to finish with `as_rtftables()` as usual:

```{r widen}
wide <- widen_ard(
  nd, cols = "TRT01A", rows = c(group = "variable"),
  cells = list(continuous  = c("Mean (SD)" = "{mean:.1f} ({sd:.2f})",
                               "Median"    = "{median:.1f}"),
               categorical = "{n:.0f} ({p:.1f%})"),
  notes = FALSE)
head(wide, 6)
```

Use `widen_ard()` to get at the numbers quickly, or to hand the frame to
something else. Use a plan for a table that is part of a report.

## Where to start: `plan_template()`

`plan_template()` reads an ARD and writes a starting plan for it, with the
roles guessed, a template for every kind of variable it finds, and the
display verbs set to common values. It is meant to be pasted into your program
and edited:

```{r template, eval = FALSE}
plan_template(ard_dm, cols = "TRT01A", file = "t_dm_plan.R")
```

## Where next

* [The plan verbs](plan-verbs.html): what each verb declares, grouped by job.
* [Listings with a plan](plan-listings.html): a plan with no ARD at all.
* [From as_rtftables() to a plan](plan-and-as-rtftables.html): which verb
  takes over each `as_rtftables()` argument.
